### 2021-12-19 doc/manuscript.Rmd
Initial commit

---
title: "Longer term All-Cause and Cardiovascular Mortality with Intensive Blood Pressure Control"
bibliography: refs.bib
csl: jacc.csl
always_allow_html: true
output: 
  officedown::rdocx_document:
    reference_docx: style_manuscript_times_new_roman.docx
---

Nicholas M. Pajewski, PhD^1^ 
Byron C. Jaeger, PhD^1^ 
Paul E. Drawz, DEGREES^2^ 
Paul K. Whelton, DEGREES^3^ 
Mark A. Supiano, DEGREES^4^ 
Jeff D. Williamson, DEGREES^5^, and 
David M. Reboussin, DEGREES^1^

^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC

^2^Division of Renal Diseases & Hypertension, University of Minnesota, Minneapolis.

^3^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.

^4^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.

^5^Section on Gerontology and Geriatric Medicine, Wake Forest School of Medicine, Winston-Salem, NC.

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE,
                      dpi = 600, 
                      fig.height = 7, 
                      fig.width = 6.75)

# specification for table tables
rspec <- round_spec() |>
  round_using_magnitude(digits = c(2, 1, 0),
                        breaks = c(10, 100, Inf))

# save it to options:
names(rspec) <- paste('table.glue', names(rspec), sep = '.')
options(rspec)

```

```{r load-targets, include=FALSE}

tar_load(names = c(recoders,
                   base_viz_inc,
                   base_ds, 
                   long_ds,
                   long_inf,
                   base_inf,
                   ndi_baseline))

base_ds <- mutate(base_ds,
                  variable = str_remove(variable, '^base_ds_'))

long_ds <- mutate(long_ds,
                  variable = str_remove(variable, '^long_ds_'))

long_inf <- mutate(long_inf,
                   variable = str_remove(variable, '^long_inf_'))

base_inf <- mutate(base_inf,
                   variable = str_remove(variable, '^base_inf_'))

inline <- list()

tbl_variables <- c('variable', 'group', 'treatment')

inline$base_ds <- base_ds |>
  mutate(across(matches("est$|lwr$|upr"), table_value),
         across(matches("^n_"), ~table_value(as.integer(.x)))) |>
  as_inline(tbl_variables = tbl_variables,
            tbl_values = setdiff(names(base_ds), tbl_variables))

tbl_variables <- c('variable', 'group', 'trial_phase')

inline$long_inf <- long_inf |>
  mutate(across(matches("est$|lwr$|upr"), table_value),
         across(matches("^n_"), ~table_value(as.integer(.x)))) |>
  as_inline(tbl_variables = tbl_variables,
            tbl_values = setdiff(names(long_inf), tbl_variables))

```

\newpage

# ABSTRACT

__Background__. Both the Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) trial have shown benefits of intensive blood pressure control on cardiovascular morbidity and mortality. However, as both trials were ended after slightly more than 3 years of follow-up, the legacy effect of intensive treatment on longer term cardiovascular and all-cause mortality is unknown.

__Methods__. We linked participants in SPRINT to the National Death Index (NDI, 2016 to 2020), assessing cardiovascular mortality using the NDI plus system. Cox and competing risk regression models were used to model the effect of intensive treatment through the trial and during post-trial follow-up on all-cause and cardiovascular mortality respectively

```{r}

# ap makes inline insertion of results less text-heavy.
# ap = access point, 
# b = baseline, l = longitudinal
# o = overall, i = intensive, s = standard,
# t = trial, c = cohort

ap_b_o <- inline$base_ds$overall$overall$overall
ap_b_i <- inline$base_ds$overall$overall$Intensive
ap_b_s <- inline$base_ds$overall$overall$Standard

ap_l_o_t <- inline$long_inf$overall$overall$trial
ap_l_o_c <- inline$long_inf$overall$overall$cohort

# make the . symbol equivalent to calling table_glue.
# Another way to makes inline insertion of results less text-heavy
. <- table.glue::table_glue

```

__Results__. Over a median follow-up of `r ap_b_o$fup_est` years, there were `r ap_b_i$n_cvd` and `r ap_b_i$n_acm` cardiovascular and all-cause deaths with intensive treatment respectively, and `r ap_b_s$n_cvd` cardiovascular / `r ap_b_s$n_acm` all-cause deaths for standard treatment. Intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial (follow-up through July 2016). However, there was no indication of benefit during post-trial follow-up for either cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`). Results were similar for subgroups based on baseline age, cognitive function, and frailty status.

__Conclusions__. Our results show a clear benefit for cardiovascular and all-cause mortality during the trial which was largely attenuated during post-trial observational follow-up. Given indications of increasing blood pressures in SPRINT participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension in line with current guidelines. 
\newpage

# INTRODUCTION

# METHODS

# RESULTS

# DISCUSSION

```{r}
# in case you want to make something landscape,
# wrap it in this code
#<!---BLOCK_LANDSCAPE_START--->
#<!---BLOCK_LANDSCAPE_STOP--->
# remove the # signs - you will be placing this in the markdown text
```

\newpage
__Table 1__: All-cause mortality by treatment group and subgroup

```{r}

tbl_inf <- long_inf |> 
  mutate(mort = .("{mort_est}\n({mort_lwr}, {mort_upr})")) |> 
  select(variable,
         group, 
         trial_phase,
         mort) |> 
  pivot_wider(names_from = trial_phase,
              values_from = mort,
              names_prefix = '.h_')

tbl_data <- long_ds |>
  left_join(base_ds) |>
  mutate(
    .n = .("{n_acm} / {n_obs}"),
    .i = .("{incidence_est}\n({incidence_lwr}, {incidence_upr})")
  ) |>
  select(-starts_with("incidence_"),
         -starts_with("n_"),
         -starts_with("fup_")) |> 
  pivot_wider(names_from = c(treatment, trial_phase),
              values_from = c(.n, .i)) |> 
  left_join(tbl_inf) |> 
  mutate(
    variable = replace(variable, group == variable, NA_character_),
    variable = recode(variable, !!!recoders$variable),
    group = recode(group, !!!recoders$group)
  )

tbl_object <- tbl_data |> 
  select(variable, group, 
         ends_with('trial'), 
         ends_with('cohort')) |> 
  select(variable,
         group,
         # .n_Standard_trial,
         # .n_Intensive_trial,
         .i_Standard_trial,
         .i_Intensive_trial,
         .h_trial,
         # .n_Standard_cohort,
         # .n_Intensive_cohort,
         .i_Standard_cohort,
         .i_Intensive_cohort,
         .h_cohort) |> 
  as_grouped_data(groups = 'variable') |> 
  drop_na_rows() |> 
  as_flextable(hide_grouplabel = TRUE) |> 
  add_header_row(
    values = c(" ", 
               # "N Events / N Total", 
               "Incidence (95% CI)", 
               "Hazard Ratio\n(95% CI)",
               # "N Events / N Total", 
               "Incidence (95% CI)",
               "Hazard Ratio\n(95% CI)"),
    colwidths = c(1, 2, 1, 2, 1)
  ) |> 
  add_header_row(
    values = c(" ", 
               "Trial Follow-up Through Close-out Visits",
               "Post-trial Follow-up"),
    colwidths = c(1, 3, 3)
  ) |> 
  set_header_labels(
    group = " ",
    # .n_Standard_trial   = "Standard",
    # .n_Intensive_trial  = "Intensive",
    .i_Standard_trial   = "Standard",
    .i_Intensive_trial  = "Intensive",
    .h_trial            = "Hazard Ratio\n(95% CI)",
    # .n_Standard_cohort  = "Standard",
    # .n_Intensive_cohort = "Intensive",
    .i_Standard_cohort  = "Standard",
    .i_Intensive_cohort = "Intensive",
    .h_cohort           = "Hazard Ratio\n(95% CI)"
  ) |> 
  merge_v(part = 'header') |>
  theme_box() |> 
  padding(j = 1, i = ~ !is.na(group), padding.left = 15, part = 'body') |> 
  bg(i = ~ is.na(group), bg = 'grey80') |> 
  italic(i = ~ is.na(group), italic = TRUE) |> 
  width(width = 0.95) |> 
  width(j=1, width = 1.5) |> 
  align(align = 'center', part = 'all') |> 
  align(j = 1, align = 'left', part = 'all') |> 
  fontsize(size = 12, part = 'all') |> 
  font(fontname = 'Times New Roman', part = 'all')

tbl_object

```

\newpage
__Figure 1__: Cumulative incidence of all-cause mortality.

```{r}

base_viz_inc$acm[[1]]

```

\newpage
__Figure 2__: Cumulative incidence of cardiovascular and non-cardiovascular mortality.

```{r}

base_viz_inc$cvd[[1]]

```

\newpage
# SUPPLEMENT

We have a total of `r nrow(base_viz_inc)`) plots showing all-cause mortality and mortality split into cardiovascular and non-cardiovascular causes. They are automated enough to dynamically set text and axis values that will look reasonable. To show what I mean, I've added the figures for adults 75 years or older, who have a much higher incidence.

\newpage
__Figure S1__: Cumulative incidence of all-cause mortality among participants 75 years or older.

```{r}

base_viz_inc$acm[[3]]

```

\newpage
__Figure S2__: Cumulative incidence of cardiovascular and non-cardiovascular mortality among participants 75 years or older.

```{r}

base_viz_inc$cvd[[3]]

```

### 2021-12-19 R/add_annotations.R

@@ -1,0 +1,52 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param fig
+add_annotations <- function(fig, cols, ymax = NULL, xmax = NULL) {
+
+  stopifnot(length(cols)==3)
+
+  fig_ymax <- if(is.null(ymax)) layer_scales(fig)$y$range$range[2] else ymax
+  fig_xmax <- if(is.null(xmax)) layer_scales(fig)$x$range$range[2] else xmax
+
+  fig +
+    annotate(geom = 'rect',
+             xmin = 0,
+             ymin = 0,
+             xmax = 2.54,
+             ymax = fig_ymax,
+             fill = cols[1],
+             alpha = 0.1) +
+    annotate(geom = 'rect',
+             xmin = 2.54,
+             ymin = 0,
+             xmax = 4.48,
+             ymax = fig_ymax,
+             fill = cols[2],
+             alpha = 0.1) +
+    annotate(geom = 'rect',
+             xmin = 4.48,
+             ymin = 0,
+             xmax = fig_xmax,
+             ymax = fig_ymax,
+             fill = cols[3],
+             alpha = 0.1) +
+    annotate(geom = 'text',
+             x = 1.27,
+             y = fig_ymax * 0.90,
+             size = 5,
+             label = 'Trial phase') +
+    annotate(geom = 'text',
+             x = 3.51,
+             y = fig_ymax * 0.90,
+             size = 5,
+             label = 'Trial And\nCohort\nPhase') +
+    annotate(geom = 'text',
+             x = 5.85,
+             y = fig_ymax * 0.90,
+             size = 5,
+             label = 'Cohort Phase')
+
+}

### 2021-12-19 R/cph_estimate.R

@@ -1,0 +1,18 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param model
+#' @param k_mat
+cph_estimate <- function(model, k_mat) {
+
+  model |>
+    glht(linfct = k_mat) |>
+    confint() |>
+    getElement('confint') |>
+    as_tibble() |>
+    mutate(across(everything(), exp)) |>
+    rename(est = Estimate)
+
+}

### 2021-12-19 R/cph_fit.R

@@ -1,0 +1,25 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param ndi_data
+#' @param formula
+cph_fit <- function(ndi_data, formula) {
+
+  fit <- do.call(what = survival::coxph,
+                 args = list(data = ndi_data, formula = formula))
+
+  pv <- as_tibble(
+    getElement(cox.zph(fit), 'table'),
+    rownames = 'variable'
+  ) |>
+    filter(variable == 'GLOBAL') |>
+    pull(p)
+
+  message("Non-proportional hazards global p-value = ", table_pvalue(pv))
+
+
+  return(fit)
+
+}

### 2021-12-19 R/cph_incidence.R

@@ -1,0 +1,27 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+##'
+##' @param time
+##' @param data 
+##' @param status
+cph_incidence <- function(data, time, status, div_time_by = 1000) {
+  
+  total_events <- sum(getElement(data, status))
+  total_time <- sum(getElement(data, time))
+  
+  total_time <- total_time / div_time_by
+  
+  estimate = total_events / total_time
+  
+  fit <- glm(total_events ~ offset(log(total_time)), family="poisson")
+  
+  fit_ci <- exp(confint(fit))
+  
+  list(incidence_est = estimate,
+       incidence_lwr = fit_ci['2.5 %'],
+       incidence_upr = fit_ci['97.5 %'])
+  
+}

### 2021-12-19 R/drop_na_rows.R

@@ -1,0 +1,10 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+
+drop_na_rows <- function(x){
+  keep_rows <- apply(x, 1, function(.x) !all(is.na(.x)))
+  return(x[keep_rows, ])
+}

### 2021-12-19 R/gg_themes.R

@@ -1,0 +1,20 @@
+
+
+theme_fig <- function(x, ...){
+  theme_bw() +
+    theme(text = element_text(family = 'serif', size = 12),
+          axis.text = element_text(family = 'serif', color = 'black', size = 12),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor = element_blank(),
+          panel.border = element_blank(),
+          ...)
+}
+
+theme_tbl <- function(x, ...) {
+  theme_void() +
+    theme(text = element_text(family = 'serif', size = 12),
+          plot.caption = element_text(hjust = 0, size = 2),
+          plot.title.position = "plot",
+          plot.caption.position =  "plot",
+          ...)
+}

### 2021-12-19 R/kaplan_plot_NP.R

@@ -1,0 +1,396 @@
+
+
+
+
+kmplot <- function(km, cumHaz=FALSE, mark=3, simple=FALSE,
+                   xaxis.at=pretty(km$time), xaxis.lab=xaxis.at,
+                   lty.surv=1, lwd.surv=1, col.surv=1,
+                   lty.ci=0, lwd.ci=1, col.ci=col.surv, #By default (lty.ci=0), confidence intervals are not plotted.
+                   group.names=NULL, group.order=seq(length(km$n)), extra.left.margin=4,
+                   label.n.at.risk=FALSE, draw.lines=TRUE, cex.axis=1,
+                   xlab='', ylab='', main='', xlim=c(0,max(km$time)), ylim=c(0,1),
+                   grid=TRUE, grid.vert=TRUE, lty.grid=1, lwd.grid=1, col.grid=grey(.9),
+                   legend=!is.null(km$strata), loc.legend='topright', add=FALSE,
+                   yaxis.at, yaxis.lab, leg.title, cex.text, add.zeroes=FALSE, no.ci=FALSE, cex.legend=1.0,
+                   use.grp.4legend=TRUE, legend.names, extra.top=0,
+                   move.lab.left=0, trunc.time.at, leg.ncol=1, extra.bottom = 0, nrisk.down = 0,
+                   ... # ... is passed to par()
+){
+  # Version 2.5.5: 2014/5/19
+
+  # km is the output from survfit() function in survival package.
+
+  # xaxis.at specifies where 'n at risk' will be computed and printed.
+  # xaxis.lab specifies what will be printed at xaixs.at.  (see example)
+
+  # If group names are long, add extra left margin by setting extra.left.margin to something greater than 0.
+
+  # line specifications (lty.surv, lwd.surv, col.surv) will be recycled.
+  # Set lty.ci to 1 if confidence intervals are needed.
+  # group.names will overwrite whatever is specified in survfit() output.
+  # group.order specifies the order of groups from top in 'n at risk'.  1 is at top, 2 next, and so on.
+
+  # if add=TRUE, then par() is not refreshed.  allows multiple panels by
+  # using, e.g., par(mfrow=c(2,2)).
+
[…]

### 2021-12-19 R/ndi_base_describe.R

@@ -1,0 +1,85 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param ndi_data_list
+ndi_base_describe <- function(ndi_data_list) {
+
+  map_dfr(
+    .x = ndi_data_list,
+    .id = 'group',
+    .f = function(ndi_data) {
+
+      time_by_tx <- ndi_data |>
+        group_by(treatment) |>
+        group_map(
+          .f = ~ prodlim(Hist(acm_years, 1 - acm_event) ~ 1, data = .x) |>
+            quantile() |>
+            getElement('quantiles.survival') |>
+            filter(q == 0.50) |>
+            bind_cols(.y) # .y is the group info
+        )
+
+      time_ovrl <-
+        prodlim(Hist(acm_years, 1 - acm_event) ~ 1,
+                data = ndi_data) |>
+        quantile() |>
+        getElement('quantiles.survival') |>
+        filter(q == 0.50) |>
+        mutate(treatment = 'overall')
+
+      time_by_tx_and_ovrl <-
+        time_ovrl |>
+        bind_rows(time_by_tx) |>
+        select(
+          treatment,
+          fup_est = quantile,
+          fup_lwr = lower,
+          fup_upr = upper
+        )
+
+
+      # cvd_by_tx_and_ovrl <-
+      #   ndi_data |>
+      #   split(f = ndi_data$treatment) |>
+      #   list_modify(overall = ndi_data) |>
+      #   map_dfr(.f = cph_incidence, 'acm_years', 'cvd_event',
+      #           .id = 'treatment') |>
+      #   rename(inc_cvd_est = incidence_est,
+      #          inc_cvd_lwr = incidence_lwr,
+      #          inc_cvd_upr = incidence_upr)
+      #
+      # acm_by_tx_and_ovrl <-
+      #   ndi_data |>
+      #   split(f = ndi_data$treatment) |>
+      #   list_modify(overall = ndi_data) |>
+      #   map_dfr(.f = cph_incidence, 'acm_years', 'acm_event',
+      #           .id = 'treatment') |>
+      #   rename(inc_acm_est = incidence_est,
+      #          inc_acm_lwr = incidence_lwr,
+      #          inc_acm_upr = incidence_upr)
+
+      events_by_tx_and_ovrl <-
+        ndi_data |>
+        split(f = ndi_data$treatment) |>
+        list_modify(overall = ndi_data) |>
+        map_dfr(.id = 'treatment',
[…]

### 2021-12-19 R/ndi_base_infer.R

@@ -1,0 +1,47 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param ndi_data_list
+ndi_base_infer <- function(ndi_data_list) {
+
+  # models and estimates
+
+  map_dfr(
+    .x = ndi_data_list,
+    .id = 'group',
+    .f = function(ndi_data) {
+
+      ndi_intervals <- ndi_split_time(ndi_data,
+                                      interval_size = 1.5,
+                                      event_var = "acm_event",
+                                      event_start_status = 0,
+                                      time_var = "acm_years")
+
+      acm_fit_timevarying <- ndi_intervals |>
+        mutate(treatment = as.numeric(treatment == 'Intensive')) |>
+        cph_fit(
+        formula = Surv(Start_time, Stop_time, acm_event) ~
+          treatment + treatment:time_factor + strata(randSite)
+      )
+
+
+      cph_estimate(acm_fit_timevarying,
+                   k_mat = rbind(c(1,0,0,0,0,0,0),
+                                 c(1,1,0,0,0,0,0),
+                                 c(1,0,1,0,0,0,0),
+                                 c(1,0,0,1,0,0,0),
+                                 c(1,0,0,0,1,0,0),
+                                 c(1,0,0,0,0,1,0),
+                                 c(1,0,0,0,0,0,1))) |>
+        mutate(years = levels(ndi_intervals$time_factor),
+               .before = 1)
+
+    }
+
+  )
+
+
+
+}

### 2021-12-19 R/ndi_base_viz_incidence.R

@@ -1,0 +1,139 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param base_sub
+ndi_base_viz_incidence <- function(ndi_data_list) {
+
+  cols_bg <- viridis(3)
+  cols_tx <- pal_lancet()(2)
+
+  figs <- map_dfr(
+    .x = ndi_data_list,
+    .id = 'group',
+    .f = function(ndi_data){
+
+      fit_surv <- survfit(Surv(acm_years, acm_event) ~ treatment,
+                          data = ndi_data)
+
+      ymax_candidates <- seq(0.05, 1, by = 0.05)
+
+      ymax_true <- max(1-fit_surv$lower)
+
+      ymax <- ymax_candidates[min(which(ymax_candidates > ymax_true))]
+
+      fig_kap <- survminer::ggsurvplot(
+        fit_surv,
+        data = ndi_data,
+        fun = 'event',
+        palette = "lancet",
+        ggtheme = theme_fig(),
+        fontsize = 12,
+        conf.int = TRUE,
+        risk.table.fontsize = 3.5,
+        legend.title = "Treatment",
+        legend.labs = c("Standard", "Intensive"),
+        risk.table = 'absolute',
+        break.x.by = 1,
+        ylab = 'Cumulative incidence',
+        xlab = 'Years since randomization',
+        ylim = c(0, ymax),
+        break.y.by = 0.05,
+        censor.size = 0,
+        surv.scale = 'percent',
+        tables.theme = theme_tbl(),
+        tables.height = 1/8
+      )
+
+      fig_kap$plot <- add_annotations(fig_kap$plot,
+                                      cols = cols_bg,
+                                      ymax = ymax)
+
+
+      fit_cr <- cuminc(ftime   = ndi_data$acm_years,
+                       fstatus = factor(ndi_data$cvd_event_cr,
+                                        levels = c(0, 1, 2),
+                                        labels = c("- Censored",
+                                                   "- CVD mortality",
+                                                   "- Non-CVD mortality")),
+                       group   = ndi_data$treatment, cencode = '- Censored')
+
+      data_cr <-
+        map_dfr(.x = fit_cr[1:4],
+                .f = as_tibble,
+                .id = 'fit') |>
+        separate(fit, into = c('treatment','event'), sep = ' - ') |>
+        mutate(
+          treatment = factor(
+            x = treatment,
+            levels = c("Standard", "Intensive")
[…]

### 2021-12-19 R/ndi_base_viz_tv_effect.R

@@ -1,0 +1,43 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param base_sub
+ndi_base_viz_tv_effect <- function(ndi_data_list) {
+
+  map_dfr(
+    .x = ndi_data_list,
+    .id = 'group',
+    .f = function(ndi_data){
+
+      fit <- comp.risk(
+        formula = Event(acm_years, cvd_event_cr) ~ treatment + cluster(randSite),
+        cause = 1,
+        model = 'prop',
+        n.sim = 1000,
+        data = ndi_data
+      )
+
+      fit_var <- as_tibble(fit$var.cum) |>
+        transmute(time, sd = sqrt(treatmentIntensive))
+
+      fit_data <- as_tibble(fit$cum) |>
+        left_join(fit_var) |>
+        transmute(
+          time,
+          hr_est = exp(treatmentIntensive),
+          hr_lwr = exp(treatmentIntensive - 1.96 * sd),
+          hr_upr = exp(treatmentIntensive + 1.96 * sd)
+        ) |>
+        add_row(.data = tibble(time = 0,
+                               hr_est = 1,
+                               hr_lwr = 1,
+                               hr_upr = 1),
+                .before = 2)
+
+    }
+  )
+
+
+}

### 2021-12-19 R/ndi_load.R

@@ -1,0 +1,50 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param sasinet_drive
+ndi_load <- function(sasinet_drive = "Z", fname, ...) {
+
+  path <- file.path(
+    paste0(sasinet_drive, ":"),
+    "npajewski",
+    "NDI",
+    "Data",
+    fname
+  )
+
+  read_csv(path) |>
+    transmute(
+      # ... makes it easy to modify the call to this function in targets
+      # for including additional variables or modifying them.
+      ...,
+      age_yrs = age,
+      age_cat = factor(sub_senior,
+                       levels = c(0, 1),
+                       labels = c("lt75_years", "gteq75_years")),
+      sex = factor(female,
+                   levels = c(0,1),
+                   labels = c("Male","Female")),
+      race = factor(is.na(race_black),
+                    levels = c(TRUE, FALSE),
+                    labels = c("Non_Black", "Black")),
+      ckd = factor(eGFR_CKDEPI < 60,
+                   levels = c(FALSE, TRUE),
+                   labels = c("No", "Yes")),
+      moca = factor(moca_status,
+                    levels = c(0, 1),
+                    labels = c("gt10th_percentile",
+                               "lteq10th_percentile")),
+      frail = factor(frail_status,
+                     levels = c(0,1,2),
+                     labels = c("Fit",
+                                "Pre_frail",
+                                "Frail")),
+      # treatment indicator with factor structure is useful for tables
+      treatment = factor(intensive,
+                         levels = c(0,1),
+                         labels = c('Standard', 'Intensive'))
+    )
+
+}

### 2021-12-19 R/ndi_long_describe.R

@@ -1,0 +1,50 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param ndi_data_list
+ndi_long_describe <- function(ndi_data_list) {
+
+  incidence_mort <- map_dfr(
+    .x = ndi_data_list,
+    .id = 'group',
+    .f = function(ndi_data){
+
+
+      ..ndi_data <-
+        ndi_data |>
+        mutate(
+          time = if_else(trial_phase == 'trial',
+                         end_yrs,
+                         end_yrs - start_yrs),
+          cvd_mortality = as.numeric(cvd_mortality == 1)
+        ) |>
+        split(f = list(ndi_data$trial_phase,
+                       ndi_data$treatment))
+
+      ..ndi_data |>
+        map_dfr(cph_incidence,
+                time = 'time',
+                status = 'mortality',
+                .id = 'split_me') |>
+        separate(col = 'split_me',
+                 into = c('trial_phase', 'treatment'),
+                 sep = '\\.')
+
+    }
+  )
+
+  # ndi_data |>
+  #   split(ndi_data$trial_phase) |>
+  #   map()
+  # cvd_inc <-
+  #   cuminc(ftime   = ..ndi_data$end_yrs,
+  #          fstatus = ..ndi_data$cvd_mortality,
+  #          cencode = 0,
+  #          group = ..ndi_data$treatment)
+  #
+  # ndi_data <- ndi_data_list[[1]]
+
+
+}

### 2021-12-19 R/ndi_long_infer.R

@@ -1,0 +1,148 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param ndi_data_list
+
+ndi_long_infer <- function(ndi_data_list, n_sim = 1000) {
+
+  mort_by_phase <- map_dfr(
+    .x = ndi_data_list,
+    .id = 'group',
+    .f = function(ndi_data){
+
+      frmla <-
+        Surv(start_yrs, end_yrs, mortality) ~
+        treatment +
+        trial_phase +
+        treatment:trial_phase +
+        strata(randSite)
+
+      fit <- cph_fit(ndi_data, frmla)
+
+      k_mat <- rbind(c(1,0,0),
+                     c(1,0,1))
+
+      cph_estimate(fit, k_mat) |>
+        mutate(trial_phase = levels(ndi_data$trial_phase), .before = 1) |>
+        rename(mort_est = est,
+               mort_lwr = lwr,
+               mort_upr = upr)
+
+    }
+
+  )
+
+  stratify <- all(names(ndi_data_list) == c("lt75_years", "gteq75_years"))
+
+  cvd_by_phase <- map_dfr(
+    .x = ndi_data_list,
+    .id = 'group',
+    .f = function(ndi_data){
+
+      # if the time-varying approach fails to converge or gives
+      # impossible hazard ratio estimates, estimate hazard ratios
+      # separately in the two phases.
+
+      if(stratify){
+
+        ..ndi_data <- ndi_data |>
+          mutate(time = if_else(trial_phase == 'trial',
+                                end_yrs,
+                                end_yrs - start_yrs),
+                 .before = start_yrs)
+
+        fits <- ..ndi_data |>
+          split(..ndi_data$trial_phase) |>
+          map(
+            .f = ~ {
+              m <- comp.risk(
+                formula = Event(time, cvd_mortality) ~
+                  const(treatment) + cluster(randSite),
+                cause = 1,
+                model = 'prop',
+                n.sim = 1000,
+                data = .x
+              )
+              class(m) <- c('bcj_comprisk', class(m))
+              m
+            }
+          )
+
+        return(
+          map_dfr(.x = fits,
+                  .f = cph_estimate,
+                  k_mat = matrix(1),
+                  .id = 'trial_phase') |>
+            rename(cvd_est = est,
[…]

### 2021-12-19 R/ndi_split_group.R

@@ -1,0 +1,33 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param ndi_data
+#' @param group
+ndi_split_group <- function(ndi_data, group) {
+
+  if(group == 'overall') return(list(overall = ndi_data))
+
+  split(ndi_data, f = ndi_data[[group]])
+
+  # long_dat <- 'trial_phase' %in% names(ndi_data)
+  #
+  # if(group == 'overall' && !long_dat) return(list(overall = ndi_data))
+  #
+  # fctr <- vector(mode = 'list', length = 2)
+  # fctr_index <- 1
+  #
+  # if(long_dat){
+  #   fctr[[fctr_index]] <- ndi_data$trial_phase_fctr
+  #   fctr_index <- fctr_index + 1
+  # }
+  #
+  # if(group != 'overall'){
+  #   fctr[[fctr_index]] <- ndi_data[[group]]
+  #   fctr_index <- fctr_index + 1
+  # }
+  #
+  # split(ndi_data, f = fctr[seq(fctr_index-1, 1)], sep = '')
+
+}

### 2021-12-19 R/ndi_split_time.R

@@ -1,0 +1,21 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param ndi_data
+ndi_split_time <- function(ndi_data,
+                           interval_size,
+                           time_var,
+                           event_var,
+                           event_start_status) {
+
+  ndi_data |>
+    timeSplitter(by = interval_size,
+                 time_var = time_var,
+                 event_var = event_var,
+                 event_start_status = event_start_status) |>
+    as_tibble() |>
+    mutate(time_factor = factor(Start_time))
+
+}

### 2021-12-19 R/recoders_make.R

@@ -1,0 +1,39 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+
+recoders_make <- function() {
+
+  list(
+
+    variable = c('age_cat' = 'Age, years',
+                 'sex' = 'Sex',
+                 'race' = 'Race',
+                 'ckd' = 'Chronic Kidney Disease',
+                 'moca' = 'Cognitive Function',
+                 'frail' = 'Frailty Status'),
+
+    group = c(
+      "overall" = "Overall",
+      "lt75_years" = "<75 years",
+      "gteq75_years" = "\u226575 years",
+      "Male" = "Male",
+      "Female" = "Female",
+      "Non_Black" = "Non-Black",
+      "Black" = "Black",
+      "No" = "No",
+      "Yes" = "Yes",
+      "gt10th_percentile" = ">10th percentile",
+      "lteq10th_percentile" = "\u226410th percentile",
+      "Fit" = "Fit (FI\u22640.10)",
+      "Pre_frail" = "Pre-frail (0.10<FI\u22640.21)",
+      "Frail" = "Frail (FI>0.21)"
+    )
+
+  )
+
+
+
+}

### 2021-12-19 _targets.R

@@ -1,0 +1,115 @@
+## Load your packages, e.g. library(targets).
+source("./packages.R")
+
+## Load your R files
+lapply(list.files("./R", full.names = TRUE), source)
+
+subgroups <- tibble(group = c("overall",
+                              "age_cat",
+                              "sex",
+                              "race",
+                              "ckd",
+                              "moca",
+                              "frail"))
+list(
+
+  tar_target(recoders, recoders_make()),
+
+  tar_target(
+    ndi_baseline,
+    ndi_load(sasinet_drive = 'Z',
+             fname = 'longterm_death.csv',
+             # participant identifier
+             pid,
+             # study site
+             randSite,
+             # all cause mortality time and statue
+             acm_years, acm_event,
+             # cvd mortality status
+             cvd_event = as.numeric(cvd_event_cr == 1),
+             cvd_event_cr)
+  ),
+
+  tar_target(
+    ndi_longitudinal,
+    ndi_load(sasinet_drive = 'Z',
+             fname = 'longterm_death_td_subgroup.csv',
+             pid,
+             randSite,
+             trial_phase = factor(trial_phase,
+                                  levels = c(0,1),
+                                  labels = c("trial", "cohort")),
+             start_yrs, end_yrs,
+             mortality, cvd_mortality)
+  ),
+
+  # arrow assignment is necessary to make 'long' an object that
+  # will be recognized in downstream tar_ functions
+  long <- tar_map(
+    values = subgroups,
+    # make the sub-groups
+    tar_target(long_sub, ndi_split_group(ndi_longitudinal, group)),
+    # ds = descriptive statistics
+    tar_target(long_ds, ndi_long_describe(long_sub)),
+    # inf = inference
+    tar_target(long_inf, ndi_long_infer(long_sub))
+  ),
+
+  tar_combine(
+    long_ds,
+    long[[2]],
+    command = bind_rows(!!!.x, .id = "variable")
+  ),
+
+  tar_combine(
+    long_inf,
+    long[[3]],
+    command = bind_rows(!!!.x, .id = "variable")
+  ),
+
+  base <- tar_map(
+    values = subgroups,
+    # make the sub-groups
+    tar_target(base_sub, ndi_split_group(ndi_baseline, group)),
[…]

### 2021-12-19 packages.R

@@ -1,0 +1,29 @@
+## library() calls go here
+library(conflicted)
+library(dotenv)
+library(targets)
+library(tarchetypes)
+
+library(ggsci)
+library(tidyverse)
+library(survival)
+library(Greg)
+library(timereg)
+library(multcomp)
+library(riskRegression)
+library(prodlim)
+library(viridis)
+library(table.glue)
+library(flextable)
+library(survminer)
+library(cmprsk)
+
+conflict_prefer("filter", "dplyr")
+conflict_prefer("select", "dplyr")
+conflict_prefer("border", "flextable")
+conflict_prefer("font", "flextable")
+conflict_prefer("rotate", "flextable")
+
+
+
+library(rmarkdown)

### 2021-12-23 doc/manuscript.Rmd
multi-panel landscape setup

@@
 
-Nicholas M. Pajewski, PhD^1^ 
 Byron C. Jaeger, PhD^1^ 
@@
 Mark A. Supiano, DEGREES^4^ 
-Jeff D. Williamson, DEGREES^5^, and 
-David M. Reboussin, DEGREES^1^
+Adam P. Bress, PharmD, MS^5,6^
+Jeff D. Williamson, DEGREES^7^
+David M. Reboussin, DEGREES^1^, and
+Nicholas M. Pajewski, PhD^1^
+
+<!-- Jackson T. Wright Jr., MD, PhD, Alfred K. Cheung, MD, William C. Cushman, MD, Cora E. Lewis, MD, Suzanne Oparil, MD, Michael V. Rocco, MD and Karen Johnson, MD -->
 
@@
 
-^5^Section on Gerontology and Geriatric Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+^5^	Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT
+
+^6^	Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT
+
+^7^Section on Gerontology and Geriatric Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
 
@@
                       message = FALSE,
+                      warning = FALSE,
                       dpi = 600, 
-                      fig.height = 7, 
-                      fig.width = 6.75)
+                      fig.width = 11, 
+                      fig.height = 6.25)
 
@@
                    base_viz_inc,
+                   base_viz_eff,
+                   long_viz_forest,
                    base_ds, 
@@
 
-base_ds <- mutate(base_ds,
-                  variable = str_remove(variable, '^base_ds_'))
-
-long_ds <- mutate(long_ds,
-                  variable = str_remove(variable, '^long_ds_'))
-
-long_inf <- mutate(long_inf,
-                   variable = str_remove(variable, '^long_inf_'))
-
-base_inf <- mutate(base_inf,
-                   variable = str_remove(variable, '^base_inf_'))
+fig <- left_join(base_viz_inc, base_viz_eff)
 
@@
 
-\newpage
-__Figure 1__: Cumulative incidence of all-cause mortality.
+```{r}
+
+block_section(
+  prop_section(
+    page_size = page_size(orient = "portrait"),
+    type = "continuous"
+  )
+)
+
+```
+
+__Figure 1__: (A) Cumulative incidence of all-cause mortality by treatment group. (B) Time-dependent effect of being randomized to intensive treatment on all-cause mortality.
 
@@
 
-base_viz_inc$acm[[1]]
+blank_plot <- ggplot(data.frame()) + aes() + theme_void()
+
+legend <- get_legend(
+  fig$inc_acm[[1]]$plot + 
+    theme(legend.position = c(.55, .2),
+          legend.direction = 'horizontal')
+)
+
+plot_grid(
+  # row 1 (legend)
+  legend, 
+  blank_plot,
+  # row 2 (figures)
+  fig$inc_acm[[1]]$plot + theme(legend.position = ''), 
+  fig$eff_acm[[1]],
+  # row 3 (table)
+  fig$inc_acm[[1]]$table, 
+  blank_plot,
+  ncol = 2,
+  align = 'v',
+  axis = 'lr',
+  rel_heights = c(1,10,2),
+  labels = c("","","A","B","",""),
+  label_x = 0.1,
+  label_y = 1.05
+)
 
@@
 \newpage
-__Figure 2__: Cumulative incidence of cardiovascular and non-cardiovascular mortality.
+__Figure 2__: (A) Cumulative incidence of cardiovascular and non-cardiovascular mortality by treatment group.
+(B) Time-dependent effect of being randomized to intensive treatment on cardiovascular mortality.
 
@@
 
-base_viz_inc$cvd[[1]]
+plot_grid(
+  # row 1 (legend)
+  legend, 
+  blank_plot,
+  # row 2 (figures)
+  fig$inc_cvd[[1]] + theme(legend.position = ''), 
+  fig$eff_cvd[[1]],
+  ncol = 2,
+  align = 'v',
+  axis = 'lr',
+  rel_heights = c(1,10,2),
+  labels = c("","","A","B","",""),
+  label_x = 0.1,
+  label_y = 1.05
+)
 
@@
 \newpage
+__Figure 3__: Hazard ratio for all-cause mortality for participants randomized to intensive versus standard treatment, overall and in subgroups.
+
+```{r}
+
+long_viz_forest
+
+```
+
 # SUPPLEMENT
 
-We have a total of `r nrow(base_viz_inc)`) plots showing all-cause mortality and mortality split into cardiovascular and non-cardiovascular causes.
+We have a total of `r nrow(base_viz_inc)` plots showing all-cause mortality and mortality split into cardiovascular and non-cardiovascular causes.
 They are automated enough to dynamically set text and axis values that will look reasonable.
@@
 
-base_viz_inc$acm[[3]]
+# base_viz_inc$acm[[3]]
 
@@
 
-base_viz_inc$cvd[[3]]
+# base_viz_inc$cvd[[3]]
 
 ```
+
+```{r}
+block_section(
+  prop_section(
+    page_size = page_size(orient = "landscape"),
+    page_margins = page_mar(bottom = 1/8, top = 1/8, 
+                            right = 1/4, left = 1/4),
+    type = "continuous"
+  )
+)
+```

### 2021-12-23 R/add_annotations.R

@@ -4,49 +4,60 @@
 #'
 #' @title
 #' @param fig
-add_annotations <- function(fig, cols, ymax = NULL, xmax = NULL) {
+add_annotations <- function(fig,
+                            cols,
+                            ymax = NULL,
+                            ymin = NULL,
+                            xmax = NULL,
+                            xmin = NULL,
+                            ymult = 0.90,
+                            x_cohort_phase = 6,
+                            textsize = 4.5) {
 
   stopifnot(length(cols)==3)
 
   fig_ymax <- if(is.null(ymax)) layer_scales(fig)$y$range$range[2] else ymax
   fig_xmax <- if(is.null(xmax)) layer_scales(fig)$x$range$range[2] else xmax
 
+  fig_ymin <- if(is.null(ymin)) 0 else ymin
+  fig_xmin <- if(is.null(xmin)) 0 else xmin
+
   fig +
     annotate(geom = 'rect',
-             xmin = 0,
-             ymin = 0,
+             xmin = fig_xmin,
+             ymin = fig_ymin,
              xmax = 2.54,
              ymax = fig_ymax,
              fill = cols[1],
              alpha = 0.1) +
     annotate(geom = 'rect',
              xmin = 2.54,
-             ymin = 0,
+             ymin = fig_ymin,
              xmax = 4.48,
              ymax = fig_ymax,
              fill = cols[2],
              alpha = 0.1) +
     annotate(geom = 'rect',
              xmin = 4.48,
-             ymin = 0,
+             ymin = fig_ymin,
              xmax = fig_xmax,
              ymax = fig_ymax,
              fill = cols[3],
              alpha = 0.1) +
     annotate(geom = 'text',
              x = 1.27,
-             y = fig_ymax * 0.90,
-             size = 5,
+             y = fig_ymax * ymult,
+             size = textsize,
              label = 'Trial phase') +
     annotate(geom = 'text',
              x = 3.51,
-             y = fig_ymax * 0.90,
-             size = 5,
+             y = fig_ymax * ymult,
+             size = textsize,
              label = 'Trial And\nCohort\nPhase') +
     annotate(geom = 'text',
-             x = 5.85,
-             y = fig_ymax * 0.90,
-             size = 5,
+             x = x_cohort_phase,
+             y = fig_ymax * ymult,
+             size = textsize,
              label = 'Cohort Phase')
 
 }

### 2021-12-23 R/ndi_base_viz_incidence.R

@@ -82,7 +82,7 @@ ndi_base_viz_incidence <- function(ndi_data_list) {
             y = est,
             col = treatment,
             group = interaction(treatment, event)) +
-        geom_line(size = 1.2) +
+        geom_line(size = 1.5) +
         geom_ribbon(
           aes(x = time,
               ymin = est - sqrt(var) * 1.96,
@@ -103,7 +103,7 @@ ndi_base_viz_incidence <- function(ndi_data_list) {
                            limits = c(0, ymax)) +
         scale_x_continuous(breaks = c(0:10),
                            expand = c(0, 0, 0, 0),
-                           limits = c(0, 12)) +
+                           limits = c(0, 13)) +
         scale_color_manual(values = cols_tx) +
         scale_fill_manual(values = cols_tx)
 
@@ -118,20 +118,22 @@ ndi_base_viz_incidence <- function(ndi_data_list) {
       fig_cr <- add_annotations(fig_cr,
                                 cols = cols_bg,
                                 ymax = ymax,
-                                xmax = 12) +
+                                xmax = 13,
+                                x_cohort_phase = 6.25) +
         annotate(geom = 'text',
-                 x = 11.1,
+                 x = 11.5,
                  y = y_text$est[y_text$event == 'Non-CVD mortality'],
                  size = 5,
                  label = 'Non-CVD\nMortality') +
         annotate(geom = 'text',
-                 x = 11.1,
+                 x = 11.5,
                  y = y_text$est[y_text$event == 'CVD mortality'],
                  size = 5,
                  label = 'CVD\nMortality')
 
 
-      tibble(acm = list(fig_kap), cvd = list(fig_cr))
+      tibble(inc_acm = list(fig_kap),
+             inc_cvd = list(fig_cr))
 
     }
   )

### 2021-12-23 R/ndi_base_viz_tv_effect.R

@@ -6,35 +6,103 @@
 #' @param base_sub
 ndi_base_viz_tv_effect <- function(ndi_data_list) {
 
+  cols_bg <- viridis(3)
+  cols_tx <- pal_lancet()(2)
+
   map_dfr(
     .x = ndi_data_list,
     .id = 'group',
     .f = function(ndi_data){
 
-      fit <- comp.risk(
+
+      fit_acm <- timecox(
+        Surv(acm_years, acm_event) ~ treatment + cluster(randSite),
+        data = ndi_data
+      )
+
+      var_acm <- as_tibble(fit_acm$robvar.cum) |>
+        transmute(time, se = sqrt(treatmentIntensive))
+
+      fit_acm_data <- as_tibble(fit_acm$cum) |>
+        left_join(var_acm) |>
+        transmute(
+          time,
+          hr_est = exp(treatmentIntensive),
+          hr_lwr = exp(treatmentIntensive - 1.96 * se),
+          hr_upr = exp(treatmentIntensive + 1.96 * se)
+        )
+
+      fig_eff_acm <- ggplot(fit_acm_data) +
+        aes(x = time, y = hr_est, ymin = hr_lwr, ymax = hr_upr) +
+        geom_ribbon(alpha = 0.2) +
+        geom_line(color = cols_tx[2]) +
+        scale_y_log10(limits = c(0.15, 2.5), breaks = c(0.15, 0.25, 0.5, 1)) +
+        scale_x_continuous(breaks = seq(0,10)) +
+        theme_fig() +
+        geom_segment(aes(x = 0, xend = max(time),
+                         y = 1, yend = 1),
+                     linetype = 2,
+                     color = 'black') +
+        labs(x = 'Years since randomization',
+             y = 'Hazard ratio (95% CI)')
+
+      fig_eff_acm <- add_annotations(fig_eff_acm, cols = cols_bg,
+                                     ymax = 2.5,
+                                     xmax = 10,
+                                     ymin = 0.15,
+                                     ymult = 0.75)
+
+      times <- ndi_data$acm_years[ndi_data$acm_event>0]
+      times <- times[times > 1.5]
+
+      fit_cvd <- comp.risk(
         formula = Event(acm_years, cvd_event_cr) ~ treatment + cluster(randSite),
         cause = 1,
         model = 'prop',
         n.sim = 1000,
+        # times = times,
         data = ndi_data
       )
 
-      fit_var <- as_tibble(fit$var.cum) |>
[…]

### 2021-12-23 R/ndi_long_viz_forest.R

@@ -1,0 +1,128 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param long_inf
+ndi_long_viz_forest <- function(long_inf, recoders) {
+
+  rspec <- round_spec() |>
+    round_using_decimal(digits = 2)
+
+  data_init <- long_inf |>
+    mutate(variable = str_remove(variable, 'long_inf_'),
+           variable = recode(variable, !!!recoders$variable),
+           group = recode(group, !!!recoders$group))
+
+  data_bind <- tibble(variable = unique(data_init$variable),
+                      group = recode(variable, !!!recoders$variable)) |>
+    expand_grid(tibble(trial_phase = c("cohort", "trial")))
+
+  data_gg <- bind_rows(data_bind, data_init) |>
+    mutate(variable = fct_inorder(variable),
+           group = fct_inorder(group)) |>
+    arrange(variable, group, trial_phase) |>
+    slice(-c(1,3)) |>
+    group_by(trial_phase) |>
+    mutate(
+      x = seq(n(), 1),
+      .before = 1,
+      across(where(is.factor), as.character),
+      group = if_else(
+        condition = group == variable | group == 'Overall',
+        true = group,
+        false = paste(" ", group)
+      ),
+      trial_phase = factor(
+        trial_phase,
+        levels = c('trial', 'cohort'),
+        labels = c("Trial Phase", "Cohort Phase")
+      ),
+      fill = 1 +
+        as.numeric(variable == 'Overall') +
+        as.numeric(variable == 'Overall' & trial_phase == 'Cohort Phase') +
+        as.numeric(variable != group) +
+        as.numeric(variable != group & trial_phase == 'Cohort Phase'),
+      fill = factor(fill),
+      label = table_value(mort_est, rspec = rspec),
+      label = replace(label, label == 'NA', "")
+    )
+
+  range_y <- range(data_gg$mort_lwr, data_gg$mort_upr, na.rm = TRUE)
+
+  data_gg <- data_gg |>
+    mutate(
+      text_y = if_else(
+        condition = trial_phase == 'Trial Phase',
+        true = range_y[2],
+        false = range_y[1]
+      ),
+      hjust = if_else(
+        condition = trial_phase == 'Trial Phase',
+        true = 0,
+        false = 1
+      ),
+      .before = fill
+    )
+
+
[…]

### 2021-12-23 R/recoders_make.R

@@ -8,12 +8,13 @@ recoders_make <- function() {
 
   list(
 
-    variable = c('age_cat' = 'Age, years',
-                 'sex' = 'Sex',
-                 'race' = 'Race',
-                 'ckd' = 'Chronic Kidney Disease',
-                 'moca' = 'Cognitive Function',
-                 'frail' = 'Frailty Status'),
+    variable = c("overall" = "Overall",
+                 "age_cat" = "Age, years",
+                 "sex" = "Sex",
+                 "race" = "Race",
+                 "ckd" = "Chronic Kidney Disease",
+                 "moca" = "Cognitive Function",
+                 "frail" = "Frailty Status"),
 
     group = c(
       "overall" = "Overall",

### 2021-12-23 _targets.R

@@ -11,6 +11,7 @@ subgroups <- tibble(group = c("overall",
                               "ckd",
                               "moca",
                               "frail"))
+
 list(
 
   tar_target(recoders, recoders_make()),
@@ -58,13 +59,19 @@ list(
   tar_combine(
     long_ds,
     long[[2]],
-    command = bind_rows(!!!.x, .id = "variable")
+    command = bind_rows(!!!.x, .id = "variable") |>
+      mutate(variable = str_remove(variable, '^long_ds_'))
   ),
 
   tar_combine(
     long_inf,
     long[[3]],
-    command = bind_rows(!!!.x, .id = "variable")
+    command = bind_rows(!!!.x, .id = "variable") |>
+      mutate(variable = str_remove(variable, '^long_inf_'))
+  ),
+
+  tar_target(
+    long_viz_forest, ndi_long_viz_forest(long_inf, recoders)
   ),
 
   base <- tar_map(
@@ -86,19 +93,29 @@ list(
   tar_combine(
     base_ds,
     base[[2]],
-    command = bind_rows(!!!.x, .id = "variable")
+    command = bind_rows(!!!.x, .id = "variable") |>
+      mutate(variable = str_remove(variable, '^base_ds_'))
   ),
 
   tar_combine(
     base_inf,
     base[[3]],
-    command = bind_rows(!!!.x, .id = "variable")
+    command = bind_rows(!!!.x, .id = "variable") |>
+      mutate(variable = str_remove(variable, '^base_inf_'))
   ),
 
   tar_combine(
     base_viz_inc,
     base[[4]],
-    command = bind_rows(!!!.x, .id = "variable")
+    command = bind_rows(!!!.x, .id = "variable") |>
+      mutate(variable = str_remove(variable, '^base_viz_inc_'))
+  ),
+
+  tar_combine(
+    base_viz_eff,
+    base[[5]],
+    command = bind_rows(!!!.x, .id = "variable") |>
+      mutate(variable = str_remove(variable, '^base_viz_eff_'))
   ),
 
   tar_render(manuscript, "doc/manuscript.Rmd")

### 2021-12-23 packages.R

@@ -17,13 +17,15 @@ library(table.glue)
 library(flextable)
 library(survminer)
 library(cmprsk)
+library(officer)
+library(cowplot)
 
 conflict_prefer("filter", "dplyr")
 conflict_prefer("select", "dplyr")
 conflict_prefer("border", "flextable")
 conflict_prefer("font", "flextable")
 conflict_prefer("rotate", "flextable")
-
+conflict_prefer("get_legend", "cowplot")
 
 
 library(rmarkdown)

### 2021-12-31 doc/manuscript.Rmd
Initial commit

@@
 bibliography: refs.bib
-csl: jacc.csl
+csl: lancet.csl
 always_allow_html: true
@@
                    base_viz_eff,
-                   long_viz_forest,
+                   long_viz_mort,
+                   long_viz_cvd,
                    base_ds, 
@@
 
+<!-- END OF ABSTRACT -->
+
 # INTRODUCTION
@@
 
+___Statistical Analysis___: The effect of randomization to intensive treatment was estimated as a function of time using two approaches.
+The first approach split the follow-up of study participants into non-overlapping trial and cohort phases, then estimated the effect of randomization to intensive treatment in each phase [@zhang_time_varying_2018].
+The second approach estimated the cumulative time-dependent effect of randomization to intensive treatment [@martinussen2007dynamic].
+While the phase-specific estimate of the treatment effect treats the trial and cohort phases independently, the cumulative estimate enforces continuity of the treatment effect through time so that the effect of treatment during the trial phase is not 'forgotten' during the cohort phase.
+All analyses accounted for correlation within study sites [@lin_robust_1989], and analyses of cardiovascular mortality accounted for the competing risk of non-cardiovascular mortality [@scheike_flexible_2008].
+
 # RESULTS
+
+In both treatment groups, median follow-up time was `r ap_b_o$fup_est` years.
+A total of `r ap_b_i$n_acm` and `r ap_b_s$n_acm` all-cause mortality events occurred among participants randomized to intensive and standard treatment, respectively (__Table 1__).
+The hazard ratio (HR) for all-cause mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during the cohort phase.
+The cumulative time-varying effect of intensive versus standard treatment indicated lower risk for all-cause mortality during the trial phase and was attenuated during the cohort phase (__Figure 1__).
 
@@
 ```{r}
-# in case you want to make something landscape,
-# wrap it in this code
-#<!---BLOCK_LANDSCAPE_START--->
-#<!---BLOCK_LANDSCAPE_STOP--->
-# remove the # signs - you will be placing this in the markdown text
+
+block_section(
+  prop_section(
+    page_size = page_size(orient = "portrait"),
+    type = "continuous"
+  )
+)
+
 ```
 
-\newpage
 __Table 1__: All-cause mortality by treatment group and subgroup
@@
 tbl_data <- long_ds |>
-  left_join(base_ds) |>
   mutate(
+    across(starts_with("n_"), as.integer),
     .n = .("{n_acm} / {n_obs}"),
-    .i = .("{incidence_est}\n({incidence_lwr}, {incidence_upr})")
+    .i = .("{inc_mort_est}\n({inc_mort_lwr}, {inc_mort_upr})")
   ) |>
-  select(-starts_with("incidence_"),
-         -starts_with("n_"),
-         -starts_with("fup_")) |> 
+  select(-starts_with("inc_"),
+         -starts_with("n_")) |> 
   pivot_wider(names_from = c(treatment, trial_phase),
@@
 tbl_object <- tbl_data |> 
-  select(variable, group, 
-         ends_with('trial'), 
-         ends_with('cohort')) |> 
   select(variable,
          group,
-         # .n_Standard_trial,
-         # .n_Intensive_trial,
+         .n_Standard_trial,
+         .n_Intensive_trial,
          .i_Standard_trial,
@@
          .h_trial,
-         # .n_Standard_cohort,
-         # .n_Intensive_cohort,
+         .n_Standard_cohort,
+         .n_Intensive_cohort,
          .i_Standard_cohort,
@@
     values = c(" ", 
-               # "N Events / N Total", 
-               "Incidence (95% CI)", 
+               "N Events / N Total", 
+               "Incidence\n(95% CI)", 
                "Hazard Ratio\n(95% CI)",
-               # "N Events / N Total", 
-               "Incidence (95% CI)",
+               "N Events / N Total", 
+               "Incidence\n(95% CI)",
                "Hazard Ratio\n(95% CI)"),
-    colwidths = c(1, 2, 1, 2, 1)
+    colwidths = c(1, 2, 2, 1, 2, 2, 1)
   ) |> 
@@
                "Post-trial Follow-up"),
-    colwidths = c(1, 3, 3)
+    colwidths = c(1, 5, 5)
   ) |> 
@@
     group = " ",
-    # .n_Standard_trial   = "Standard",
-    # .n_Intensive_trial  = "Intensive",
+    .n_Standard_trial   = "Standard",
+    .n_Intensive_trial  = "Intensive",
     .i_Standard_trial   = "Standard",
@@
     .h_trial            = "Hazard Ratio\n(95% CI)",
-    # .n_Standard_cohort  = "Standard",
-    # .n_Intensive_cohort = "Intensive",
+    .n_Standard_cohort  = "Standard",
+    .n_Intensive_cohort = "Intensive",
     .i_Standard_cohort  = "Standard",
@@
 
+\newpage
+__Table 2__: Cardiovascular and non-cardiovascular mortality by treatment group and subgroup
+
 ```{r}
 
-block_section(
-  prop_section(
-    page_size = page_size(orient = "portrait"),
-    type = "continuous"
+tbl_inf <- long_inf |> 
+  mutate(cvd = .("{cvd_est}\n({cvd_lwr}, {cvd_upr})"),
+         non = .("{other_est}\n({other_lwr}, {other_upr})")) |> 
+  select(variable,
+         group, 
+         trial_phase,
+         cvd,
+         non) |> 
+  pivot_wider(names_from = trial_phase,
+              values_from = c(cvd, non),
+              names_glue = ".h_{.value}_{trial_phase}")
+
+tbl_data <- long_ds |>
+  mutate(
+    across(starts_with("n_"), as.integer),
+    .n = .("{n_cvd} / {n_other} / {n_obs}")
+  ) |>
+  select(-starts_with("inc_"),
+         -starts_with("n_")) |> 
+  pivot_wider(names_from = c(treatment, trial_phase),
+              values_from = c(.n),
+              names_prefix = '.n_') |> 
+  left_join(tbl_inf) |> 
+  mutate(
+    variable = replace(variable, group == variable, NA_character_),
+    variable = recode(variable, !!!recoders$variable),
+    group = recode(group, !!!recoders$group)
   )
-)
+
+tbl_object <- tbl_data |> 
+  select(variable,
+         group,
+         .n_Standard_trial,
+         .n_Intensive_trial,
+         .h_cvd_trial,
+         .h_non_trial,
+         .n_Standard_cohort,
+         .n_Intensive_cohort,
+         .h_cvd_cohort,
+         .h_non_cohort) |> 
+  as_grouped_data(groups = 'variable') |> 
+  drop_na_rows() |> 
+  as_flextable(hide_grouplabel = TRUE) |> 
+  add_header_row(
+    values = c(" ", 
+               "N\nCVD / non-CVD / Total", 
+               "Hazard Ratio\n(95% CI)",
+               "N\nCVD / non-CVD / Total", 
+               "Hazard Ratio\n(95% CI)"),
+    colwidths = c(1, 2, 2, 2, 2)
+  ) |> 
+  add_header_row(
+    values = c(" ", 
+               "Trial Follow-up Through Close-out Visits",
+               "Post-trial Follow-up"),
+    colwidths = c(1, 4, 4)
+  ) |> 
+  set_header_labels(
+    group = " ",
+    .n_Standard_trial= "Standard",
+    .n_Intensive_trial= "Intensive",
+    .h_cvd_trial = "CVD",
+    .h_non_trial = "Non-CVD",
+    .n_Standard_cohort = "Standard",
+    .n_Intensive_cohort = "Intensive",
+    .h_cvd_cohort = "CVD",
+    .h_non_cohort = "Non-CVD"
+  ) |> 
+  merge_v(part = 'header') |>
+  theme_box() |> 
+  padding(j = 1, i = ~ !is.na(group), padding.left = 15, part = 'body') |> 
+  bg(i = ~ is.na(group), bg = 'grey80') |> 
+  italic(i = ~ is.na(group), italic = TRUE) |> 
+  width(width = 0.95) |> 
+  width(j=1, width = 1.5) |> 
+  align(align = 'center', part = 'all') |> 
+  align(j = 1, align = 'left', part = 'all') |> 
+  fontsize(size = 12, part = 'all') |> 
+  font(fontname = 'Times New Roman', part = 'all')
+
+tbl_object
 
@@
 
-__Figure 1__: (A) Cumulative incidence of all-cause mortality by treatment group. (B) Time-dependent effect of being randomized to intensive treatment on all-cause mortality.
+\newpage
+__Figure 1__: (A) Cumulative incidence of all-cause mortality by treatment group. (B) Time-dependent effect of randomization to intensive treatment for all-cause mortality.
 
@@
 __Figure 2__: (A) Cumulative incidence of cardiovascular and non-cardiovascular mortality by treatment group.
-(B) Time-dependent effect of being randomized to intensive treatment on cardiovascular mortality.
+(B) Time-dependent effect of randomization to intensive treatment for cardiovascular mortality.
 
@@
 \newpage
-__Figure 3__: Hazard ratio for all-cause mortality for participants randomized to intensive versus standard treatment, overall and in subgroups.
+__Figure 3__: Mean systolic blood pressure over time by treatment group. Shaded areas indicate a 95% confidence interval for the mean.
 
@@
 
-long_viz_forest
+knitr::include_graphics(
+  path = 'Z:/bjaeger/sprint-mortality-ndi/fig/SPRINT_EHR_SBP_LongTerm_Comparison_122721.png', 
+  dpi = 600
+)
+
+```
+
+\newpage
+__Figure 4__: All-cause mortality hazard ratio for participants randomized to intensive versus standard treatment.
+
+```{r}
+
+long_viz_mort
+
+```
+
+\newpage
+__Figure 5__: Cardiovascular mortality hazard ratio for participants randomized to intensive versus standard treatment.
+
+```{r}
+
+long_viz_cvd
 
@@
 
-We have a total of `r nrow(base_viz_inc)` plots showing all-cause mortality and mortality split into cardiovascular and non-cardiovascular causes.
-They are automated enough to dynamically set text and axis values that will look reasonable.
-To show what I mean, I've added the figures for adults 75 years or older, who have a much higher incidence.
-
-\newpage
-__Figure S1__: Cumulative incidence of all-cause mortality among participants 75 years or older.
-
-```{r}
-
-# base_viz_inc$acm[[3]]
-
-```
-
-\newpage
-__Figure S2__: Cumulative incidence of cardiovascular and non-cardiovascular mortality among participants 75 years or older.
-
-```{r}
-
-# base_viz_inc$cvd[[3]]
-
-```
+To be decided.
 
@@
 ```
+
+# REFERENCES

### 2021-12-31 R/ndi_base_viz_incidence.R

@@ -57,7 +57,8 @@ ndi_base_viz_incidence <- function(ndi_data_list) {
                                         labels = c("- Censored",
                                                    "- CVD mortality",
                                                    "- Non-CVD mortality")),
-                       group   = ndi_data$treatment, cencode = '- Censored')
+                       group   = ndi_data$treatment,
+                       cencode = '- Censored')
 
       data_cr <-
         map_dfr(.x = fit_cr[1:4],

### 2021-12-31 R/ndi_long_describe.R

@@ -11,14 +11,13 @@ ndi_long_describe <- function(ndi_data_list) {
     .id = 'group',
     .f = function(ndi_data){
 
-
       ..ndi_data <-
         ndi_data |>
         mutate(
           time = if_else(trial_phase == 'trial',
                          end_yrs,
                          end_yrs - start_yrs),
-          cvd_mortality = as.numeric(cvd_mortality == 1)
+          .before = start_yrs
         ) |>
         split(f = list(ndi_data$trial_phase,
                        ndi_data$treatment))
@@ -30,21 +29,38 @@ ndi_long_describe <- function(ndi_data_list) {
                 .id = 'split_me') |>
         separate(col = 'split_me',
                  into = c('trial_phase', 'treatment'),
-                 sep = '\\.')
+                 sep = '\\.') |>
+        rename(inc_mort_est = incidence_est,
+               inc_mort_lwr = incidence_lwr,
+               inc_mort_upr = incidence_upr)
+
+
+
 
     }
   )
 
-  # ndi_data |>
-  #   split(ndi_data$trial_phase) |>
-  #   map()
-  # cvd_inc <-
-  #   cuminc(ftime   = ..ndi_data$end_yrs,
-  #          fstatus = ..ndi_data$cvd_mortality,
-  #          cencode = 0,
-  #          group = ..ndi_data$treatment)
-  #
-  # ndi_data <- ndi_data_list[[1]]
+  n_events <- map_dfr(
+    .x = ndi_data_list,
+    .id = 'group',
+    .f = function(ndi_data){
+      ndi_data |>
+        group_by(trial_phase, treatment) |>
+        summarize(n_obs = n(),
+                  n_acm = sum(mortality),
+                  n_cvd = sum(cvd_mortality == 1),
+                  n_other = sum(cvd_mortality == 2))
+    }
+  )
 
+
+  reduce(
+    .x = list(incidence_mort,
+              n_events),
+    .f = left_join,
+    by = c('trial_phase',
+           'treatment',
+           'group')
+  )
 
 }

### 2021-12-31 R/ndi_long_infer.R

@@ -34,81 +34,86 @@ ndi_long_infer <- function(ndi_data_list, n_sim = 1000) {
 
   )
 
-  stratify <- all(names(ndi_data_list) == c("lt75_years", "gteq75_years"))
+  stratify <- TRUE
 
-  cvd_by_phase <- map_dfr(
-    .x = ndi_data_list,
-    .id = 'group',
-    .f = function(ndi_data){
+  cvd_by_phase <- ndi_data_list |>
+    expand_grid(cause = c('cvd', 'other')) |>
+    transmute(
+      group = names(ndi_data_list),
+      cause,
+      estimates = map2(
+        .x = ndi_data_list,
+        .y = cause,
+        .f = function(ndi_data, cause_string){
 
-      # if the time-varying approach fails to converge or gives
-      # impossible hazard ratio estimates, estimate hazard ratios
-      # separately in the two phases.
+          # if the time-varying approach fails to converge or gives
+          # impossible hazard ratio estimates, estimate hazard ratios
+          # separately in the two phases.
 
-      if(stratify){
+          cause <- switch(cause_string,
+                          'cvd' = 1,
+                          'other' = 2)
 
-        ..ndi_data <- ndi_data |>
-          mutate(time = if_else(trial_phase == 'trial',
-                                end_yrs,
-                                end_yrs - start_yrs),
-                 .before = start_yrs)
+          if(stratify){
 
-        fits <- ..ndi_data |>
-          split(..ndi_data$trial_phase) |>
-          map(
-            .f = ~ {
-              m <- comp.risk(
-                formula = Event(time, cvd_mortality) ~
-                  const(treatment) + cluster(randSite),
-                cause = 1,
-                model = 'prop',
-                n.sim = 1000,
-                data = .x
+            ..ndi_data <- ndi_data |>
+              mutate(time = if_else(trial_phase == 'trial',
+                                    end_yrs,
+                                    end_yrs - start_yrs),
+                     .before = start_yrs)
+
+            fits <- ..ndi_data |>
+              split(..ndi_data$trial_phase) |>
+              map(
+                .f = ~ {
+                  m <- comp.risk(
+                    formula = Event(time, cvd_mortality) ~
+                      const(treatment) + cluster(randSite),
+                    cause = cause,
+                    model = 'prop',
+                    n.sim = 1000,
+                    data = .x
+                  )
[…]

### 2021-12-31 R/ndi_long_viz_forest.R

@@ -4,7 +4,7 @@
 #'
 #' @title
 #' @param long_inf
-ndi_long_viz_forest <- function(long_inf, recoders) {
+ndi_long_viz_mort <- function(long_inf, recoders) {
 
   rspec <- round_spec() |>
     round_using_decimal(digits = 2)
@@ -121,6 +121,131 @@ ndi_long_viz_forest <- function(long_inf, recoders) {
           axis.title.y = element_blank()) +
     facet_wrap(~trial_phase) +
     labs(y = 'Hazard ratio for all-cause mortality, intensive versus standard treatment') +
+    scale_y_log10()
+
+  fig_hrs
+
+}
+
+
+ndi_long_viz_cvd <- function(long_inf, recoders) {
+
+  rspec <- round_spec() |>
+    round_using_decimal(digits = 2)
+
+  data_init <- long_inf |>
+    mutate(variable = str_remove(variable, 'long_inf_'),
+           variable = recode(variable, !!!recoders$variable),
+           group = recode(group, !!!recoders$group),
+           cvd_lwr = if_else(cvd_lwr < 0.10, true = cvd_est, false = cvd_lwr))
+
+  data_bind <- tibble(variable = unique(data_init$variable),
+                      group = recode(variable, !!!recoders$variable)) |>
+    expand_grid(tibble(trial_phase = c("cohort", "trial")))
+
+  data_gg <- bind_rows(data_bind, data_init) |>
+    mutate(variable = fct_inorder(variable),
+           group = fct_inorder(group)) |>
+    arrange(variable, group, trial_phase) |>
+    slice(-c(1,3)) |>
+    group_by(trial_phase) |>
+    mutate(
+      x = seq(n(), 1),
+      .before = 1,
+      across(where(is.factor), as.character),
+      group = if_else(
+        condition = group == variable | group == 'Overall',
+        true = group,
+        false = paste(" ", group)
+      ),
+      trial_phase = factor(
+        trial_phase,
+        levels = c('trial', 'cohort'),
+        labels = c("Trial Phase", "Cohort Phase")
+      ),
+      fill = 1 +
+        as.numeric(variable == 'Overall') +
+        as.numeric(variable == 'Overall' & trial_phase == 'Cohort Phase') +
+        as.numeric(variable != group) +
+        as.numeric(variable != group & trial_phase == 'Cohort Phase'),
+      fill = factor(fill),
[…]

### 2021-12-31 _targets.R

@@ -70,9 +70,8 @@ list(
       mutate(variable = str_remove(variable, '^long_inf_'))
   ),
 
-  tar_target(
-    long_viz_forest, ndi_long_viz_forest(long_inf, recoders)
-  ),
+  tar_target(long_viz_mort, ndi_long_viz_mort(long_inf, recoders)),
+  tar_target(long_viz_cvd, ndi_long_viz_cvd(long_inf, recoders)),
 
   base <- tar_map(
     values = subgroups,

### 2021-12-31 R/ndi_base_viz_tv_effect.R
changing to coxph

@@ -20,6 +20,13 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
         data = ndi_data
       )
 
+      fit_acm <- coxph(
+        formula = Surv(acm_years, acm_event) ~ treatment + cluster(randSite),
+        data = ndi_data
+      )
+
+      x <- cox.zph(fit_acm)
+
       var_acm <- as_tibble(fit_acm$robvar.cum) |>
         transmute(time, se = sqrt(treatmentIntensive))
 

### 2021-12-31 R/ndi_base_viz_tv_effect.R
coxph for tv effect
we tried to use timecox from the timereg package but the time-varying
cumulative effect plots did not give a time-varying effect that matched
timecox's own estimate of the overall effect, i.e. the cumulative effect
at the end of the trial is much lower than the estimated marginal
effect.

@@ -14,36 +14,33 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
     .id = 'group',
     .f = function(ndi_data){
 
-
-      fit_acm <- timecox(
-        Surv(acm_years, acm_event) ~ treatment + cluster(randSite),
-        data = ndi_data
-      )
-
       fit_acm <- coxph(
         formula = Surv(acm_years, acm_event) ~ treatment + cluster(randSite),
         data = ndi_data
       )
 
-      x <- cox.zph(fit_acm)
+      zph <- cox.zph(fit_acm)
 
-      var_acm <- as_tibble(fit_acm$robvar.cum) |>
-        transmute(time, se = sqrt(treatmentIntensive))
+      zph_data <- tibble(time = zph$time,
+                         resid = zph$y[, 'treatment', drop = TRUE])
 
-      fit_acm_data <- as_tibble(fit_acm$cum) |>
-        left_join(var_acm) |>
-        transmute(
-          time,
-          hr_est = exp(treatmentIntensive),
-          hr_lwr = exp(treatmentIntensive - 1.96 * se),
-          hr_upr = exp(treatmentIntensive + 1.96 * se)
-        )
+      resid_fit <- lm(resid ~ ns(time, df = 6), zph_data)
+
+      resid_pred <- predict(resid_fit, se.fit = TRUE)
+
+      fit_acm_data <- tibble(
+        time = zph$time,
+        hr_est = exp(resid_pred$fit),
+        hr_lwr = exp(resid_pred$fit - 1.96 * resid_pred$se.fit),
+        hr_upr = exp(resid_pred$fit + 1.96 * resid_pred$se.fit)
+      )
 
       fig_eff_acm <- ggplot(fit_acm_data) +
         aes(x = time, y = hr_est, ymin = hr_lwr, ymax = hr_upr) +
         geom_ribbon(alpha = 0.2) +
         geom_line(color = cols_tx[2]) +
-        scale_y_log10(limits = c(0.15, 2.5), breaks = c(0.15, 0.25, 0.5, 1)) +
+        scale_y_log10(limits = c(0.35, 2.5),
+                      breaks = c(0.5, 1)) +
         scale_x_continuous(breaks = seq(0,10)) +
         theme_fig() +
         geom_segment(aes(x = 0, xend = max(time),
@@ -56,7 +53,7 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
[…]

### 2021-12-31 packages.R

@@ -7,6 +7,7 @@ library(tarchetypes)
 library(ggsci)
 library(tidyverse)
 library(survival)
+library(splines)
 library(Greg)
 library(timereg)
 library(multcomp)

### 2022-01-01 R/ndi_base_viz_tv_effect.R
tt instead of cox.zph
cox.zph gave reasonable looking time-varying effect plots but would not
really be 'good enough' for a paper as the trend line is just tracing
a line of best fit through residuals. Using tt() goes one step further
and fits a model with time-varying effect for treatment that we can then
use to plot the estimate hazard ratio across the full study timeline.
Notably, model plot and residual plot look very similar.

@@ -14,25 +14,50 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
     .id = 'group',
     .f = function(ndi_data){
 
+      # Deprecated b/c this was an analysis of residuals
+      # fit_acm <- coxph(
+      #   formula = Surv(acm_years, acm_event) ~ treatment + cluster(randSite),
+      #   data = ndi_data
+      # )
+      # zph <- cox.zph(fit_acm)
+      #
+      # zph_data <- tibble(time = zph$time,
+      #                    resid = zph$y[, 'treatment', drop = TRUE])
+      #
+      # resid_fit <- lm(resid ~ ns(time, df = 6), zph_data)
+      #
+      # resid_pred <- predict(resid_fit, se.fit = TRUE)
+      #
+      # fit_acm_data <- tibble(
+      #   time = zph$time,
+      #   hr_est = exp(resid_pred$fit),
+      #   hr_lwr = exp(resid_pred$fit - 1.96 * resid_pred$se.fit),
+      #   hr_upr = exp(resid_pred$fit + 1.96 * resid_pred$se.fit)
+      # )
+
+      # Using the tt() approach instead of residual analysis
+      # this gives model estimates of the time-varying effect
       fit_acm <- coxph(
-        formula = Surv(acm_years, acm_event) ~ treatment + cluster(randSite),
+        formula = Surv(acm_years, acm_event) ~ treatment +
+          tt(treatment) + cluster(randSite),
+        tt = function(x, t, ...)
+          as.numeric(x == 'Intensive') * ns(t, df=6),
         data = ndi_data
       )
 
-      zph <- cox.zph(fit_acm)
+      plot_times <- seq(0, 10, length.out = 1000)
 
-      zph_data <- tibble(time = zph$time,
-                         resid = zph$y[, 'treatment', drop = TRUE])
+      spline_x <- cbind(1, ns(x = plot_times, df=6))
 
-      resid_fit <- lm(resid ~ ns(time, df = 6), zph_data)
+      yhat <- spline_x %*% coef(fit_acm)
 
-      resid_pred <- predict(resid_fit, se.fit = TRUE)
+      ses <- sqrt(diag(spline_x %*% vcov(fit_acm) %*% t(spline_x)))
 
       fit_acm_data <- tibble(
-        time = zph$time,
-        hr_est = exp(resid_pred$fit),
-        hr_lwr = exp(resid_pred$fit - 1.96 * resid_pred$se.fit),
[…]

### 2022-01-03 R/fig_tv_eff_make.R
memory for ggplots
ggplot saves an environment object that can make targets take up
large amounts of memory.

@@ -1,0 +1,11 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+
+fig_tv_eff_make <- function() {
+
+  NULL
+
+}

### 2022-01-03 R/ndi_base_viz_tv_effect.R

@@ -60,6 +60,11 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
         hr_upr = exp(yhat + 1.96 * ses)
       )
 
+      # do not remove the rm() statement.
+      # ggplot saves fit_acm in its environment, which causes
+      # the file size of the current target to be massive.
+      rm(fit_acm)
+
       fig_eff_acm <- ggplot(fit_acm_data) +
         aes(x = time, y = hr_est, ymin = hr_lwr, ymax = hr_upr) +
         geom_ribbon(alpha = 0.2) +

### 2022-01-03 doc/manuscript.Rmd
methods from NP
comments and edits from Nick.

@@
 ---
-title: "Longer term All-Cause and Cardiovascular Mortality with Intensive Blood Pressure Control"
+title: "Longer term All-Cause and Cardiovascular Mortality with Intensive Blood Pressure Control: A Secondary Analysis of SPRINT"
 bibliography: refs.bib
@@
 
+both the The Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) trial have shown the benefits of intensive blood pressure control on cardiovascular morbidity and mortality [@sprint_research_group_randomized_2015; @zhang_trial_2021].
+
+__Reminder: bring in the recent BPTTC meta-analyses into the introduction__[@rahimi_age_stratified_2021].
+
 # METHODS
+
+__Trial Design__: The trial design and methods have been published previously [@ambrosius_design_2014].
+Briefly, we conducted a multicenter randomized clinical trial that compared two strategies for managing systolic BP (SBP) in older adults with hypertension who were at increased risk of cardiovascular disease.
+Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed.
+Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (defined by an estimated glomerular filtration rate of <60 mL/min/1.73 m^2^), or a Framingham Risk Score of 15% or greater or if they were aged 75 years or older.
+Individuals residing in a nursing home, persons with a diagnosis of dementia (based on medical record review), and those treated with medications primarily used for dementia therapy were excluded, as were persons with prevalent diabetes mellitus, history of stroke, proteinuria > 1 gram per day, or polycystic kidney disease.
+Individuals at 102 sites in the United States and Puerto Rico were randomized (1:1) to a SBP goal of less than 120 mm Hg (intensive treatment group, n = 4678) or a goal of less than 140 mm Hg (standard treatment group, n = 4683), using random permuted blocks with the randomization stratified by clinic site.
+The algorithms and formulary for the trial are listed in the published study protocol ( __citations needed__?
+).
+Trial enrollment began in November 2010 and ended in March 2013, with active follow-up through July 1, 2016.
+The study was approved by the institutional review board at each participating site, and each participant provided written informed consent.
+The study is registered at ClinicalTrials.gov (NCT01206062).
+
+__National Death Index Linkage__:
+
+__Baseline Study Measurements__: Sociodemographic data were collected at baseline, with race or ethnicity information collected via self-report.
+The estimated glomerular filtration rate (eGFR) was calculated by the four-variable Modification of Diet in Renal Disease study equation.
+Cognitive function was assessed using Montreal Cognitive Assessment (MoCA) [@nasreddine_montreal_2005].
+Lower cognitive function was defined as scoring 18 or lower (less than high school education) or 20 or lower (high school education or higher) on the MoCA.
+This roughly corresponds to the estimated normative 25th percentile at 80 years of age in the Irish Longitudinal Study of Aging [@kenny_normative_2013].
+We defined frailty status at baseline using a previously developed Frailty Index (FI) based upon the model of deficit accumulation [@pajewski_characterizing_2016].
+Briefly, the FI comprises a total of 36 items, and is calculated as the sum of the score for each deficit divided by the total number of nonmissing items.
+We categorized frailty status as fit (FI ≤ 0.10), less fit (0.10 < FI ≤ 0.21), or frail (FI > 0.21).
+
+__EHR Ancillary Study__: We examined the trajectory of systolic blood pressure (SBP) following the conclusion of the trial using outpatient SBPs extracted from the electronic health record (EHR).
+Methods for the linkage of participants to their medical record number and the extraction of vital sign data have been previously described [@drawz_concordance_2020].
+Because encounter type information was inconsistently available (i.e.
+outpatient, inpatient, observation, etc.), we defined a BP measurement as outpatient if there were was not a BP measurement on the preceding or following day, and if there were 2 or less BP measurements on a particular day.
+We averaged outpatient EHR BP readings when there were 2 on the same day.
 
@@
 The first approach split the follow-up of study participants into non-overlapping trial and cohort phases, then estimated the effect of randomization to intensive treatment in each phase [@zhang_time_varying_2018].
-The second approach estimated the cumulative time-dependent effect of randomization to intensive treatment [@martinussen2007dynamic].
+The second approach estimated a continuous time-dependent effect [@martinussen2007dynamic; @therneau2017using].
 While the phase-specific estimate of the treatment effect treats the trial and cohort phases independently, the cumulative estimate enforces continuity of the treatment effect through time so that the effect of treatment during the trial phase is not 'forgotten' during the cohort phase.
@@
 blank_plot <- ggplot(data.frame()) + aes() + theme_void()
-
 legend <- get_legend(

### 2022-01-03 R/ckd_epi_2021_compute.R
new 2021 eGFR equation
race-agnostic is preferred.

@@ -1,0 +1,28 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param age
+#' @param sex
+#' @param screat
+ckd_epi_2021_compute <- function(age, sex, screat) {
+
+  min_term <- case_when(
+    sex == 'Female' ~ pmin(screat / 0.7, 1),
+    sex == 'Male' ~ pmin(screat / 0.9, 1)
+  )
+
+  max_term <- case_when(
+    sex == 'Female' ~ pmax(screat / 0.7, 1),
+    sex == 'Male' ~ pmax(screat / 0.9, 1)
+  )
+
+  case_when(
+    sex == 'Female' ~
+      142*(min_term^(-0.241))*(max_term^(-1.200))*(0.9938^(age))*1.012,
+    sex == 'Male' ~
+      142*(min_term^(-0.302))*(max_term^(-1.200))*(0.9938^(age))
+  )
+
+}

### 2022-01-03 R/ndi_load.R

@@ -32,6 +32,7 @@ ndi_load <- function(sasinet_drive = "Z", fname, ...) {
       ckd = factor(eGFR_CKDEPI < 60,
                    levels = c(FALSE, TRUE),
                    labels = c("No", "Yes")),
+      egfr_2021 = ckd_epi_2021_compute(age, sex, screat),
       moca = factor(moca_status,
                     levels = c(0, 1),
                     labels = c("gt10th_percentile",

### 2022-01-03 _targets.R

@@ -24,6 +24,7 @@ list(
              pid,
              # study site
              randSite,
+             screat,
              # all cause mortality time and statue
              acm_years, acm_event,
              # cvd mortality status

### 2022-01-04 doc/manuscript.Rmd
new CKD epi eqn matches online calc

@@
 Byron C. Jaeger, PhD^1^ 
-Paul E. Drawz, DEGREES^2^ 
-Paul K. Whelton, DEGREES^3^ 
-Mark A. Supiano, DEGREES^4^ 
+Paul E. Drawz, MD^2^ 
+Paul K. Whelton, MD^3^ 
+Mark A. Supiano, MD^4^ 
 Adam P. Bress, PharmD, MS^5,6^
-Jeff D. Williamson, DEGREES^7^
-David M. Reboussin, DEGREES^1^, and
+Jeff D. Williamson, MD, MHS^7^
+David M. Reboussin, PhD^1^, and
 Nicholas M. Pajewski, PhD^1^
@@
 The second approach estimated a continuous time-dependent effect [@martinussen2007dynamic; @therneau2017using].
-While the phase-specific estimate of the treatment effect treats the trial and cohort phases independently, the cumulative estimate enforces continuity of the treatment effect through time so that the effect of treatment during the trial phase is not 'forgotten' during the cohort phase.
 All analyses accounted for correlation within study sites [@lin_robust_1989], and analyses of cardiovascular mortality accounted for the competing risk of non-cardiovascular mortality [@scheike_flexible_2008].
+The mean between-group differences in SBP following the conclusion of trial follow-up (after July 2016) were estimated using linear mixed models.
+Models included random effects for participant and clinic site.
+Our primary models included an interaction between treatment group and time since randomization, which was flexibly modeled using B-splines.
+
+All analyses were performed using SAS version 9.4 (SAS Institute Inc, Cary, NC) and R version `r paste(version$major, version$minor, sep = '.')` (R Project for Statistical Computing [http://www.r-project.org]) with multiple auxiliary R packages [@r_language; @table.glue; @tidyverse; @survival_package; @timereg; @targets].
+All hypothesis tests were 2-sided, and P values less than 0.05 were considered statistically significant.
+No adjustments for multiple comparisons were made.
 
 # RESULTS
+
+```{r}
+
+# fig$eff_acm[[1]]$data$hr_lwr
+
+```
 

### 2022-01-04 R/ckd_epi_2021_compute.R

@@ -26,3 +26,46 @@ ckd_epi_2021_compute <- function(age, sex, screat) {
   )
 
 }
+
+
+# TESTS comparing my eqn to the online calculator available here:
+# https://www.kidney.org/professionals/kdoqi/gfr_calculator
+
+# FORMAT for results below
+# test <number>
+# my_r_function(characteristics of test patient)
+#                my answer: <result using my function>
+# online calculator answer: <result from url above>
+
+# test 1
+# ckd_epi_2021_compute(age = 60, sex = 'Female', screat = 1.2)
+#                             MY EQN: 51.82133
+# CKD-EPI creatinine equation (2021): 52
+
+# test 2
+# ckd_epi_2021_compute(age = 60, sex = 'Male', screat = 1.2)
+#                             MY EQN: 69.23113
+# CKD-EPI creatinine equation (2021): 69
+
+# test 3
+# ckd_epi_2021_compute(age = 60, sex = 'Male', screat = 0.6)
+#                             MY EQN: 110.5115
+# CKD-EPI creatinine equation (2021): 111
+
+# test 4
+# ckd_epi_2021_compute(age = 70, sex = 'Male', screat = 0.6)
+#                             MY EQN: 103.8478
+# CKD-EPI creatinine equation (2021): 104
+
+# test 5
+# ckd_epi_2021_compute(age = 70, sex = 'Female', screat = 1)
+#                             MY EQN: 60.60605
+# CKD-EPI creatinine equation (2021): 61
+
+# test 6
+# ckd_epi_2021_compute(age = 70, sex = 'Female', screat = 1.5)
+#                             MY EQN: 37.25688
+# CKD-EPI creatinine equation (2021): 37
+
+
+

### 2022-01-04 R/ndi_base_viz_tv_effect.R

@@ -49,7 +49,7 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
 
       spline_x <- cbind(1, ns(x = plot_times, df=6))
 
-      yhat <- spline_x %*% coef(fit_acm)
+      yhat <- as.numeric(spline_x %*% coef(fit_acm))
 
       ses <- sqrt(diag(spline_x %*% vcov(fit_acm) %*% t(spline_x)))
 

### 2022-01-04 R/ndi_load.R
new CKD group; eGFR 2021

@@ -32,7 +32,13 @@ ndi_load <- function(sasinet_drive = "Z", fname, ...) {
       ckd = factor(eGFR_CKDEPI < 60,
                    levels = c(FALSE, TRUE),
                    labels = c("No", "Yes")),
-      egfr_2021 = ckd_epi_2021_compute(age, sex, screat),
+      # New creatinine based race-agnostic eGFR equation
+      egfr_2021 = ckd_epi_2021_compute(age = age,
+                                       sex = sex,
+                                       screat = screat),
+      ckd_2021 = factor(egfr_2021 < 60,
+                        levels = c(FALSE, TRUE),
+                        labels = c("No", "Yes")),
       moca = factor(moca_status,
                     levels = c(0, 1),
                     labels = c("gt10th_percentile",

### 2022-01-04 R/recoders_make.R

@@ -13,6 +13,7 @@ recoders_make <- function() {
                  "sex" = "Sex",
                  "race" = "Race",
                  "ckd" = "Chronic Kidney Disease",
+                 "ckd_2021" = "Chronic Kidney Disease",
                  "moca" = "Cognitive Function",
                  "frail" = "Frailty Status"),
 

### 2022-01-04 _targets.R

@@ -8,7 +8,7 @@ subgroups <- tibble(group = c("overall",
                               "age_cat",
                               "sex",
                               "race",
-                              "ckd",
+                              "ckd_2021",
                               "moca",
                               "frail"))
 
@@ -24,7 +24,6 @@ list(
              pid,
              # study site
              randSite,
-             screat,
              # all cause mortality time and statue
              acm_years, acm_event,
              # cvd mortality status

### 2022-01-04 R/get_sasinet_drive.R
no more manually switching sasinet drive location

@@ -1,0 +1,17 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+
+get_sasinet_drive <- function() {
+
+  username <- getElement(gitcreds::gitcreds_get(), 'username')
+
+  switch(
+    username,
+    'bcjaeger' = 'Z',
+    'nmpieyeskey' = 'O'
+  )
+
+}

### 2022-01-04 _targets.R

@@ -18,7 +18,7 @@ list(
 
   tar_target(
     ndi_baseline,
-    ndi_load(sasinet_drive = 'Z',
+    ndi_load(sasinet_drive = get_sasinet_drive(),
              fname = 'longterm_death.csv',
              # participant identifier
              pid,

### 2022-01-04 doc/manuscript.Rmd
first draft of results section

@@
 
-# fig$eff_acm[[1]]$data$hr_lwr
+acm_overall_epochs <- fig |> 
+  filter(variable == 'overall') |> 
+  pull(eff_acm) |> 
+  unlist(recursive = FALSE) |> 
+  getElement('data') |> 
+  filter(find_signchange_lgl(hr_upr-1)) |> 
+  mutate(label = table_value(time))
+
+t_signif <- acm_overall_epochs$label[1] 
+t_atten <- acm_overall_epochs$label[2]
 
@@
 The hazard ratio (HR) for all-cause mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during the cohort phase.
-The cumulative time-varying effect of intensive versus standard treatment indicated lower risk for all-cause mortality during the trial phase and was attenuated during the cohort phase (__Figure 1__).
+In subgroups based on age, sex, race, CKD, cognitive function, and frailty, there was no evidence that intensive treatment during the trial phase had benefit for all-cause mortality during the cohort phase (__eFigure 1__).
+The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 1__).
+
+```{r}
+
+cvd_overall_epochs <- fig |> 
+  filter(variable == 'overall') |> 
+  pull(eff_cvd) |> 
+  unlist(recursive = FALSE) |> 
+  getElement('data') |> 
+  slice(-1) |> 
+  filter(find_signchange_lgl(hr_upr-1)) |> 
+  mutate(label = table_value(time))
+
+t_signif <- cvd_overall_epochs$label[1] 
+t_atten  <- cvd_overall_epochs$label[2]
+
+```
+
+A total of `r ap_b_i$n_cvd` and `r ap_b_s$n_cvd` CVD mortality events occurred among participants randomized to intensive and standard treatment, respectively (__Table 2__).
+The HR for CVD mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the cohort phase.
+Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase had benefit for CVD mortality during the cohort phase (__eFigure 2__).
+The time-dependent effect of intensive versus standard treatment indicated a benefit for CVD mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort period (__Figure 2__).
+
+The estimated mean SBP among participants randomized to intensive treatment was <placeholder (95% CI)> at 5 years and <placeholder (95% CI)> at 10 years post-randomization (__Figure 3__).
+For participants randomized to standard treatment, mean SBP was estimated to be <placeholder (95% CI)> at 5 years and <placeholder (95% CI)> at 10 years post-randomization.
 
@@
 
-\newpage
-__Figure 4__: All-cause mortality hazard ratio for participants randomized to intensive versus standard treatment.
+```{r}
+block_section(
+  prop_section(
+    page_size = page_size(orient = "landscape"),
+    page_margins = page_mar(bottom = 1/8, top = 1/8, 
+                            right = 1/4, left = 1/4),
+    type = "continuous"
+  )
+)
+```
+
+Online supplement only
+
+Longer term All-Cause and Cardiovascular Mortality with Intensive Blood Pressure Control: A Secondary Analysis of SPRINT
+
+Byron C. Jaeger, PhD^1^ 
+Paul E. Drawz, MD^2^ 
+Paul K. Whelton, MD^3^ 
+Mark A. Supiano, MD^4^ 
+Adam P. Bress, PharmD, MS^5,6^
+Jeff D. Williamson, MD, MHS^7^
+David M. Reboussin, PhD^1^, and
+Nicholas M. Pajewski, PhD^1^
+
+<!-- Jackson T. Wright Jr., MD, PhD, Alfred K. Cheung, MD, William C. Cushman, MD, Cora E. Lewis, MD, Suzanne Oparil, MD, Michael V. Rocco, MD and Karen Johnson, MD -->
+
+^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC
+
+^2^Division of Renal Diseases & Hypertension, University of Minnesota, Minneapolis.
+
+^3^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
+
+^4^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
+
+^5^	Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT
+
+^6^	Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT
+
+^7^Section on Gerontology and Geriatric Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+
+```{r}
+block_section(
+  prop_section(
+    page_size = page_size(orient = "portrait"),
+    type = "continuous"
+  )
+)
+```
+
+__eFigure 1__: All-cause mortality hazard ratio for participants randomized to intensive versus standard treatment.
 
@@
 \newpage
-__Figure 5__: Cardiovascular mortality hazard ratio for participants randomized to intensive versus standard treatment.
+__eFigure 2__: Cardiovascular mortality hazard ratio for participants randomized to intensive versus standard treatment.
 
@@
 ```
-
-# SUPPLEMENT
-
-To be decided.
 

### 2022-01-04 R/find_signchange_int.R

@@ -1,0 +1,11 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param x
+find_signchange_int <- function(x){
+
+  which(c(0,diff(sign(x))) != 0)
+
+}

### 2022-01-04 R/find_signchange_lgl.R

@@ -1,0 +1,14 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param x
+find_signchange_lgl <- function(x){
+
+  c(0,diff(sign(x))) != 0
+
+}
+
+
+

### 2022-01-06 R/get_sasinet_drive.R
NMPrev010522

@@ -4,14 +4,14 @@
 #'
 #' @title
 
-get_sasinet_drive <- function() {
+get_sasinet_drive <- function(username = 'bcjaeger') {
 
-  username <- getElement(gitcreds::gitcreds_get(), 'username')
+  #username <- getElement(gitcreds::gitcreds_get(), 'username')
 
   switch(
     username,
     'bcjaeger' = 'Z',
-    'nmpieyeskey' = 'O'
+    'nmpieyeskey' = 'O:/sprint'
   )
 
 }

### 2022-01-06 R/ndi_load.R

@@ -7,7 +7,7 @@
 ndi_load <- function(sasinet_drive = "Z", fname, ...) {
 
   path <- file.path(
-    paste0(sasinet_drive, ":"),
+    sasinet_drive,
     "npajewski",
     "NDI",
     "Data",

### 2022-01-06 doc/manuscript.Rmd

--- 2022-01-04 doc/manuscript.Rmd
@@
 ---
+```{r tp1, include=FALSE}
+single_par <- fp_par(line_spacing=1)
+```
 
-Byron C. Jaeger, PhD^1^ 
-Paul E. Drawz, MD^2^ 
-Paul K. Whelton, MD^3^ 
-Mark A. Supiano, MD^4^ 
-Adam P. Bress, PharmD, MS^5,6^
-Jeff D. Williamson, MD, MHS^7^
-David M. Reboussin, PhD^1^, and
-Nicholas M. Pajewski, PhD^1^
+Byron C.
+Jaeger, PhD,^1^ Adam P.
+Bress, PharmD, MS,^2,3^ Alfred K.
+Cheung, MD,^4,5^ William C.
+Cushman, MD,^6^ Paul E.
+Drawz, MD,^7^ Lawrence J.
+Fine, MD,^8^ Karen C.
+Johnson, MD,^6^ Cora E.
+Lewis, MD,^9^ Suzanne Oparil, MD,^10^ Michael V.
+Rocco, MD,^11^ Mark A.
+Supiano, MD,^12^ Paul K.
+Whelton, MD,^13^ Jeff D.
+Williamson, MD, MHS,^14^ Jackson T.
+Wright, Jr, MD, PhD,^15^ David M.
+Reboussin, PhD,^1^ and Nicholas M.
+Pajewski, PhD,^1^
 
-<!-- Jackson T. Wright Jr., MD, PhD, Alfred K. Cheung, MD, William C. Cushman, MD, Cora E. Lewis, MD, Suzanne Oparil, MD, Michael V. Rocco, MD and Karen Johnson, MD -->
+^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC.
+^2^Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT.
+^3^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT.
+^4^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT.
+^5^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT.
+^6^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis.
+^7^Division of Renal Diseases & Hypertension, University of Minnesota, MN.
+^8^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD.
+^9^Department of Epidemiology, University of Alabama at Birmingham.
+^10^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham.
+^11^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+^12^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
+^13^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
+^14^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+^15^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
+\ \
 
-^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC
+**Correspondence to:** \
+Nicholas M. Pajewski, PhD \
+Department of Biostatistics and Data Science \
+Division of Public Health Sciences \
+Wake Forest School of Medicine \
+Medical Center Boulevard \
+Winston-Salem, NC 27154 \
+(336) 713-1396 \
+npajewsk@wakehealth.edu `r single_par`
 
-^2^Division of Renal Diseases & Hypertension, University of Minnesota, Minneapolis.
-
-^3^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
-
-^4^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
-
-^5^	Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT
-
-^6^	Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT
-
-^7^Section on Gerontology and Geriatric Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+**Word Count:** Abstract: ; Manuscript:
 
@@
 
-__Background__.
+__IMPORTANCE__.
 Both the Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) trial have shown benefits of intensive blood pressure control on cardiovascular morbidity and mortality.
-However, as both trials were ended after slightly more than 3 years of follow-up, the legacy effect of intensive treatment on longer term cardiovascular and all-cause mortality is unknown.
+However, as both trials were ended after slightly more than 3 years of follow-up, the effect of intensive treatment on longer term mortality is unknown.
 
-__Methods__.
-We linked participants in SPRINT to the National Death Index (NDI, 2016 to 2020), assessing cardiovascular mortality using the NDI plus system.
-Cox and competing risk regression models were used to model the effect of intensive treatment through the trial and during post-trial follow-up on all-cause and cardiovascular mortality respectively
+__OBJECTIVE__. To evaluate the legacy effect of intensive hypertension treatment on longer term all-cause and cardiovascular mortality.
+
+__DESIGN, SETTING, AND PARTICIPANTS__.
+SPRINT, a randomized clinical trial of 9361 patients aged 50 years or older with hypertension and increased cardiovascular risk, but without diabetes or history of stroke.
+Randomization began on November 8, 2010, the trial intervention was stopped early on August 20, 2015, and trial follow-up visits occurred through July 2016.
+
+__INTERVENTIONS__. Randomization to a systolic blood pressure goal of either less than 120 mm Hg (intensive treatment, n=4678) versus less than 140 mm Hg (standard treatment, N=4683).
+
+__MAIN OUTCOMES AND MEASURES__.
+Cardiovascular and all-cause mortality assessed via the US National Death Index, beginning in 2016 through December 31st, 2020.
+Outpatient blood pressures measured in routine clinical practice were examined in a subset of trial participants (N=3644).
 
@@
 
-__Results__.
-Over a median follow-up of `r ap_b_o$fup_est` years, there were `r ap_b_i$n_cvd` and `r ap_b_i$n_acm` cardiovascular and all-cause deaths with intensive treatment respectively, and `r ap_b_s$n_cvd` cardiovascular / `r ap_b_s$n_acm` all-cause deaths for standard treatment.
-Intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial (follow-up through July 2016).
+__RESULTS__.
+Among 9361 randomized participants (mean age, 67.9 years; 3332 women [35.6%]), the median intervention period was 3.34 years.
+Over a median follow-up of `r ap_b_o$fup_est` years, intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial (follow-up through July 2016).
 However, there was no indication of benefit during post-trial follow-up for either cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
 Results were similar for subgroups based on baseline age, cognitive function, and frailty status.
+Analyses of outpatient blood pressures indicated a steady decline in the mean between group difference following the trial, largely driven by increases in mean systolic blood pressure in participnts randomized to intensive treatment, increasing from a mean of X at 4.5 years of follow-up to Y at 9 years of follow-up.
 
-__Conclusions__.
-Our results show a clear benefit for cardiovascular and all-cause mortality during the trial which was largely attenuated during post-trial observational follow-up.
-Given indications of increasing blood pressures in SPRINT participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension in line with current guidelines.
+there were `r ap_b_i$n_cvd` and `r ap_b_i$n_acm` cardiovascular and all-cause deaths with intensive treatment respectively, and `r ap_b_s$n_cvd` cardiovascular / `r ap_b_s$n_acm` all-cause deaths for standard treatment.
+
+__CONCLUSIONS AND RELEVANCE__.
+The observed benefit of intensive treatment on cardiovascular and all-cause mortality was largely attenuated during post-trial observational follow-up.
+Given increasing blood pressures in participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension.
+
 \newpage
-
-<!-- END OF ABSTRACT -->
 
@@
 both the The Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) trial have shown the benefits of intensive blood pressure control on cardiovascular morbidity and mortality [@sprint_research_group_randomized_2015; @zhang_trial_2021].
+However, both trials were stopped early after a median follow-up of 3 years.
 
 __Reminder: bring in the recent BPTTC meta-analyses into the introduction__[@rahimi_age_stratified_2021].
+
+Here we examined the longer term effect of randomization to intensive treatment on through linkage to the National Death Index (NDI).
 
@@
 __Trial Design__: The trial design and methods have been published previously [@ambrosius_design_2014].
-Briefly, we conducted a multicenter randomized clinical trial that compared two strategies for managing systolic BP (SBP) in older adults with hypertension who were at increased risk of cardiovascular disease.
+Briefly, we conducted a multicenter randomized clinical trial that compared two strategies for managing systolic BP (SBP) in older adults with hypertension who were at increased risk of cardiovascular disease (CVD).
 Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed.
@@
 
-__National Death Index Linkage__:
-
 __Baseline Study Measurements__: Sociodemographic data were collected at baseline, with race or ethnicity information collected via self-report.
-The estimated glomerular filtration rate (eGFR) was calculated by the four-variable Modification of Diet in Renal Disease study equation.
-Cognitive function was assessed using Montreal Cognitive Assessment (MoCA) [@nasreddine_montreal_2005].
+The estimated glomerular filtration rate (eGFR) was calculated by the race-free 2021 CKD-EPI creatinine equation.[@inker_new_2021] Cognitive function was assessed using Montreal Cognitive Assessment (MoCA) [@nasreddine_montreal_2005].
 Lower cognitive function was defined as scoring 18 or lower (less than high school education) or 20 or lower (high school education or higher) on the MoCA.
-This roughly corresponds to the estimated normative 25th percentile at 80 years of age in the Irish Longitudinal Study of Aging [@kenny_normative_2013].
+This roughly corresponds to the estimated normative 10th percentile in the Irish Longitudinal Study of Aging [@kenny_normative_2013].
 We defined frailty status at baseline using a previously developed Frailty Index (FI) based upon the model of deficit accumulation [@pajewski_characterizing_2016].
@@
 We categorized frailty status as fit (FI ≤ 0.10), less fit (0.10 < FI ≤ 0.21), or frail (FI > 0.21).
+
+__National Death Index Linkage__: Outcomes of interest included all-cause and CVD mortality.
+Methods of ascertainment and adjudication through the course of trial follow-up have been previously described.
+Subsequently, mortality was ascertained through a US National Death Index (NDI) search.
+Possible matches were identified according to NDI guidelines.
+To be considered a confirmed death, we required 4 or more of 5 matches among Social Security number, name, date of birth, city, and state in the NDI.
+NDI follow-up began in 2016 and ended on the date of death or date of the NDI search (2020).
+Deaths ascertained in 2020 were based on the preliminary data release.
+CVD mortality for NDI-based follow-up used the NDI Plus System, which automatically identifies underlying causes of death from death certificates, including conversion to ICD-10 codes.
+we defined CVD mortality as any death containing the ICD-10 codes of I00 to I99.
 
@@
 
+A striking aspect of our results is the quickly weakening level of BP control for participants randomized to intensive treatment.
+While an attenuation of the between-group BP delta subsequent to the trial was certainly expected, one hypothesis was that such an attenuation would be driven by participants randomized to standard treatment pursuing a lower BP goal, given the results of SPRINT and subsequent changes to hypertension guidelines.
+However, this is clearly not what occurred.
+While we do not have access to prescription records to know how participant medication regimes may have changed after the trial, these results likely show some contribution of clinician therapeutic inertia, which has been identified as a significant barrier to improving population level control of hypertension.
+Combined with evidence showing recent decreases in the population of prevalence of controlled hypertension in the US,[@muntner2020htn] our results highlight the sobering reality facing the hypertension community.
+Sustainability is a clear limiting factor, especially with trying to implement lower BP goals and interventions earlier in adulthood.
+
+This study has several limitations.
+First, while we restricted analyses to high quality NDI matches, some small degree of misclassification in linking participants to the NDI is likely.
+Second, while several studies have shown reasonable performance of using NDI diagnosis codes for defining CVD mortality, it is clearly not as robust as the adjudication process used in the primary follow-up for the trial.
+Third, information about BP control after the trial was limited to routine outpatient BPs extracted from the EHR, which are well known to poorly reflect the standardized BP measurement process used during the trial.[@drawz_concordance_2020]
+
+# ACKNOWLEDGMENTS 
+The views expressed in this paper are those of the authors and do not represent the official position of the National Institutes of Health (NIH), the National Heart, Lung, and Blood Institute, the Department of Veterans Affairs, or the U.S.
+Government, or the SPRINT Research Group.
+This paper was not reviewed by the SPRINT Publications and Presentations Committee.
+The authors also wish to acknowledge computing support provided the Veterans Affairs Informatics and Computing Infrastructure (VINCI).
+
+# FINANCIAL DISCLOSURE 
+The work presented here was supported by R01HL136679, R01AG055606, and R01AG065805.
+The Systolic Blood Pressure Intervention Trial was funded by the National Institutes of Health (including the National Heart, Lung, and Blood Institute, the National Institute of Diabetes and Digestive and Kidney Diseases, the National Institute on Aging, and the National Institute of Neurological Disorders and Stroke) under contracts HHSN268200900040C, HHSN268200900046C, HHSN268200900047C, HHSN268200900048C, and HHSN268200900049C and interagency agreement A-HL-13-002-001.
+It was also supported in part with resources and use of facilities through the Department of Veterans Affairs.
+Azilsartan and chlorthalidone (combined with azilsartan) were provided by Takeda Pharmaceuticals International Inc.
+
+Additional support was provided through the following National Center for Advancing Translational Sciences clinical and translational science awards: UL1TR000439 (awarded to Case Western Reserve University); UL1RR025755 (Ohio State University); UL1RR024134 and UL1TR000003 (University of Pennsylvania); UL1RR025771 (Boston University); UL1TR000093 (Stanford University); UL1RR025752, UL1TR000073, and UL1TR001064 (Tufts University); UL1TR000050 (University of Illinois); UL1TR000005 (University of Pittsburgh); U54TR000017-06 (University of Texas Southwestern Medical Center); UL1TR000105-05 (University of Utah); UL1 TR000445 (Vanderbilt University); UL1TR000075 (George Washington University); UL1 TR000002 (University of California, Davis); UL1 TR000064 (University of Florida); and UL1TR000433 (University of Michigan); and by National Institute of General Medical Sciences, Centers of Biomedical Research Excellence award NIGMS P30GM103337 (awarded to Tulane University).
+
+\newpage 
 ```{r}
@@
   font(fontname = 'Times New Roman', part = 'all')
+
+tbl_object <- add_footer_lines(tbl_object, values="Abbreviations: CI, confidence interval; FI, frailty index.
+aChronic Kidney Disease defined as an estimated glomerular filtration rate <60 ml/min/1.73 m2 based on the 2021 CKD-EPI creatinine equation.")
 
@@
 knitr::include_graphics(
-  path = 'Z:/bjaeger/sprint-mortality-ndi/fig/SPRINT_EHR_SBP_LongTerm_Comparison_122721.png', 
+  path = 'O:/sprint/npajewski/sprint-mortality-ndi/fig/SPRINT_EHR_SBP_LongTerm_Comparison_122721.png', 
   dpi = 600

### 2022-01-06 doc/manuscript.Rmd

@@
 ---
-```{r tp1, include=FALSE}
+```{r setup, include=FALSE}
+
+knitr::opts_chunk$set(echo = FALSE, 
+                      message = FALSE,
+                      warning = FALSE,
+                      dpi = 600, 
+                      fig.width = 11, 
+                      fig.height = 6.25)
+
+# specification for table tables
+rspec <- round_spec() |>
+  round_using_magnitude(digits = c(2, 1, 0),
+                        breaks = c(10, 100, Inf))
+
+# save it to options:
+names(rspec) <- paste('table.glue', names(rspec), sep = '.')
+options(rspec)
+
+# default is double space 
+# single spacing for title page, abstract, captions
 single_par <- fp_par(line_spacing=1)
+
+# make the . symbol equivalent to calling table_glue.
+# Another way to makes inline insertion of results less text-heavy
+. <- table.glue::table_glue
+
 ```
@@
 Supiano, MD,^12^ Paul K.
-Whelton, MD,^13^ Jeff D.
+Whelton, MD,^13^ Joshua D.
+Bundy, PhD,^13^ Jeff D.
 Williamson, MD, MHS,^14^ Jackson T.
@@
 
-```{r setup, include=FALSE}
-
-knitr::opts_chunk$set(echo = FALSE, 
-                      message = FALSE,
-                      warning = FALSE,
-                      dpi = 600, 
-                      fig.width = 11, 
-                      fig.height = 6.25)
-
-# specification for table tables
-rspec <- round_spec() |>
-  round_using_magnitude(digits = c(2, 1, 0),
-                        breaks = c(10, 100, Inf))
-
-# save it to options:
-names(rspec) <- paste('table.glue', names(rspec), sep = '.')
-options(rspec)
-
-```
-
 ```{r load-targets, include=FALSE}
@@
 
+# ap makes inline insertion of results less text-heavy.
+# ap = access point, 
+# b = baseline, l = longitudinal
+# o = overall, i = intensive, s = standard,
+# t = trial, c = cohort
+
+ap_b_o <- inline$base_ds$overall$overall$overall
+ap_b_i <- inline$base_ds$overall$overall$Intensive
+ap_b_s <- inline$base_ds$overall$overall$Standard
+
+ap_l_o_t <- inline$long_inf$overall$overall$trial
+ap_l_o_c <- inline$long_inf$overall$overall$cohort
+
 ```
@@
 Outpatient blood pressures measured in routine clinical practice were examined in a subset of trial participants (N=3644).
-
-```{r}
-
-# ap makes inline insertion of results less text-heavy.
-# ap = access point, 
-# b = baseline, l = longitudinal
-# o = overall, i = intensive, s = standard,
-# t = trial, c = cohort
-
-ap_b_o <- inline$base_ds$overall$overall$overall
-ap_b_i <- inline$base_ds$overall$overall$Intensive
-ap_b_s <- inline$base_ds$overall$overall$Standard
-
-ap_l_o_t <- inline$long_inf$overall$overall$trial
-ap_l_o_c <- inline$long_inf$overall$overall$cohort
-
-# make the . symbol equivalent to calling table_glue.
-# Another way to makes inline insertion of results less text-heavy
-. <- table.glue::table_glue
-
-```
 
@@
 In both treatment groups, median follow-up time was `r ap_b_o$fup_est` years.
-A total of `r ap_b_i$n_acm` and `r ap_b_s$n_acm` all-cause mortality events occurred among participants randomized to intensive and standard treatment, respectively (__Table 1__).
+A total of `r ap_b_i$n_acm` and `r ap_b_s$n_acm` all-cause mortality events occurred among participants randomized to intensive and standard treatment, respectively (__Table 2__).
 The hazard ratio (HR) for all-cause mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during the cohort phase.
-The cumulative time-varying effect of intensive versus standard treatment indicated lower risk for all-cause mortality during the trial phase and was attenuated during the cohort phase (__Figure 1__).
+In subgroups based on age, sex, race, CKD, cognitive function, and frailty, there was no evidence that intensive treatment during the trial phase had benefit for all-cause mortality during the cohort phase (__eFigure 1__).
+The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 1__).
+
+```{r}
+
+cvd_overall_epochs <- fig |> 
+  filter(variable == 'overall') |> 
+  pull(eff_cvd) |> 
+  unlist(recursive = FALSE) |> 
+  getElement('data') |> 
+  slice(-1) |> 
+  filter(find_signchange_lgl(hr_upr-1)) |> 
+  mutate(label = table_value(time))
+
+t_signif <- cvd_overall_epochs$label[1] 
+t_atten  <- cvd_overall_epochs$label[2]
+
+```
+
+A total of `r ap_b_i$n_cvd` and `r ap_b_s$n_cvd` CVD mortality events occurred among participants randomized to intensive and standard treatment, respectively (__Table 3__).
+The HR for CVD mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the cohort phase.
+Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase had benefit for CVD mortality during the cohort phase (__eFigure 2__).
+The time-dependent effect of intensive versus standard treatment indicated a benefit for CVD mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort period (__Figure 2__).
+
+The estimated mean SBP among participants randomized to intensive treatment was <placeholder (95% CI)> at 5 years and <placeholder (95% CI)> at 10 years post-randomization (__Figure 3__).
+For participants randomized to standard treatment, mean SBP was estimated to be <placeholder (95% CI)> at 5 years and <placeholder (95% CI)> at 10 years post-randomization.
 
@@
 
+```{r tbl_footers}
+
+ckd_defn <- as_paragraph("Chronic Kidney Disease was defined by an estimated glomerular filtration rate <60 ml/min/1.73 m", as_sup(2L), " based on the 2021 CKD-EPI creatinine equation.")
+
+```
+
+```{r tbl_abbrvs}
+
+abbrvs <- list(
+  CI   = 'confidence interval',
+  FI   = 'frailty index',
+  SD   = 'standard deviation',
+  IQR  = 'interquartile range',
+  MoCA = 'Montreal cognitive assessment',
+  EHR  = 'electronic health records'
+)
+
+abbrvs_paste <- function(strings){
+
+  strings[1] <- str_replace(strings[1], ' ', ' represents ')
+
+  glue::glue_collapse(strings, sep = ', ', last = ' and ')
+
+}
+
+abbrvs_write <- function(abbr){
+
+  sorted <- abbr[sort(names(abbr))]
+
+  strings <- map2_chr(names(sorted), sorted, paste)
+
+  as_paragraph(abbrvs_paste(strings))
+
+}
+
+```
+
 \newpage
+__Table 1__: Characteristics of study participants overall and stratified by inclusion in the electronic health record ancillary study. `r single_par`
+
+```{r}
+
+tar_load(tbl_characteristics)
+
+header_labels <- tbl_characteristics |> 
+  show_header_names(quiet = TRUE) |> 
+  mutate(
+    label = str_remove_all(label, fixed("*")),
+    label = str_replace(label, ', ', '\n')
+  ) |> 
+  deframe()
+
+tbl_characteristics |> 
+  modify_footnote(update = everything() ~ NA) |> 
+  as_flex_table() |> 
+  add_header_row(
+    values = c("Characteristic",
+               header_labels['stat_0'], 
+               "Included in ancillary EHR study", 
+               header_labels['p.value']),
+    colwidths = c(1, 1, 2, 1)
+  ) |> 
+  theme_box() |> 
+  set_header_labels(stat_0 = header_labels['stat_0'],
+                    stat_1 = header_labels['stat_1'],
+                    stat_2 = header_labels['stat_2']) |> 
+  merge_v(part = 'header') |> 
+  align(j = c(2, 3, 4, 5), align = 'center', part = 'all') |> 
+  footnote(
+    i = 1, 
+    j = 1, 
+    ref_symbols = '',
+    value = as_paragraph(
+      "Table values are percentages unless otherwise specified"
+    )
+  ) |> 
+  footnote(
+    i = 1, 
+    j = ~p.value,
+    part = 'header',
+    ref_symbols = 'a',
+    value = as_paragraph(
+      "P-values computed using Wilcoxon rank sum test for age in years, chi-square test for age \u226575 years, and chi-square test for all other variables"
+    )
+  ) |> 
+  footnote(
+    i = ~ label == "Chronic Kidney Disease",
+    j = 1, 
+    ref_symbols = 'b',
+    value = ckd_defn,
+    part = 'body'
+  ) |> 
+  footnote(
+    i = 1,
+    j = 1,
+    ref_symbols = '',
+    value = abbrvs_write(abbrvs[c('SD', 'EHR', 'IQR', 'MoCA', 'FI')]),
+    part = 'body'
+  ) |>
+  fontsize(size = 12, part = 'all') |> 
+  font(fontname = 'Times New Roman', part = 'all')
+
+```
+
+<!-- new page not needed at the end of a block section -->
 ```{r}
@@
 
-__Table 1__: All-cause mortality by treatment group and subgroup
+__Table 2__: All-cause mortality by treatment group and subgroup `r single_par`
 
@@
 
-tbl_object <- add_footer_lines(tbl_object, values="Abbreviations: CI, confidence interval; FI, frailty index.
-aChronic Kidney Disease defined as an estimated glomerular filtration rate <60 ml/min/1.73 m2 based on the 2021 CKD-EPI creatinine equation.")
+tbl_object <- add_footer_lines(tbl_object, values="Abbreviations: CI, confidence interval; FI, frailty index.")
 
@@
 \newpage
-__Table 2__: Cardiovascular and non-cardiovascular mortality by treatment group and subgroup
+__Table 3__: Cardiovascular and non-cardiovascular mortality by treatment group and subgroup `r single_par`
 
@@
 \newpage
-__Figure 1__: (A) Cumulative incidence of all-cause mortality by treatment group. (B) Time-dependent effect of randomization to intensive treatment for all-cause mortality.
+__Figure 1__: (A) Cumulative incidence of all-cause mortality by treatment group. (B) Time-dependent effect of randomization to intensive treatment for all-cause mortality. `r single_par`
 
@@
 (B) Time-dependent effect of randomization to intensive treatment for cardiovascular mortality.
+`r single_par`
 
@@
 \newpage
-__Figure 3__: Mean systolic blood pressure over time by treatment group. Shaded areas indicate a 95% confidence interval for the mean.
+__Figure 3__: Mean systolic blood pressure over time by treatment group. Shaded areas indicate a 95% confidence interval for the mean. `r single_par`
 
@@
 
+fpath <- file.path(
+  paste0(get_sasinet_drive(), ":"),
+  'bjaeger',
+  'sprint-mortality-ndi',
+  'fig',
+  'SPRINT_EHR_SBP_LongTerm_Comparison_122721.png'
+)
+
 knitr::include_graphics(
-  path = 'O:/sprint/npajewski/sprint-mortality-ndi/fig/SPRINT_EHR_SBP_LongTerm_Comparison_122721.png', 
+  path = fpath, 
   dpi = 600

### 2022-01-07 doc/manuscript.Rmd
address #1 and #2 with bp fig in targets

@@
 bibliography: refs.bib
-csl: lancet.csl
+csl: jama.csl
 always_allow_html: true
@@
 Reboussin, PhD,^1^ and Nicholas M.
-Pajewski, PhD,^1^
+Pajewski, PhD^1^
 
@@
 ^15^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
-\ \
+\
 
@@
   MoCA = 'Montreal cognitive assessment',
-  EHR  = 'electronic health records'
+  EHR  = 'electronic health records',
+  CVD  = 'cardiovascular disease'
 )
@@
 
-  strings[1] <- str_replace(strings[1], ' ', ' represents ')
+  front <- 'Abbreviations'
 
-  glue::glue_collapse(strings, sep = ', ', last = ' and ')
+  matter <- glue::glue_collapse(strings, sep = '; ', last = '; and ')
+
+  paste(front, matter, sep = ': ')
 
@@
 
-  strings <- map2_chr(names(sorted), sorted, paste)
+  strings <- map2_chr(names(sorted), sorted, paste, sep = ' = ')
 
@@
 \newpage
-__Table 1__: Characteristics of study participants overall and stratified by inclusion in the electronic health record ancillary study. `r single_par`
+__Table 1__: Characteristics of study participants stratified by inclusion in the electronic health record ancillary study. `r single_par`
 
@@
 header_labels <- tbl_characteristics |> 
+  getElement('main') |> 
   show_header_names(quiet = TRUE) |> 
@@
 tbl_characteristics |> 
+  getElement('main') |> 
   modify_footnote(update = everything() ~ NA) |> 
@@
     values = c("Characteristic",
-               header_labels['stat_0'], 
                "Included in ancillary EHR study", 
                header_labels['p.value']),
-    colwidths = c(1, 1, 2, 1)
+    colwidths = c(1, 2, 1)
   ) |> 
   theme_box() |> 
-  set_header_labels(stat_0 = header_labels['stat_0'],
-                    stat_1 = header_labels['stat_1'],
+  set_header_labels(stat_1 = header_labels['stat_1'],
                     stat_2 = header_labels['stat_2']) |> 
   merge_v(part = 'header') |> 
-  align(j = c(2, 3, 4, 5), align = 'center', part = 'all') |> 
+  align(j = c(2, 3, 4), align = 'center', part = 'all') |> 
   footnote(
@@
   fontsize(size = 12, part = 'all') |> 
-  font(fontname = 'Times New Roman', part = 'all')
-
-tbl_object <- add_footer_lines(tbl_object, values="Abbreviations: CI, confidence interval; FI, frailty index.")
+  font(fontname = 'Times New Roman', part = 'all') |> 
+  footnote(i = ~ variable == 'Chronic Kidney Disease', 
+           j = 1,
+           part = 'body',
+           ref_symbols = 'a',
+           value = ckd_defn) |> 
+  footnote(i = 1, 
+           j = 1, 
+           ref_symbols = '',
+           value = abbrvs_write(abbrvs[c('CI', 'FI')]))
 
@@
   fontsize(size = 12, part = 'all') |> 
-  font(fontname = 'Times New Roman', part = 'all')
+  font(fontname = 'Times New Roman', part = 'all') |> 
+  footnote(i = ~ variable == 'Chronic Kidney Disease', 
+           j = 1,
+           part = 'body',
+           ref_symbols = 'a',
+           value = ckd_defn) |> 
+  footnote(i = 1, 
+           j = 1, 
+           ref_symbols = '',
+           value = abbrvs_write(abbrvs[c('CI', 'CVD', 'FI')]))
 
@@
 blank_plot <- ggplot(data.frame()) + aes() + theme_void()
+
 legend <- get_legend(
@@
 \newpage
-__Figure 3__: Mean systolic blood pressure over time by treatment group. Shaded areas indicate a 95% confidence interval for the mean. `r single_par`
+__Figure 3__: (A) Mean systolic blood pressure over time by treatment group.
+(B) Mean difference in systolic blood pressure over time.
+Shaded areas indicate a 95% confidence interval for the mean.
+Horizontal lines during the trial period show target blood pressure values for each treatment group `r single_par`
 
@@
 
-fpath <- file.path(
-  paste0(get_sasinet_drive(), ":"),
-  'bjaeger',
-  'sprint-mortality-ndi',
-  'fig',
-  'SPRINT_EHR_SBP_LongTerm_Comparison_122721.png'
+tar_load(bp_viz)
+
+blank_plot <- ggplot(data.frame()) + aes() + theme_void()
+
+legend <- get_legend(
+  bp_viz$levels + 
+    theme(legend.position = c(.55, .2),
+          legend.direction = 'horizontal')
 )
 
-knitr::include_graphics(
-  path = fpath, 
-  dpi = 600
+plot_grid(
+  # row 1 (legend)
+  legend, 
+  blank_plot,
+  # row 2 (figures)
+  bp_viz$levels + theme(legend.position = ''), 
+  bp_viz$diff,
+  ncol = 2,
+  align = 'v',
+  axis = 'lr',
+  rel_heights = c(1,10,2),
+  labels = c("","","A","B","",""),
+  label_x = 0.1,
+  label_y = 1.05
 )
+
+# fpath <- file.path(
+#   paste0(get_sasinet_drive(), ":"),
+#   'bjaeger',
+#   'sprint-mortality-ndi',
+#   'fig',
+#   'SPRINT_EHR_SBP_LongTerm_Comparison_122721.png'
+# )
+# 
+# knitr::include_graphics(
+#   path = fpath, 
+#   dpi = 600
+# )
 
@@
 
-Longer term All-Cause and Cardiovascular Mortality with Intensive Blood Pressure Control: A Secondary Analysis of SPRINT
+Byron C.
+Jaeger, PhD,^1^ Adam P.
+Bress, PharmD, MS,^2,3^ Alfred K.
+Cheung, MD,^4,5^ William C.
+Cushman, MD,^6^ Paul E.
+Drawz, MD,^7^ Lawrence J.
+Fine, MD,^8^ Karen C.
+Johnson, MD,^6^ Cora E.
+Lewis, MD,^9^ Suzanne Oparil, MD,^10^ Michael V.
+Rocco, MD,^11^ Mark A.
+Supiano, MD,^12^ Paul K.
+Whelton, MD,^13^ Joshua D.
+Bundy, PhD,^13^ Jeff D.
+Williamson, MD, MHS,^14^ Jackson T.
+Wright, Jr, MD, PhD,^15^ David M.
+Reboussin, PhD,^1^ and Nicholas M.
+Pajewski, PhD^1^
 
-Byron C. Jaeger, PhD^1^ 
-Paul E. Drawz, MD^2^ 
-Paul K. Whelton, MD^3^ 
-Mark A. Supiano, MD^4^ 
-Adam P. Bress, PharmD, MS^5,6^
-Jeff D. Williamson, MD, MHS^7^
-David M. Reboussin, PhD^1^, and
-Nicholas M. Pajewski, PhD^1^
+^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC.
+^2^Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT.
+^3^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT.
+^4^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT.
+^5^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT.
+^6^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis.
+^7^Division of Renal Diseases & Hypertension, University of Minnesota, MN.
+^8^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD.
+^9^Department of Epidemiology, University of Alabama at Birmingham.
+^10^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham.
+^11^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+^12^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
+^13^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
+^14^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+^15^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
+\
 
-<!-- Jackson T. Wright Jr., MD, PhD, Alfred K. Cheung, MD, William C. Cushman, MD, Cora E. Lewis, MD, Suzanne Oparil, MD, Michael V. Rocco, MD and Karen Johnson, MD -->
-
-^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC
-
-^2^Division of Renal Diseases & Hypertension, University of Minnesota, Minneapolis.
-
-^3^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
-
-^4^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
-
-^5^	Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT
-
-^6^	Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT
-
-^7^Section on Gerontology and Geriatric Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+**Correspondence to:** \
+Nicholas M. Pajewski, PhD \
+Department of Biostatistics and Data Science \
+Division of Public Health Sciences \
+Wake Forest School of Medicine \
+Medical Center Boulevard \
+Winston-Salem, NC 27154 \
+(336) 713-1396 \
+npajewsk@wakehealth.edu `r single_par`
 
@@
 ```
+
+__eTable 1__: Characteristics of study participants stratified by inclusion in the electronic health record ancillary study and by treatment group. `r single_par`
+
+```{r}
+
+header_labels <- tbl_characteristics |> 
+  getElement('supp') |> 
+  show_header_names(quiet = TRUE) |> 
+  mutate(
+    label = str_remove_all(label, fixed("*")),
+    label = str_replace(label, ', ', '\n')
+  ) |> 
+  deframe()
+
+tbl_characteristics |> 
+  getElement('supp') |> 
+  modify_footnote(update = everything() ~ NA) |> 
+  as_flex_table() |> 
+  theme_box() |> 
+  width(width = 0.95) |> 
+  width(j=1, width = 1.75) |> 
+  set_header_labels(stat_1_1 = header_labels['stat_1_1'],
+                    stat_1_2 = header_labels['stat_1_2'],
+                    stat_2_1 = header_labels['stat_2_1'],
+                    stat_2_2 = header_labels['stat_2_2']) |> 
+  align(j = 2:7, align = 'center', part = 'all') |> 
+  footnote(
+    i = 1, 
+    j = 1, 
+    ref_symbols = '',
+    value = as_paragraph(
+      "Table values are percentages unless otherwise specified"
+    )
+  ) |> 
+  footnote(
+    i = 2, 
+    j = ~ p.value_1 + p.value_2,
+    part = 'header',
+    ref_symbols = 'a',
+    value = as_paragraph(
+      "P-values computed using Wilcoxon rank sum test for age in years, chi-square test for age \u226575 years, and chi-square test for all other variables"
+    )
+  ) |> 
+  footnote(
+    i = ~ label == "Chronic Kidney Disease",
+    j = 1, 
+    ref_symbols = 'b',
+    value = ckd_defn,
+    part = 'body'
+  ) |> 
+  footnote(
+    i = 1,
+    j = 1,
+    ref_symbols = '',
+    value = abbrvs_write(abbrvs[c('SD', 'EHR', 'IQR', 'MoCA', 'FI')]),
+    part = 'body'
+  ) |>
+  fontsize(size = 12, part = 'all') |> 
+  font(fontname = 'Times New Roman', part = 'all')
+
+```
+
+\newpage
 

### 2022-01-07 R/bp_long_viz.R

@@ -1,0 +1,118 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param date
+bp_long_viz <- function(date = "12-27-21") {
+
+  cols_bg <- viridis(3)
+  cols_tx <- pal_lancet()(2)
+
+  .date <- str_remove_all(date, fixed('-'))
+
+  fname <- paste0("SPRINT_EHR_SBP_GLIMMIX_LongTerm_", .date, ".csv")
+
+  bp <- read_csv(file.path('ehr', fname)) |>
+    select(-Statement, -Alpha, -tValue, -DF, -StdErr) |>
+    separate(Label, into = c('group', 'time_months'), sep = ';') |>
+    mutate(time_months = as.numeric(time_months),
+           group = recode(group,
+                          "STD" = "Standard",
+                          "INT" = "Intensive",
+                          "DIFF" = "Difference"))
+
+  bp_levels <- filter(bp, group %in% c("Standard",
+                                       "Intensive"))
+  bp_diff <- filter(bp, group == 'Difference')
+
+  bp_targets <- tibble(
+    x = c(0, 0),
+    xend = c(2.54, 2.54),
+    y = c(140, 120),
+    yend = c(140, 120),
+    group = c("Standard", "Intensive")
+  )
+
+  bp_diff_targets <- tibble(
+    x = 0,
+    xend = 2.54,
+    y = 20,
+    yend = 20
+  )
+
+  fig_levels <- ggplot(bp_levels) +
+    aes(x = time_months/12,
+        y = Estimate,
+        color = group) +
+    geom_line(size = 1) +
+    labs(y = 'Systolic blood pressure, mm Hg',
+         x = 'Years since randomization',
+         fill = 'Treatment',
+         color = 'Treatment') +
+    scale_x_continuous(limits = c(0,10),
+                       breaks = seq(0,10)) +
+    scale_y_continuous(limits = c(120, 142),
+                       breaks = seq(120, 140, by=5)) +
+    scale_color_manual(values = cols_tx) +
+    scale_fill_manual(values = cols_tx) +
+    theme_fig(legend.position = 'top') +
+    geom_ribbon(alpha = 0.2,
+                mapping = aes(x = time_months/12,
+                              ymin = Lower,
+                              ymax = Upper,
+                              fill = group),
+                inherit.aes = FALSE) +
+    geom_segment(data=bp_targets,
[…]

### 2022-01-07 R/ndi_load.R

@@ -14,10 +14,12 @@ ndi_load <- function(sasinet_drive = "Z", fname, ...) {
     fname
   )
 
-  read_csv(path) |>
+
+  data_in <- read_csv(path) |>
     transmute(
       # ... makes it easy to modify the call to this function in targets
       # for including additional variables or modifying them.
+      pid,
       ...,
       age_yrs = age,
       age_cat = factor(sub_senior,
@@ -53,5 +55,21 @@ ndi_load <- function(sasinet_drive = "Z", fname, ...) {
                          levels = c(0,1),
                          labels = c('Standard', 'Intensive'))
     )
+
+
+  # indicate ancillary study membership for baseline data
+  if(fname == 'longterm_death.csv'){
+
+    ancillary <- read_csv('ehr/SPRINT_EHR_SBP_LongTerm_PID_010622.csv')
+
+    data_in$ehr_ancillary <- if_else(
+      condition = data_in$pid %in% ancillary$pid,
+      true = 'Yes',
+      false = 'No'
+    )
+
+  }
+
+  data_in
 
 }

### 2022-01-07 R/ndi_tabulate_characteristics.R

@@ -7,13 +7,10 @@
 ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
 
 
-  ndi_baseline$ehr_ancillary <- sample(x = c('No','Yes'),
-                                       size = nrow(ndi_baseline),
-                                       replace = TRUE)
-
   ndi_tbl_data <- ndi_baseline |>
     select(ehr_ancillary,
            age_yrs,
+           treatment,
            any_of(names(recoders$variable)))
 
   for(r in seq_along(recoders$variable)){
@@ -37,32 +34,73 @@ ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
 
   }
 
-  tbl_summary(data = ndi_tbl_data,
-              by = 'ehr_ancillary',
-              statistic = list(
-                age_yrs ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
-                all_categorical() ~ "{p}"
-              ),
-              label = list(age_cat ~ paste("Age", recoders$levels$age_cat[2]),
-                           sex ~ recoders$levels$sex[2],
-                           race ~ recoders$levels$race[2],
-                           moca ~ paste("MoCA", recoders$levels$moca[2])),
-              type = list(age_yrs ~ "continuous2",
-                          age_cat ~ "dichotomous",
-                          sex ~ "dichotomous",
-                          race ~ "dichotomous",
-                          ckd_2021 ~ "dichotomous",
-                          moca ~ "dichotomous",
-                          frail ~ "categorical"),
-              value = list(age_cat = recoders$levels$age_cat[2],
-                           sex = recoders$levels$sex[2],
-                           race = recoders$levels$race[2],
-                           ckd_2021 = recoders$levels$ckd_2021[2],
-                           moca = recoders$levels$moca[2]),
-              missing = 'no'
-              ) |>
-    add_overall() |>
+  tbl_by_anci <- ndi_tbl_data |>
+    select(-treatment) |>
+    tbl_summary(
+      by = 'ehr_ancillary',
+      statistic = list(
+        age_yrs ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
+        all_categorical() ~ "{p}"
+      ),
+      label = list(age_cat ~ paste("Age", recoders$levels$age_cat[2]),
+                   sex ~ recoders$levels$sex[2],
+                   race ~ recoders$levels$race[2],
[…]

### 2022-01-07 _targets.R

@@ -98,6 +98,8 @@ list(
     tar_target(base_viz_eff, ndi_base_viz_tv_effect(base_sub))
   ),
 
+  tar_target(bp_viz, bp_long_viz(date='12-27-21')),
+
   # ndi_long_describe(ndi_data_list)
 
   tar_combine(

### 2022-01-07 doc/manuscript.Rmd
bcj intro edits

@@
 
-both the The Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) trial have shown the benefits of intensive blood pressure control on cardiovascular morbidity and mortality [@sprint_research_group_randomized_2015; @zhang_trial_2021].
-However, both trials were stopped early after a median follow-up of 3 years.
+Blood pressure (BP) is the leading modifiable risk factor for cardiovascular disease (CVD).
+A meta analysis of pooled clinical trial data has shown that pharmacological BP reduction reduces the risk of major cardiovascular events, with larger absolute risk reduction for older adults [@rahimi_age_stratified_2021].
+The Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) each showed that intensive systolic BP (SBP) control, defined by a SBP < 120 mm Hg in SPRINT and SBP < 130 mm Hg in STEP, reduced the risk for cardiovascular morbidity and mortality [@sprint_research_group_randomized_2015; @zhang_trial_2021].
+However, as both trials were stopped early after a median follow-up of 3 years, the longer term effect of randomization to intensive SBP control on CVD risk and SBP levels following the trial period are unclear.
 
-__Reminder: bring in the recent BPTTC meta-analyses into the introduction__[@rahimi_age_stratified_2021].
-
-Here we examined the longer term effect of randomization to intensive treatment on through linkage to the National Death Index (NDI).
+The objective of the current study was to estimate the longer term effect of randomization to intensive SBP control on CVD risk for SPRINT participants.
+A secondary objective was to describe SBP levels over time following the trial period.
+To accomplish these objectives, we linked SPRINT participant data to the National Death Index (NDI) and pulled longitudinal SBP data for participants with available BP values in the electronic health record (EHR).
 
@@
 
-__Trial Design__: The trial design and methods have been published previously [@ambrosius_design_2014].
-Briefly, we conducted a multicenter randomized clinical trial that compared two strategies for managing systolic BP (SBP) in older adults with hypertension who were at increased risk of cardiovascular disease (CVD).
+__Trial Design__: The design and methods of SPRINT have been published previously [@ambrosius_design_2014].
+Briefly, we conducted a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk of CVD.
 Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed.

### 2022-01-07 doc/manuscript.Rmd
bcj edits to Results

@@
 
+NOTE: ADD Steve Rapp AS A CO-AUTHOR
+
 ^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC.
@@
                    long_viz_cvd,
+                   bp_viz,
+                   tbl_characteristics,
                    base_ds, 
@@
 
-__EHR Ancillary Study__: We examined the trajectory of systolic blood pressure (SBP) following the conclusion of the trial using outpatient SBPs extracted from the electronic health record (EHR).
+__EHR Ancillary Study__: We examined the trajectory of SBP following the conclusion of the trial using outpatient SBPs extracted from the EHR.
 Methods for the linkage of participants to their medical record number and the extraction of vital sign data have been previously described [@drawz_concordance_2020].
@@
 # RESULTS
+
+__ADD 1-2 SENTENCES ON CHARACTERISTICS__ (__Table 1__).
+Among participants included and not included in the ancillary EHR study, there was no evidence of a difference in study characteristics between participants randomized to intensive versus standard treatment (__eTable 1__).
 
@@
 
-The estimated mean SBP among participants randomized to intensive treatment was <placeholder (95% CI)> at 5 years and <placeholder (95% CI)> at 10 years post-randomization (__Figure 3__).
-For participants randomized to standard treatment, mean SBP was estimated to be <placeholder (95% CI)> at 5 years and <placeholder (95% CI)> at 10 years post-randomization.
+```{r}
+
+bp_viz_5yr <- bp_viz |> 
+  getElement('levels') |> 
+  getElement('data') |> 
+  filter(time_months == 60)
+
+bp_viz_5yr <- split(bp_viz_5yr, bp_viz_5yr$group)
+
+bp_viz_10yr <- bp_viz |> 
+  getElement('levels') |> 
+  getElement('data') |> 
+  filter(time_months == '120')
+
+bp_viz_10yr <- split(bp_viz_10yr, bp_viz_10yr$group)
+
+# a temporary function for inline reporting below.
+f <- function(d) with(d, .("{Estimate} ({Lower}, {Upper})"))
+
+bp_diff_5yr <- bp_viz |> 
+  getElement('diff') |> 
+  getElement('data') |> 
+  filter(time_months == 60)
+
+t_atten <- bp_viz |> 
+  getElement('diff') |> 
+  getElement('data') |> 
+  filter(find_signchange_lgl(Lower)) |> 
+  pull(time_months) |> 
+  divide_by(12)
+
+```
+
+The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization (__Figure 3__; panel A).
+For participants randomized to standard treatment, mean (95% CI) SBP was estimated to be `r f(bp_viz_5yr$Standard)` at 5 years and `r f(bp_viz_10yr$Standard)` post-randomization.
+The difference in mean SBP levels between participants randomized to intensive versus standard treatment, in mm Hg, was `r f(bp_diff_5yr)` at 5 years post-randomization, and was attenuated by `r t_atten` years post-randomization (__Figure 3__; Panel B).
 
@@
 ```{r}
-
-tar_load(tbl_characteristics)
 
@@
 
-tar_load(bp_viz)
-
 blank_plot <- ggplot(data.frame()) + aes() + theme_void()

### 2022-01-17 doc/manuscript.Rmd
nickwork2

@@
 Jaeger, PhD,^1^ Adam P.
-Bress, PharmD, MS,^2,3^ Alfred K.
-Cheung, MD,^4,5^ William C.
-Cushman, MD,^6^ Paul E.
-Drawz, MD,^7^ Lawrence J.
-Fine, MD,^8^ Karen C.
-Johnson, MD,^6^ Cora E.
-Lewis, MD,^9^ Suzanne Oparil, MD,^10^ Michael V.
-Rocco, MD,^11^ Mark A.
-Supiano, MD,^12^ Paul K.
-Whelton, MD,^13^ Joshua D.
-Bundy, PhD,^13^ Jeff D.
-Williamson, MD, MHS,^14^ Jackson T.
-Wright, Jr, MD, PhD,^15^ David M.
+Bress, PharmD, MS,^2,3^ Joshua D.
+Bundy, PhD,^4^ Alfred K.
+Cheung, MD,^5,6^ William C.
+Cushman, MD,^7^ Paul E.
+Drawz, MD,^8^ Lawrence J.
+Fine, MD,^9^ Karen C.
+Johnson, MD,^7^ Cora E.
+Lewis, MD,^10^ Suzanne Oparil, MD,^11^ Michael V.
+Rocco, MD,^12^ Stephen R.
+Rapp, PhD, ^13^ Joni K.
+Snyder, RN,BSN,^9^ Mark A.
+Supiano, MD,^14^ Paul K.
+Whelton, MD,^4^ Jeff D.
+Williamson, MD, MHS,^15^ Jackson T.
+Wright, Jr, MD, PhD,^16^ David M.
 Reboussin, PhD,^1^ and Nicholas M.
 Pajewski, PhD^1^
-
-NOTE: ADD Steve Rapp AS A CO-AUTHOR
 
@@
 ^3^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT.
-^4^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT.
-^5^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT.
-^6^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis.
-^7^Division of Renal Diseases & Hypertension, University of Minnesota, MN.
-^8^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD.
-^9^Department of Epidemiology, University of Alabama at Birmingham.
-^10^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham.
-^11^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^12^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
-^13^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
-^14^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^15^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
+^4^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
+^5^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT.
+^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT.
+^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis.
+^8^Division of Renal Diseases & Hypertension, University of Minnesota, MN.
+^9^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD.
+^10^Department of Epidemiology, University of Alabama at Birmingham.
+^11^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham.
+^12^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+^13^Departments of Psychiatry and Behavioral Medicine & Social Sciences and Health Policy, Wakfe Forest School of Medicine, Winston-Salem, NC.
+^14^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
+^15^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+^16^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
 \
@@
 Cardiovascular and all-cause mortality assessed via the US National Death Index, beginning in 2016 through December 31st, 2020.
-Outpatient blood pressures measured in routine clinical practice were examined in a subset of trial participants (N=3644).
+Outpatient blood pressures measured in routine clinical practice after the trial were examined in a subset of trial participants (N=3644).
 
@@
 Among 9361 randomized participants (mean age, 67.9 years; 3332 women [35.6%]), the median intervention period was 3.34 years.
-Over a median follow-up of `r ap_b_o$fup_est` years, intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial (follow-up through July 2016).
+Over a median follow-up of `r ap_b_o$fup_est` years, intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial.
 However, there was no indication of benefit during post-trial follow-up for either cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
 Results were similar for subgroups based on baseline age, cognitive function, and frailty status.
-Analyses of outpatient blood pressures indicated a steady decline in the mean between group difference following the trial, largely driven by increases in mean systolic blood pressure in participnts randomized to intensive treatment, increasing from a mean of X at 4.5 years of follow-up to Y at 9 years of follow-up.
-
-there were `r ap_b_i$n_cvd` and `r ap_b_i$n_acm` cardiovascular and all-cause deaths with intensive treatment respectively, and `r ap_b_s$n_cvd` cardiovascular / `r ap_b_s$n_acm` all-cause deaths for standard treatment.
+Analyses of outpatient blood pressures indicated a steady decline in the mean between group difference following the trial, largely driven by increases in mean systolic blood pressure in participants randomized to intensive treatment, increasing from a mean of 131.7 mm Hg (95% CI 130.7 to 132.6 mm Hg) at 4.5 years of follow-up to 136.8 mm Hg (95% CI 135.7 to 137.8 mm Hg) at 9 years of follow-up.
 
@@
 
-Blood pressure (BP) is the leading modifiable risk factor for cardiovascular disease (CVD).
-A meta analysis of pooled clinical trial data has shown that pharmacological BP reduction reduces the risk of major cardiovascular events, with larger absolute risk reduction for older adults [@rahimi_age_stratified_2021].
-The Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) each showed that intensive systolic BP (SBP) control, defined by a SBP < 120 mm Hg in SPRINT and SBP < 130 mm Hg in STEP, reduced the risk for cardiovascular morbidity and mortality [@sprint_research_group_randomized_2015; @zhang_trial_2021].
-However, as both trials were stopped early after a median follow-up of 3 years, the longer term effect of randomization to intensive SBP control on CVD risk and SBP levels following the trial period are unclear.
+Hypertension is the leading modifiable risk factor for cardiovascular disease (CVD).[@gdbJACC2020] A meta analysis of pooled individual-level data from randomized trials has shown that pharmacological blood pressure (BP) reduction reduces the risk of major cardiovascular events across the spectrum of initial BP, with larger absolute risk reduction for older adults.[@rahimi_age_stratified_2021] The Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) each showed that intensive systolic BP (SBP) control, defined by a SBP target < 120 mm Hg in SPRINT and < 130 mm Hg in STEP, reduced the risk for cardiovascular morbidity and mortality.
+[@sprint_research_group_randomized_2015; @zhang_trial_2021] However,
+only SPRINT showed a benefit of intensive treatment on all-cause mortality.
+Another contrasting result was that intensive treatment in SPRINT, but not STEP, increased the risk of acute kidney injury (AKI).
+While this may be attributable to the much lower prevalence of chronic kidney disease in STEP (2.3% vs 28.3%), the consistent association of AKI with increased mortality risk[@chawla2014aki; @rocco2018aki] raises questions about the longer term effect of intensive treatment on CVD and all-cause mortality.
 
-The objective of the current study was to estimate the longer term effect of randomization to intensive SBP control on CVD risk for SPRINT participants.
-A secondary objective was to describe SBP levels over time following the trial period.
-To accomplish these objectives, we linked SPRINT participant data to the National Death Index (NDI) and pulled longitudinal SBP data for participants with available BP values in the electronic health record (EHR).
+Because both trials were stopped early after a median follow-up of 3 years, the objective of the current study was to estimate the longer term, legacy effect of randomization to intensive treatment on cardiovascular and all-cause mortality risk for participants in SPRINT.
+A secondary objective, in a subset of trial participants, was to examine attained SBP following the conclusion of trial follow-up.
+To accomplish these objectives, we linked SPRINT participants to the National Death Index (NDI) from 2016 to 2020, and extracted longitudinal outpatient measurements of SBP available in the electronic health record (EHR) over the same time period.
 
@@
 
-__Trial Design__: The design and methods of SPRINT have been published previously [@ambrosius_design_2014].
-Briefly, we conducted a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk of CVD.
+__Trial Design__: The design and methods of SPRINT have been published previously.[@ambrosius_design_2014; @sprint_research_group_randomized_2015] Briefly, it was a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk of CVD.
 Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed.
@@
 ).
-Trial enrollment began in November 2010 and ended in March 2013, with active follow-up through July 1, 2016.
-The study was approved by the institutional review board at each participating site, and each participant provided written informed consent.
+Trial enrollment began in November 2010 and ended in March 2013, with active follow-up through July 1, 2016.[@sprintFinal] The study was approved by the institutional review board at each participating site, and each participant provided written informed consent.
 The study is registered at ClinicalTrials.gov (NCT01206062).
@@
 __Baseline Study Measurements__: Sociodemographic data were collected at baseline, with race or ethnicity information collected via self-report.
-The estimated glomerular filtration rate (eGFR) was calculated by the race-free 2021 CKD-EPI creatinine equation.[@inker_new_2021] Cognitive function was assessed using Montreal Cognitive Assessment (MoCA) [@nasreddine_montreal_2005].
-Lower cognitive function was defined as scoring 18 or lower (less than high school education) or 20 or lower (high school education or higher) on the MoCA.
-This roughly corresponds to the estimated normative 10th percentile in the Irish Longitudinal Study of Aging [@kenny_normative_2013].
-We defined frailty status at baseline using a previously developed Frailty Index (FI) based upon the model of deficit accumulation [@pajewski_characterizing_2016].
-Briefly, the FI comprises a total of 36 items, and is calculated as the sum of the score for each deficit divided by the total number of nonmissing items.
+The estimated glomerular filtration rate (eGFR) was calculated by the race-free 2021 CKD-EPI creatinine equation.[@inker_new_2021] Cognitive function was assessed using the Montreal Cognitive Assessment (MoCA) [@nasreddine_montreal_2005].
+Lower cognitive function was defined as scoring at or below the estimated age and education-specific normative 10th percentile from the Irish Longitudinal Study of Aging,[@kenny_normative_2013] after adding 3 points to the scores of non-White participants.[@sachsMoCA2021] We defined frailty status at baseline using a 36-item Frailty Index (FI) based upon the model of deficit accumulation.[@pajewski_characterizing_2016] The FI is calculated as the sum of the score for each deficit divided by the total number of nonmissing items.
 We categorized frailty status as fit (FI ≤ 0.10), less fit (0.10 < FI ≤ 0.21), or frail (FI > 0.21).
@@
 __National Death Index Linkage__: Outcomes of interest included all-cause and CVD mortality.
-Methods of ascertainment and adjudication through the course of trial follow-up have been previously described.
-Subsequently, mortality was ascertained through a US National Death Index (NDI) search.
-Possible matches were identified according to NDI guidelines.
-To be considered a confirmed death, we required 4 or more of 5 matches among Social Security number, name, date of birth, city, and state in the NDI.
-NDI follow-up began in 2016 and ended on the date of death or date of the NDI search (2020).
-Deaths ascertained in 2020 were based on the preliminary data release.
+Methods of ascertainment and adjudication through the course of trial follow-up have been previously described.[@sprint_research_group_randomized_2015; @sprintFinal] Subsequently, mortality was ascertained through a US National Death Index (NDI) search.
+Possible matches were identified according to NDI guidelines.[@ndi_guide] Deaths were treated as confirmed if they were a Class 1 match, or a Class 2, 3, or 4 match with a probabilistic score above cutoffs recommended by the NDI.[@ndi_guide] NDI follow-up began in 2016 and ended on the date of death or December 31, 2020.
+Deaths ascertained in 2020 were based on the NDI preliminary data release.
 CVD mortality for NDI-based follow-up used the NDI Plus System, which automatically identifies underlying causes of death from death certificates, including conversion to ICD-10 codes.
-we defined CVD mortality as any death containing the ICD-10 codes of I00 to I99.
+we defined CVD mortality as any death containing the ICD-10 codes of I00 to I99.[@regards_ndi]
 
 __EHR Ancillary Study__: We examined the trajectory of SBP following the conclusion of the trial using outpatient SBPs extracted from the EHR.
-Methods for the linkage of participants to their medical record number and the extraction of vital sign data have been previously described [@drawz_concordance_2020].
-Because encounter type information was inconsistently available (i.e.
+Methods for the linkage of participants to their medical record number and the extraction of vital sign data have been previously described.[@drawz_concordance_2020] Because encounter type information was inconsistently available (i.e.
 outpatient, inpatient, observation, etc.), we defined a BP measurement as outpatient if there were was not a BP measurement on the preceding or following day, and if there were 2 or less BP measurements on a particular day.
@@
 
-___Statistical Analysis___: The effect of randomization to intensive treatment was estimated as a function of time using two approaches.
-The first approach split the follow-up of study participants into non-overlapping trial and cohort phases, then estimated the effect of randomization to intensive treatment in each phase [@zhang_time_varying_2018].
-The second approach estimated a continuous time-dependent effect [@martinussen2007dynamic; @therneau2017using].
-All analyses accounted for correlation within study sites [@lin_robust_1989], and analyses of cardiovascular mortality accounted for the competing risk of non-cardiovascular mortality [@scheike_flexible_2008].
-The mean between-group differences in SBP following the conclusion of trial follow-up (after July 2016) were estimated using linear mixed models.
-Models included random effects for participant and clinic site.
-Our primary models included an interaction between treatment group and time since randomization, which was flexibly modeled using B-splines.
-
-All analyses were performed using SAS version 9.4 (SAS Institute Inc, Cary, NC) and R version `r paste(version$major, version$minor, sep = '.')` (R Project for Statistical Computing [http://www.r-project.org]) with multiple auxiliary R packages [@r_language; @table.glue; @tidyverse; @survival_package; @timereg; @targets].
-All hypothesis tests were 2-sided, and P values less than 0.05 were considered statistically significant.
+__Statistical Analysis__: Given the a priori expectation that treatment group differences may not be constant as a function of follow-up time (i.e.
+the proportional hazards assumption was likely to be invalid), we modeled treatment group differences as a function of time using two approaches.
+The first approach used Cox proportional hazards regression but split follow-up time into non-overlapping trial and cohort phases, estimating separate treatment group differences for each phase.[@zhang_time_varying_2018].
+The second approach estimated the effect of treatment group as a continuous time-dependent effect.[@martinussen2007dynamic; @therneau2017using] All analyses accounted for correlation within study sites,[@lin_robust_1989] and analyses of cardiovascular mortality accounted for the competing risk of non-cardiovascular mortality.[@scheike_flexible_2008].
+Mean between-group differences in outpatient SBP following the conclusion of trial follow-up were estimated using linear mixed models.
+Models included random effects for participant and clinic site and an interaction between treatment group and time since randomization, which was flexibly modeled using B-splines.
+All analyses were performed using SAS version 9.4 (SAS Institute Inc, Cary, NC) and R version `r paste(version$major, version$minor, sep = '.')` (R Project for Statistical Computing [http://www.r-project.org]) with multiple auxiliary R packages.[@r_language; @table.glue; @tidyverse; @survival_package; @timereg; @targets] All hypothesis tests were 2-sided, and P values less than 0.05 were considered statistically significant.
 No adjustments for multiple comparisons were made.
@@
 # RESULTS
-
-__ADD 1-2 SENTENCES ON CHARACTERISTICS__ (__Table 1__).
+___Study Participants___.
+A total of 9361 participants were randomized between November 2010 and March 2013.
+The mean age was 67.9 years (SD, 9.4 years), with 28.2% of participants aged 75 years or older.
+Randomized participants were 35.6% female, 30.0% black, and 10.5% Hispanic.
+The mean SBP at baseline was 139.7 mm Hg (SD, 15.6 mm Hg), and the median MoCA score was 23 (interquartile range, 20-26) (__Table 1__).
 Among participants included and not included in the ancillary EHR study, there was no evidence of a difference in study characteristics between participants randomized to intensive versus standard treatment (__eTable 1__).
@@
 
+___All-cause Mortality___.
 In both treatment groups, median follow-up time was `r ap_b_o$fup_est` years.
-A total of `r ap_b_i$n_acm` and `r ap_b_s$n_acm` all-cause mortality events occurred among participants randomized to intensive and standard treatment, respectively (__Table 2__).
-The hazard ratio (HR) for all-cause mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during the cohort phase.
+A total of `r ap_b_i$n_acm` and `r ap_b_s$n_acm` deaths occurred among participants randomized to intensive and standard treatment, respectively (__Table 2__).
+The hazard ratio (HR) for all-cause mortality comparing intensive to standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial, and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during observational follow-up following the trial.
 In subgroups based on age, sex, race, CKD, cognitive function, and frailty, there was no evidence that intensive treatment during the trial phase had benefit for all-cause mortality during the cohort phase (__eFigure 1__).
@@
 
+___Cardiovascular Mortality___.
 A total of `r ap_b_i$n_cvd` and `r ap_b_s$n_cvd` CVD mortality events occurred among participants randomized to intensive and standard treatment, respectively (__Table 3__).
@@
 
+___Blood Pressure___.
 The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization (__Figure 3__; panel A).
@@
 
-Online supplement only
+__Online supplement only__
 
@@
 Jaeger, PhD,^1^ Adam P.
-Bress, PharmD, MS,^2,3^ Alfred K.
-Cheung, MD,^4,5^ William C.
-Cushman, MD,^6^ Paul E.
-Drawz, MD,^7^ Lawrence J.
-Fine, MD,^8^ Karen C.
-Johnson, MD,^6^ Cora E.
-Lewis, MD,^9^ Suzanne Oparil, MD,^10^ Michael V.
-Rocco, MD,^11^ Mark A.
-Supiano, MD,^12^ Paul K.
-Whelton, MD,^13^ Joshua D.
-Bundy, PhD,^13^ Jeff D.
-Williamson, MD, MHS,^14^ Jackson T.
-Wright, Jr, MD, PhD,^15^ David M.
+Bress, PharmD, MS,^2,3^ Joshua D.
+Bundy, PhD,^4^ Alfred K.
+Cheung, MD,^5,6^ William C.
+Cushman, MD,^7^ Paul E.
+Drawz, MD,^8^ Lawrence J.
+Fine, MD,^9^ Karen C.
+Johnson, MD,^7^ Cora E.
+Lewis, MD,^10^ Suzanne Oparil, MD,^11^ Michael V.
+Rocco, MD,^12^ Stephen R.
+Rapp, PhD, ^13^ Joni K.
+Snyder, RN,BSN,^9^ Mark A.
+Supiano, MD,^14^ Paul K.
+Whelton, MD,^4^ Jeff D.
+Williamson, MD, MHS,^15^ Jackson T.
+Wright, Jr, MD, PhD,^16^ David M.
 Reboussin, PhD,^1^ and Nicholas M.
@@
 ^3^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT.
-^4^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT.
-^5^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT.
-^6^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis.
-^7^Division of Renal Diseases & Hypertension, University of Minnesota, MN.
-^8^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD.
-^9^Department of Epidemiology, University of Alabama at Birmingham.
-^10^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham.
-^11^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^12^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
-^13^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
-^14^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^15^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
+^4^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
+^5^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT.
+^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT.
+^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis.
+^8^Division of Renal Diseases & Hypertension, University of Minnesota, MN.
+^9^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD.
+^10^Department of Epidemiology, University of Alabama at Birmingham.
+^11^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham.
+^12^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+^13^Departments of Psychiatry and Behavioral Medicine & Social Sciences and Health Policy, Wakfe Forest School of Medicine, Winston-Salem, NC.
+^14^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
+^15^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+^16^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
 \
-
-**Correspondence to:** \
-Nicholas M. Pajewski, PhD \
-Department of Biostatistics and Data Science \
-Division of Public Health Sciences \
-Wake Forest School of Medicine \
-Medical Center Boulevard \
-Winston-Salem, NC 27154 \
-(336) 713-1396 \
-npajewsk@wakehealth.edu `r single_par`
 

### 2022-01-17 R/ndi_load.R

@@ -7,7 +7,7 @@
 ndi_load <- function(sasinet_drive = "Z", fname, ...) {
 
   path <- file.path(
-    paste0(sasinet_drive, ":"),
+    paste0(sasinet_drive, ""),
     "npajewski",
     "NDI",
     "Data",

### 2022-01-17 _targets.R

@@ -4,8 +4,8 @@
 source("./packages.R")
 
 # used by NMP to run target pipeline
-# Sys.setenv(RSTUDIO_PANDOC="C:/Program Files/RStudio/bin/pandoc")
-# setwd("O:/sprint/npajewski/sprint-mortality-ndi")
+Sys.setenv(RSTUDIO_PANDOC="C:/Program Files/RStudio/bin/pandoc")
+setwd("O:/sprint/npajewski/sprint-mortality-ndi")
 
 ## Load your R files
 lapply(list.files("./R", full.names = TRUE), source)
@@ -24,7 +24,7 @@ list(
 
   tar_target(
     ndi_baseline,
-    ndi_load(sasinet_drive = get_sasinet_drive(),
+    ndi_load(sasinet_drive = get_sasinet_drive("nmpieyeskey"),
              fname = 'longterm_death.csv',
              # participant identifier
              pid,
@@ -39,7 +39,7 @@ list(
 
   tar_target(
     ndi_longitudinal,
-    ndi_load(sasinet_drive = get_sasinet_drive(),
+    ndi_load(sasinet_drive = get_sasinet_drive("nmpieyeskey"),
              fname = 'longterm_death_td_subgroup.csv',
              pid,
              randSite,
@@ -98,7 +98,7 @@ list(
     tar_target(base_viz_eff, ndi_base_viz_tv_effect(base_sub))
   ),
 
-  tar_target(bp_viz, bp_long_viz(date='12-27-21')),
+  tar_target(bp_viz, bp_long_viz(date='01-15-22')),
 
   # ndi_long_describe(ndi_data_list)
 

### 2022-01-18 R/ndi_tabulate_characteristics.R
#8

@@ -35,7 +35,7 @@ ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
   }
 
   tbl_by_anci <- ndi_tbl_data |>
-    select(-treatment) |>
+    relocate(treatment, .after = race) |>
     tbl_summary(
       by = 'ehr_ancillary',
       statistic = list(
@@ -44,16 +44,19 @@ ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
       ),
       label = list(age_cat ~ paste("Age", recoders$levels$age_cat[2]),
                    sex ~ recoders$levels$sex[2],
+                   treatment ~ "Intensive treatment",
                    race ~ recoders$levels$race[2],
                    moca ~ paste("MoCA", recoders$levels$moca[2])),
       type = list(age_yrs ~ "continuous2",
                   age_cat ~ "dichotomous",
+                  treatment ~ "dichotomous",
                   sex ~ "dichotomous",
                   race ~ "dichotomous",
                   ckd_2021 ~ "dichotomous",
                   moca ~ "dichotomous",
                   frail ~ "categorical"),
       value = list(age_cat = recoders$levels$age_cat[2],
+                   treatment = 'Intensive',
                    sex = recoders$levels$sex[2],
                    race = recoders$levels$race[2],
                    ckd_2021 = recoders$levels$ckd_2021[2],
@@ -62,44 +65,75 @@ ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
     ) |>
     add_p()
 
-  tbls_by_tx <- split(ndi_tbl_data, ndi_tbl_data$ehr_ancillary) |>
-    map(
-      ~ .x |>
-        select(-ehr_ancillary) |>
-        tbl_summary(
-          by = 'treatment',
-          statistic = list(
-            age_yrs ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
-            all_categorical() ~ "{p}"
-          ),
-          label = list(age_cat ~ paste("Age", recoders$levels$age_cat[2]),
-                       sex ~ recoders$levels$sex[2],
-                       race ~ recoders$levels$race[2],
-                       moca ~ paste("MoCA", recoders$levels$moca[2])),
-          type = list(age_yrs ~ "continuous2",
-                      age_cat ~ "dichotomous",
-                      sex ~ "dichotomous",
-                      race ~ "dichotomous",
-                      ckd_2021 ~ "dichotomous",
[…]

### 2022-01-18 _targets.R
specify user to update workflow and filepaths

@@ -3,9 +3,15 @@
 ## Load your packages, e.g. library(targets).
 source("./packages.R")
 
+# determine the paths used for accessing sasinet
+user <- 'bcjaeger'
+
 # used by NMP to run target pipeline
-Sys.setenv(RSTUDIO_PANDOC="C:/Program Files/RStudio/bin/pandoc")
-setwd("O:/sprint/npajewski/sprint-mortality-ndi")
+if(user == 'nmpieyeskey'){
+  Sys.setenv(RSTUDIO_PANDOC="C:/Program Files/RStudio/bin/pandoc")
+  setwd("O:/sprint/npajewski/sprint-mortality-ndi")
+}
+
 
 ## Load your R files
 lapply(list.files("./R", full.names = TRUE), source)
@@ -18,13 +24,14 @@ subgroups <- tibble(group = c("overall",
                               "moca",
                               "frail"))
 
+
 list(
 
   tar_target(recoders, recoders_make()),
 
   tar_target(
     ndi_baseline,
-    ndi_load(sasinet_drive = get_sasinet_drive("nmpieyeskey"),
+    ndi_load(sasinet_drive = get_sasinet_drive(user),
              fname = 'longterm_death.csv',
              # participant identifier
              pid,
@@ -39,7 +46,7 @@ list(
 
   tar_target(
     ndi_longitudinal,
-    ndi_load(sasinet_drive = get_sasinet_drive("nmpieyeskey"),
+    ndi_load(sasinet_drive = get_sasinet_drive(user),
              fname = 'longterm_death_td_subgroup.csv',
              pid,
              randSite,

### 2022-01-18 doc/manuscript.Rmd
#9 fixed plus inline text

@@
 Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed.
-Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (defined by an estimated glomerular filtration rate of <60 mL/min/1.73 m^2^), or a Framingham Risk Score of 15% or greater or if they were aged 75 years or older.
+Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (CKD; defined by an estimated glomerular filtration rate of <60 mL/min/1.73 m^2^), or a Framingham Risk Score of 15% or greater or if they were aged 75 years or older.
 Individuals residing in a nursing home, persons with a diagnosis of dementia (based on medical record review), and those treated with medications primarily used for dementia therapy were excluded, as were persons with prevalent diabetes mellitus, history of stroke, proteinuria > 1 gram per day, or polycystic kidney disease.
@@
 # RESULTS
+
 ___Study Participants___.
-A total of 9361 participants were randomized between November 2010 and March 2013.
-The mean age was 67.9 years (SD, 9.4 years), with 28.2% of participants aged 75 years or older.
-Randomized participants were 35.6% female, 30.0% black, and 10.5% Hispanic.
-The mean SBP at baseline was 139.7 mm Hg (SD, 15.6 mm Hg), and the median MoCA score was 23 (interquartile range, 20-26) (__Table 1__).
-Among participants included and not included in the ancillary EHR study, there was no evidence of a difference in study characteristics between participants randomized to intensive versus standard treatment (__eTable 1__).
+A total of `r nrow(ndi_baseline)` participants were randomized between November 2010 and March 2013.
+The mean (standard deviation [SD]) age was `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years, with `r inline_text(tbl_characteristics, variable = 'age_cat', column = 'stat_0_1')`% of participants aged 75 years or older (__eTable 1__).
+Participants were `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% female and `r inline_text(tbl_characteristics, variable = 'race', column = 'stat_0_1')`% black.
+The mean (SD) SBP at baseline was `r inline_text(tbl_characteristics, variable = 'sbp', column = 'stat_0_1')` and `r inline_text(tbl_characteristics, variable = 'moca', column = 'stat_0_1')`% of participants MoCA scores were below the 10^th^ percentile.
+Participants included in the ancillary EHR study had greater age, lower SBP, less cognitive impairment, higher prevalence of CKD, and were more likely to be male or black than participants not included.
 
 ```{r}
+
+# `r inline_text(tbl_characteristics, variable = 'moca', column = 'stat_0_1')`
 
@@
 In both treatment groups, median follow-up time was `r ap_b_o$fup_est` years.
-A total of `r ap_b_i$n_acm` and `r ap_b_s$n_acm` deaths occurred among participants randomized to intensive and standard treatment, respectively (__Table 2__).
+A total of `r ap_b_i$n_acm` and `r ap_b_s$n_acm` deaths occurred among participants randomized to intensive and standard treatment, respectively (__eTable 2__).
 The hazard ratio (HR) for all-cause mortality comparing intensive to standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial, and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during observational follow-up following the trial.
-In subgroups based on age, sex, race, CKD, cognitive function, and frailty, there was no evidence that intensive treatment during the trial phase had benefit for all-cause mortality during the cohort phase (__eFigure 1__).
 The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 1__).
+In subgroups based on age, sex, race, CKD, cognitive function, and frailty, there was no evidence that intensive treatment during the trial phase had benefit for all-cause mortality during the cohort phase (__Figure 2__).
 
@@
 ___Cardiovascular Mortality___.
-A total of `r ap_b_i$n_cvd` and `r ap_b_s$n_cvd` CVD mortality events occurred among participants randomized to intensive and standard treatment, respectively (__Table 3__).
+A total of `r ap_b_i$n_cvd` and `r ap_b_s$n_cvd` CVD mortality events occurred among participants randomized to intensive and standard treatment, respectively (__eTable 3__).
 The HR for CVD mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the cohort phase.
-Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase had benefit for CVD mortality during the cohort phase (__eFigure 2__).
-The time-dependent effect of intensive versus standard treatment indicated a benefit for CVD mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort period (__Figure 2__).
+The time-dependent effect of intensive versus standard treatment indicated a benefit for CVD mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort period (__Figure 3__).
+Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase had benefit for CVD mortality during the cohort phase (__Figure 4__).
 
@@
 ___Blood Pressure___.
-The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization (__Figure 3__; panel A).
+The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization (__Figure 5__; panel A).
 For participants randomized to standard treatment, mean (95% CI) SBP was estimated to be `r f(bp_viz_5yr$Standard)` at 5 years and `r f(bp_viz_10yr$Standard)` post-randomization.
-The difference in mean SBP levels between participants randomized to intensive versus standard treatment, in mm Hg, was `r f(bp_diff_5yr)` at 5 years post-randomization, and was attenuated by `r t_atten` years post-randomization (__Figure 3__; Panel B).
+The difference in mean SBP levels between participants randomized to intensive versus standard treatment, in mm Hg, was `r f(bp_diff_5yr)` at 5 years post-randomization, and was attenuated by `r t_atten` years post-randomization (__Figure 5__; Panel B).
 
@@
 
+<!-- new page not needed at the end of a block section -->
+```{r}
+
+margins_normal <- page_mar(bottom = 2.54 * 0.3937, 
+                           top    = 2.54 * 0.3937, 
+                           right  = 2.54 * 0.3937, 
+                           left   = 2.54 * 0.3937)
+
+block_section(
+  prop_section(
+    page_size = page_size(orient = "portrait"),
+    type = "continuous",
+    page_margins = margins_normal
+  )
+)
+
+```
+
+__Figure 1__: (A) Cumulative incidence of all-cause mortality by treatment group. (B) Time-dependent effect of randomization to intensive treatment for all-cause mortality. `r single_par`
+
+```{r}
+
+blank_plot <- ggplot(data.frame()) + aes() + theme_void()
+
+legend <- get_legend(
+  fig$inc_acm[[1]]$plot + 
+    theme(legend.position = c(.55, .2),
+          legend.direction = 'horizontal')
+)
+
+plot_grid(
+  # row 1 (legend)
+  legend, 
+  blank_plot,
+  # row 2 (figures)
+  fig$inc_acm[[1]]$plot + theme(legend.position = ''), 
+  fig$eff_acm[[1]],
+  # row 3 (table)
+  fig$inc_acm[[1]]$table, 
+  blank_plot,
+  ncol = 2,
+  align = 'v',
+  axis = 'lr',
+  rel_heights = c(1,10,2),
+  labels = c("","","A","B","",""),
+  label_x = 0.1,
+  label_y = 1.05
+)
+
+```
+
 \newpage
-__Table 1__: Characteristics of study participants stratified by inclusion in the electronic health record ancillary study. `r single_par`
+__Figure 2__: All-cause mortality hazard ratio for participants randomized to intensive versus standard treatment.
+
+```{r}
+
+long_viz_mort
+
+```
+
+\newpage
+__Figure 3__: (A) Cumulative incidence of cardiovascular and non-cardiovascular mortality by treatment group.
+(B) Time-dependent effect of randomization to intensive treatment for cardiovascular mortality.
+`r single_par`
+
+```{r}
+
+plot_grid(
+  # row 1 (legend)
+  legend, 
+  blank_plot,
+  # row 2 (figures)
+  fig$inc_cvd[[1]] + theme(legend.position = ''), 
+  fig$eff_cvd[[1]],
+  ncol = 2,
+  align = 'v',
+  axis = 'lr',
+  rel_heights = c(1,10,2),
+  labels = c("","","A","B","",""),
+  label_x = 0.1,
+  label_y = 1.05
+)
+
+```
+
+\newpage
+__Figure 4__: Cardiovascular mortality hazard ratio for participants randomized to intensive versus standard treatment.
+
+```{r}
+
+long_viz_cvd
+
+```
+
+\newpage
+__Figure 5__: (A) Mean systolic blood pressure over time by treatment group.
+(B) Mean difference in systolic blood pressure over time.
+Shaded areas indicate a 95% confidence interval for the mean.
+Horizontal lines during the trial period show target blood pressure values for each treatment group `r single_par`
+
+```{r}
+
+blank_plot <- ggplot(data.frame()) + aes() + theme_void()
+
+legend <- get_legend(
+  bp_viz$levels + 
+    theme(legend.position = c(.55, .2),
+          legend.direction = 'horizontal')
+)
+
+plot_grid(
+  # row 1 (legend)
+  legend, 
+  blank_plot,
+  # row 2 (figures)
+  bp_viz$levels + theme(legend.position = ''), 
+  bp_viz$diff,
+  ncol = 2,
+  align = 'v',
+  axis = 'lr',
+  rel_heights = c(1,10,2),
+  labels = c("","","A","B","",""),
+  label_x = 0.1,
+  label_y = 1.05
+)
+
+# fpath <- file.path(
+#   paste0(get_sasinet_drive(), ":"),
+#   'bjaeger',
+#   'sprint-mortality-ndi',
+#   'fig',
+#   'SPRINT_EHR_SBP_LongTerm_Comparison_122721.png'
+# )
+# 
+# knitr::include_graphics(
+#   path = fpath, 
+#   dpi = 600
+# )
+
+```
+
+```{r}
+block_section(
+  prop_section(
+    page_size = page_size(orient = "landscape"),
+    page_margins = page_mar(bottom = 1/8, top = 1/8, 
+                            right = 1/4, left = 1/4),
+    type = "continuous"
+  )
+)
+```
+
+__Online supplement only__
+
+Byron C.
+Jaeger, PhD,^1^ Adam P.
+Bress, PharmD, MS,^2,3^ Joshua D.
+Bundy, PhD,^4^ Alfred K.
+Cheung, MD,^5,6^ William C.
+Cushman, MD,^7^ Paul E.
+Drawz, MD,^8^ Lawrence J.
+Fine, MD,^9^ Karen C.
+Johnson, MD,^7^ Cora E.
+Lewis, MD,^10^ Suzanne Oparil, MD,^11^ Michael V.
+Rocco, MD,^12^ Stephen R.
+Rapp, PhD, ^13^ Joni K.
+Snyder, RN,BSN,^9^ Mark A.
+Supiano, MD,^14^ Paul K.
+Whelton, MD,^4^ Jeff D.
+Williamson, MD, MHS,^15^ Jackson T.
+Wright, Jr, MD, PhD,^16^ David M.
+Reboussin, PhD,^1^ and Nicholas M.
+Pajewski, PhD^1^
+
+^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC.
+^2^Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT.
+^3^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT.
+^4^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
+^5^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT.
+^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT.
+^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis.
+^8^Division of Renal Diseases & Hypertension, University of Minnesota, MN.
+^9^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD.
+^10^Department of Epidemiology, University of Alabama at Birmingham.
+^11^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham.
+^12^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+^13^Departments of Psychiatry and Behavioral Medicine & Social Sciences and Health Policy, Wakfe Forest School of Medicine, Winston-Salem, NC.
+^14^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
+^15^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
+^16^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
+\
+
+```{r}
+block_section(
+  prop_section(
+    page_size = page_size(orient = "portrait"),
+    type = "continuous",
+    page_margins = margins_normal
+  )
+)
+```
+
+__eTable 1__: Characteristics of study participants.`r single_par`
 
@@
 header_labels <- tbl_characteristics |> 
-  getElement('main') |> 
   show_header_names(quiet = TRUE) |> 
@@
     label = str_remove_all(label, fixed("*")),
+    label = if_else(column == 'stat_0_1', 
+                    paste0("Overall, ", label),
+                    label),
     label = str_replace(label, ', ', '\n')
@@
 tbl_characteristics |> 
-  getElement('main') |> 
   modify_footnote(update = everything() ~ NA) |> 
+  modify_spanning_header(update = everything() ~ NA) |> 
   as_flex_table() |> 
+  set_header_labels(
+    label = "Characteristic",
+    stat_0_1 = "Overall\nN = 9,361",
+    stat_1_2 = "Standard\nN = 4,683",
+    stat_2_2 = "Intensive\nN = 4,678",
+    p.value_2 = "p-value",
+    stat_1_3 = "No\nN = 5,952",
+    stat_2_3 = "Yes\nN = 3,409",
+    p.value_3 = "p-value"
+  ) |> 
   add_header_row(
     values = c("Characteristic",
-               "Included in ancillary EHR study", 
-               header_labels['p.value']),
-    colwidths = c(1, 2, 1)
+               "Overall\nN = 9,361",
+               "Treatment",
+               header_labels['p.value_2'],
+               "Included in EHR ancillary study", 
+               header_labels['p.value_3']),
+    colwidths = c(1, 1, 2, 1, 2, 1)
   ) |> 
   theme_box() |> 
-  set_header_labels(stat_1 = header_labels['stat_1'],
-                    stat_2 = header_labels['stat_2']) |> 
   merge_v(part = 'header') |> 
-  align(j = c(2, 3, 4), align = 'center', part = 'all') |> 
+  align(align = 'center', part = 'all') |> 
+  align(j = 1, align = 'left', part = 'all') |> 
   footnote(
@@
     i = 1, 
-    j = ~p.value,
+    j = ~p.value_2 + p.value_3,
     part = 'header',
@@
   fontsize(size = 12, part = 'all') |> 
-  font(fontname = 'Times New Roman', part = 'all')
+  font(fontname = 'Times New Roman', part = 'all') |> 
+  width(j = ~ stat_0_1 + stat_1_2 + stat_2_2 + stat_1_3 + stat_2_3,
+        width = 1) |> 
+  width(j = 1, width = 1.75)
 
@@
 
-<!-- new page not needed at the end of a block section -->
-```{r}
+\newpage
 
-block_section(
-  prop_section(
-    page_size = page_size(orient = "portrait"),
-    type = "continuous"
-  )
-)
-
-```
-
-__Table 2__: All-cause mortality by treatment group and subgroup `r single_par`
+__eTable 2__: All-cause mortality by treatment group and subgroup `r single_par`
 
@@
 \newpage
-__Table 3__: Cardiovascular and non-cardiovascular mortality by treatment group and subgroup `r single_par`
+__eTable 3__: Cardiovascular and non-cardiovascular mortality by treatment group and subgroup `r single_par`
 
@@
 
-\newpage
-__Figure 1__: (A) Cumulative incidence of all-cause mortality by treatment group. (B) Time-dependent effect of randomization to intensive treatment for all-cause mortality. `r single_par`
-
-```{r}
-
-blank_plot <- ggplot(data.frame()) + aes() + theme_void()
-
-legend <- get_legend(
-  fig$inc_acm[[1]]$plot + 
-    theme(legend.position = c(.55, .2),
-          legend.direction = 'horizontal')
-)
-
-plot_grid(
-  # row 1 (legend)
-  legend, 
-  blank_plot,
-  # row 2 (figures)
-  fig$inc_acm[[1]]$plot + theme(legend.position = ''), 
-  fig$eff_acm[[1]],
-  # row 3 (table)
-  fig$inc_acm[[1]]$table, 
-  blank_plot,
-  ncol = 2,
-  align = 'v',
-  axis = 'lr',
-  rel_heights = c(1,10,2),
-  labels = c("","","A","B","",""),
-  label_x = 0.1,
-  label_y = 1.05
-)
-
-```
-
-\newpage
-__Figure 2__: (A) Cumulative incidence of cardiovascular and non-cardiovascular mortality by treatment group.
-(B) Time-dependent effect of randomization to intensive treatment for cardiovascular mortality.
-`r single_par`
-
-```{r}
-
-plot_grid(
-  # row 1 (legend)
-  legend, 
-  blank_plot,
-  # row 2 (figures)
-  fig$inc_cvd[[1]] + theme(legend.position = ''), 
-  fig$eff_cvd[[1]],
-  ncol = 2,
-  align = 'v',
-  axis = 'lr',
-  rel_heights = c(1,10,2),
-  labels = c("","","A","B","",""),
-  label_x = 0.1,
-  label_y = 1.05
-)
-
-```
-
-\newpage
-__Figure 3__: (A) Mean systolic blood pressure over time by treatment group.
-(B) Mean difference in systolic blood pressure over time.
-Shaded areas indicate a 95% confidence interval for the mean.
-Horizontal lines during the trial period show target blood pressure values for each treatment group `r single_par`
-
-```{r}
-
-blank_plot <- ggplot(data.frame()) + aes() + theme_void()
-
-legend <- get_legend(
-  bp_viz$levels + 
-    theme(legend.position = c(.55, .2),
-          legend.direction = 'horizontal')
-)
-
-plot_grid(
-  # row 1 (legend)
-  legend, 
-  blank_plot,
-  # row 2 (figures)
-  bp_viz$levels + theme(legend.position = ''), 
-  bp_viz$diff,
-  ncol = 2,
-  align = 'v',
-  axis = 'lr',
-  rel_heights = c(1,10,2),
-  labels = c("","","A","B","",""),
-  label_x = 0.1,
-  label_y = 1.05
-)
-
-# fpath <- file.path(
-#   paste0(get_sasinet_drive(), ":"),
-#   'bjaeger',
-#   'sprint-mortality-ndi',
-#   'fig',
-#   'SPRINT_EHR_SBP_LongTerm_Comparison_122721.png'
-# )
-# 
-# knitr::include_graphics(
-#   path = fpath, 
-#   dpi = 600
-# )
-
-```
-
-```{r}
-block_section(
-  prop_section(
-    page_size = page_size(orient = "landscape"),
-    page_margins = page_mar(bottom = 1/8, top = 1/8, 
-                            right = 1/4, left = 1/4),
-    type = "continuous"
-  )
-)
-```
-
-__Online supplement only__
-
-Byron C.
-Jaeger, PhD,^1^ Adam P.
-Bress, PharmD, MS,^2,3^ Joshua D.
-Bundy, PhD,^4^ Alfred K.
-Cheung, MD,^5,6^ William C.
-Cushman, MD,^7^ Paul E.
-Drawz, MD,^8^ Lawrence J.
-Fine, MD,^9^ Karen C.
-Johnson, MD,^7^ Cora E.
-Lewis, MD,^10^ Suzanne Oparil, MD,^11^ Michael V.
-Rocco, MD,^12^ Stephen R.
-Rapp, PhD, ^13^ Joni K.
-Snyder, RN,BSN,^9^ Mark A.
-Supiano, MD,^14^ Paul K.
-Whelton, MD,^4^ Jeff D.
-Williamson, MD, MHS,^15^ Jackson T.
-Wright, Jr, MD, PhD,^16^ David M.
-Reboussin, PhD,^1^ and Nicholas M.
-Pajewski, PhD^1^
-
-^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC.
-^2^Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT.
-^3^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT.
-^4^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
-^5^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT.
-^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT.
-^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis.
-^8^Division of Renal Diseases & Hypertension, University of Minnesota, MN.
-^9^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD.
-^10^Department of Epidemiology, University of Alabama at Birmingham.
-^11^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham.
-^12^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^13^Departments of Psychiatry and Behavioral Medicine & Social Sciences and Health Policy, Wakfe Forest School of Medicine, Winston-Salem, NC.
-^14^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
-^15^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^16^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
-\
-
-```{r}
-block_section(
-  prop_section(
-    page_size = page_size(orient = "portrait"),
-    type = "continuous"
-  )
-)
-```
-
-__eTable 1__: Characteristics of study participants stratified by inclusion in the electronic health record ancillary study and by treatment group. `r single_par`
-
-```{r}
-
-header_labels <- tbl_characteristics |> 
-  getElement('supp') |> 
-  show_header_names(quiet = TRUE) |> 
-  mutate(
-    label = str_remove_all(label, fixed("*")),
-    label = str_replace(label, ', ', '\n')
-  ) |> 
-  deframe()
-
-tbl_characteristics |> 
-  getElement('supp') |> 
-  modify_footnote(update = everything() ~ NA) |> 
-  as_flex_table() |> 
-  theme_box() |> 
-  width(width = 0.95) |> 
-  width(j=1, width = 1.75) |> 
-  set_header_labels(stat_1_1 = header_labels['stat_1_1'],
-                    stat_1_2 = header_labels['stat_1_2'],
-                    stat_2_1 = header_labels['stat_2_1'],
-                    stat_2_2 = header_labels['stat_2_2']) |> 
-  align(j = 2:7, align = 'center', part = 'all') |> 
-  footnote(
-    i = 1, 
-    j = 1, 
-    ref_symbols = '',
-    value = as_paragraph(
-      "Table values are percentages unless otherwise specified"
-    )
-  ) |> 
-  footnote(
-    i = 2, 
-    j = ~ p.value_1 + p.value_2,
-    part = 'header',
-    ref_symbols = 'a',
-    value = as_paragraph(
-      "P-values computed using Wilcoxon rank sum test for age in years, chi-square test for age \u226575 years, and chi-square test for all other variables"
-    )
-  ) |> 
-  footnote(
-    i = ~ label == "Chronic Kidney Disease",
-    j = 1, 
-    ref_symbols = 'b',
-    value = ckd_defn,
-    part = 'body'
-  ) |> 
-  footnote(
-    i = 1,
-    j = 1,
-    ref_symbols = '',
-    value = abbrvs_write(abbrvs[c('SD', 'EHR', 'IQR', 'MoCA', 'FI')]),
-    part = 'body'
-  ) |>
-  fontsize(size = 12, part = 'all') |> 
-  font(fontname = 'Times New Roman', part = 'all')
-
-```
-
-\newpage
-
-__eFigure 1__: All-cause mortality hazard ratio for participants randomized to intensive versus standard treatment.
-
-```{r}
-
-long_viz_mort
-
-```
-
-\newpage
-__eFigure 2__: Cardiovascular mortality hazard ratio for participants randomized to intensive versus standard treatment.
-
-```{r}
-
-long_viz_cvd
-
-```
-
 ```{r}

### 2022-01-18 R/ndi_tabulate_characteristics.R

@@ -35,20 +35,23 @@ ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
   }
 
   tbl_by_anci <- ndi_tbl_data |>
-    relocate(treatment, .after = race) |>
+    relocate(sbp, treatment, .after = race) |>
     tbl_summary(
       by = 'ehr_ancillary',
       statistic = list(
-        age_yrs ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
+        all_continuous() ~ c("{mean} ({sd})"),
         all_categorical() ~ "{p}"
       ),
+      digits = list(everything() ~ 1),
       label = list(age_cat ~ paste("Age", recoders$levels$age_cat[2]),
                    sex ~ recoders$levels$sex[2],
+                   sbp ~ recoders$variable['sbp'],
                    treatment ~ "Intensive treatment",
                    race ~ recoders$levels$race[2],
                    moca ~ paste("MoCA", recoders$levels$moca[2])),
-      type = list(age_yrs ~ "continuous2",
+      type = list(age_yrs ~ "continuous",
                   age_cat ~ "dichotomous",
+                  sbp ~ 'continuous',
                   treatment ~ "dichotomous",
                   sex ~ "dichotomous",
                   race ~ "dichotomous",
@@ -66,21 +69,24 @@ ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
     add_p()
 
   tbl_by_tx <- ndi_tbl_data |>
-    relocate(ehr_ancillary, .after = race) |>
+    relocate(sbp, treatment, ehr_ancillary, .after = race) |>
     tbl_summary(
       by = 'treatment',
       statistic = list(
-        age_yrs ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
+        all_continuous() ~ c("{mean} ({sd})"),
         all_categorical() ~ "{p}"
       ),
+      digits = list(everything() ~ 1),
       label = list(age_cat ~ paste("Age", recoders$levels$age_cat[2]),
                    sex ~ recoders$levels$sex[2],
+                   sbp ~ recoders$variable['sbp'],
                    ehr_ancillary ~ "Included in EHR ancillary study",
                    race ~ recoders$levels$race[2],
                    moca ~ paste("MoCA", recoders$levels$moca[2])),
-      type = list(age_yrs ~ "continuous2",
+      type = list(age_yrs ~ "continuous",
                   age_cat ~ "dichotomous",
[…]

### 2022-01-18 R/recoders_make.R

@@ -11,6 +11,7 @@ recoders_make <- function() {
     variable = c("overall" = "Overall",
                  "age_cat" = "Age group",
                  "age_yrs" = "Age, years",
+                 "sbp" = "Systolic blood pressure",
                  "sex" = "Sex",
                  "race" = "Race",
                  "ckd_2021" = "Chronic Kidney Disease",

### 2022-01-18 _targets.R

@@ -35,6 +35,8 @@ list(
              fname = 'longterm_death.csv',
              # participant identifier
              pid,
+             # systolic blood pressure at baseline
+             sbp,
              # study site
              randSite,
              # all cause mortality time and statue

### 2022-01-18 doc/manuscript.Rmd
splitting bp fig; easier to read

@@
 A total of `r ap_b_i$n_acm` and `r ap_b_s$n_acm` deaths occurred among participants randomized to intensive and standard treatment, respectively (__eTable 2__).
-The hazard ratio (HR) for all-cause mortality comparing intensive to standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial, and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during observational follow-up following the trial.
+The hazard ratio (HR) for all-cause mortality comparing intensive to standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial phase, and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during the cohort phase.
 The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 1__).
@@
   getElement('data') |> 
-  filter(time_months == 60)
+  filter(source == 'EHR', 
+         time_months == 60)
 
-bp_viz_5yr <- split(bp_viz_5yr, bp_viz_5yr$group)
+bp_viz_5yr <- split(bp_viz_5yr, bp_viz_5yr$treatment)
 
@@
   getElement('data') |> 
-  filter(time_months == '120')
+  filter(source=='EHR',
+         time_months == 120)
 
-bp_viz_10yr <- split(bp_viz_10yr, bp_viz_10yr$group)
+bp_viz_10yr <- split(bp_viz_10yr, bp_viz_10yr$treatment)
 
@@
   getElement('data') |> 
-  filter(find_signchange_lgl(Lower)) |> 
+  filter(source == 'EHR',
+         find_signchange_lgl(Lower)) |> 
   pull(time_months) |> 
@@
 ___Blood Pressure___.
-The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization (__Figure 5__; panel A).
+The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization (__Figure 5__).
 For participants randomized to standard treatment, mean (95% CI) SBP was estimated to be `r f(bp_viz_5yr$Standard)` at 5 years and `r f(bp_viz_10yr$Standard)` post-randomization.
-The difference in mean SBP levels between participants randomized to intensive versus standard treatment, in mm Hg, was `r f(bp_diff_5yr)` at 5 years post-randomization, and was attenuated by `r t_atten` years post-randomization (__Figure 5__; Panel B).
+The difference in mean SBP levels between participants randomized to intensive versus standard treatment, in mm Hg, was `r f(bp_diff_5yr)` at 5 years post-randomization, and was attenuated at `r t_atten` years post-randomization (__eFigure 1__).
 
@@
 \newpage
-__Figure 5__: (A) Mean systolic blood pressure over time by treatment group.
-(B) Mean difference in systolic blood pressure over time.
-Shaded areas indicate a 95% confidence interval for the mean.
-Horizontal lines during the trial period show target blood pressure values for each treatment group `r single_par`
+__Figure 5__: Mean systolic blood pressure over time by treatment group. Shaded areas indicate a 95% confidence interval for the mean. `r single_par`
 
@@
 
-blank_plot <- ggplot(data.frame()) + aes() + theme_void()
-
-legend <- get_legend(
 bp_viz$levels + 
-    theme(legend.position = c(.55, .2),
-          legend.direction = 'horizontal')
-)
-
-plot_grid(
-  # row 1 (legend)
-  legend, 
-  blank_plot,
-  # row 2 (figures)
-  bp_viz$levels + theme(legend.position = ''), 
-  bp_viz$diff,
-  ncol = 2,
-  align = 'v',
-  axis = 'lr',
-  rel_heights = c(1,10,2),
-  labels = c("","","A","B","",""),
-  label_x = 0.1,
-  label_y = 1.05
-)
+  theme(legend.key.size = unit(1, 'cm'))
 
@@
 
+\newpage
+
+__eFigure 1__: Mean difference in systolic blood pressure over time. Shaded areas indicate a 95% confidence interval for the mean. `r single_par`
+
+```{r}
+
+bp_viz$diff + 
+  theme(legend.key.size = unit(1, 'cm'))
+
+```
+
 ```{r}

### 2022-01-18 R/bp_long_viz.R

@@ -8,13 +8,15 @@
 bp_long_viz <- function(date = "01-15-22") {
 
   cols_bg <- viridis(3)
-  cols_tx <- c("#DF8F44FF","#ED0000FF","#79AF97FF","#00468BFF")
+  cols_tx <- pal_lancet()(2)
   cols_diff <- c("maroon4","black")
 
   .date <- str_remove_all(date, fixed('-'))
 
   fname <- paste0("SPRINT_EHR_SBP_GLIMMIX_LongTerm_", .date, ".csv")
   fname2 <- paste0("SPRINT_EHR_SBP_GLIMMIX_JAMA_IM_REDO_", .date, ".csv")
+
+  # Blood pressure levels over time ----
 
   bp <- read_csv(file.path('ehr', fname)) |>
     select(-Statement, -Alpha, -tValue, -DF, -StdErr) |>
@@ -33,8 +35,8 @@ bp_long_viz <- function(date = "01-15-22") {
   bp_levels <- filter(bp, group %in% c("Standard (EHR)",
                                        "Intensive (EHR)") & time_months>54)
 
-  bp_levels_trial <- filter(bp_trial, group %in% c("STD",
-                                                   "INT") & source=="EHR") |>
+  bp_levels_trial <- bp_trial |>
+    filter(group %in% c("STD", "INT") & source=="EHR") |>
     select(-source) |>
     mutate(group = recode(group,
                           "STD" = "Standard (EHR)",
@@ -50,52 +52,67 @@ bp_long_viz <- function(date = "01-15-22") {
                           "STD" = "Standard (trial)",
                           "INT" = "Intensive (trial)",
                           "DIFF" = "Difference (trial)"))
-  bp_levels<-rbind(bp_levels_trial, bp_levels)
+
+  bp_levels<-rbind(bp_levels_trial, bp_levels) |>
+    separate(group,
+             into = c('treatment', 'source'),
+             sep = ' ') |>
+    mutate(source = str_remove_all(source, pattern = '\\(|\\)'),
+           source = recode(source, 'trial' = 'Trial'),
+           treatment = factor(treatment, levels = c("Standard", "Intensive")))
+
+  # Blood pressure difference over time ----
 
   bp_diff_orig <- filter(bp, group == 'Difference (EHR)' & time_months>54)
-  bp_diff_ehr<- filter(bp_trial, group=="DIFF" & source=="EHR") |> select(-source) |>
+
[…]

### 2022-01-20 README.Rmd
first full draft!

--- 2022-01-18 doc/manuscript.Rmd
@@
 ---
-title: "Longer term All-Cause and Cardiovascular Mortality with Intensive Blood Pressure Control: A Secondary Analysis of SPRINT"
-bibliography: refs.bib
-csl: jama.csl
-always_allow_html: true
-output: 
-  officedown::rdocx_document:
-    reference_docx: style_manuscript_times_new_roman.docx
+output: github_document
 ---
-```{r setup, include=FALSE}
+
+<!-- README.md is generated from README.Rmd. Please edit that file -->
+
+```{r, include = FALSE}
 
 knitr::opts_chunk$set(echo = FALSE, 
+                      collapse = TRUE,
+                      comment = "#>",
                       message = FALSE,
@@
 
-# specification for table tables
+# specification for tables
 rspec <- round_spec() |>
@@
 
-# default is double space 
-# single spacing for title page, abstract, captions
-single_par <- fp_par(line_spacing=1)
-
 # make the . symbol equivalent to calling table_glue.
@@
 
-Byron C.
-Jaeger, PhD,^1^ Adam P.
-Bress, PharmD, MS,^2,3^ Joshua D.
-Bundy, PhD,^4^ Alfred K.
-Cheung, MD,^5,6^ William C.
-Cushman, MD,^7^ Paul E.
-Drawz, MD,^8^ Lawrence J.
-Fine, MD,^9^ Karen C.
-Johnson, MD,^7^ Cora E.
-Lewis, MD,^10^ Suzanne Oparil, MD,^11^ Michael V.
-Rocco, MD,^12^ Stephen R.
-Rapp, PhD, ^13^ Joni K.
-Snyder, RN,BSN,^9^ Mark A.
-Supiano, MD,^14^ Paul K.
-Whelton, MD,^4^ Jeff D.
-Williamson, MD, MHS,^15^ Jackson T.
-Wright, Jr, MD, PhD,^16^ David M.
-Reboussin, PhD,^1^ and Nicholas M.
-Pajewski, PhD^1^
+# sprint-mortality-ndi
 
-^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC.
-^2^Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT.
-^3^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT.
-^4^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
-^5^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT.
-^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT.
-^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis.
-^8^Division of Renal Diseases & Hypertension, University of Minnesota, MN.
-^9^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD.
-^10^Department of Epidemiology, University of Alabama at Birmingham.
-^11^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham.
-^12^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^13^Departments of Psychiatry and Behavioral Medicine & Social Sciences and Health Policy, Wakfe Forest School of Medicine, Winston-Salem, NC.
-^14^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
-^15^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^16^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
-\
-
-**Correspondence to:** \
-Nicholas M. Pajewski, PhD \
-Department of Biostatistics and Data Science \
-Division of Public Health Sciences \
-Wake Forest School of Medicine \
-Medical Center Boulevard \
-Winston-Salem, NC 27154 \
-(336) 713-1396 \
-npajewsk@wakehealth.edu `r single_par`
-
-**Word Count:** Abstract: ; Manuscript:
+<!-- badges: start -->
+<!-- badges: end -->
 
@@
 
-```
-
-\newpage
-
-# ABSTRACT
-
-__IMPORTANCE__.
-Both the Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) trial have shown benefits of intensive blood pressure control on cardiovascular morbidity and mortality.
-However, as both trials were ended after slightly more than 3 years of follow-up, the effect of intensive treatment on longer term mortality is unknown.
-
-__OBJECTIVE__. To evaluate the legacy effect of intensive hypertension treatment on longer term all-cause and cardiovascular mortality.
-
-__DESIGN, SETTING, AND PARTICIPANTS__.
-SPRINT, a randomized clinical trial of 9361 patients aged 50 years or older with hypertension and increased cardiovascular risk, but without diabetes or history of stroke.
-Randomization began on November 8, 2010, the trial intervention was stopped early on August 20, 2015, and trial follow-up visits occurred through July 2016.
-
-__INTERVENTIONS__. Randomization to a systolic blood pressure goal of either less than 120 mm Hg (intensive treatment, n=4678) versus less than 140 mm Hg (standard treatment, N=4683).
-
-__MAIN OUTCOMES AND MEASURES__.
-Cardiovascular and all-cause mortality assessed via the US National Death Index, beginning in 2016 through December 31st, 2020.
-Outpatient blood pressures measured in routine clinical practice after the trial were examined in a subset of trial participants (N=3644).
-
-__RESULTS__.
-Among 9361 randomized participants (mean age, 67.9 years; 3332 women [35.6%]), the median intervention period was 3.34 years.
-Over a median follow-up of `r ap_b_o$fup_est` years, intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial.
-However, there was no indication of benefit during post-trial follow-up for either cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
-Results were similar for subgroups based on baseline age, cognitive function, and frailty status.
-Analyses of outpatient blood pressures indicated a steady decline in the mean between group difference following the trial, largely driven by increases in mean systolic blood pressure in participants randomized to intensive treatment, increasing from a mean of 131.7 mm Hg (95% CI 130.7 to 132.6 mm Hg) at 4.5 years of follow-up to 136.8 mm Hg (95% CI 135.7 to 137.8 mm Hg) at 9 years of follow-up.
-
-__CONCLUSIONS AND RELEVANCE__.
-The observed benefit of intensive treatment on cardiovascular and all-cause mortality was largely attenuated during post-trial observational follow-up.
-Given increasing blood pressures in participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension.
-
-\newpage
-
-# INTRODUCTION
-
-Hypertension is the leading modifiable risk factor for cardiovascular disease (CVD).[@gdbJACC2020] A meta analysis of pooled individual-level data from randomized trials has shown that pharmacological blood pressure (BP) reduction reduces the risk of major cardiovascular events across the spectrum of initial BP, with larger absolute risk reduction for older adults.[@rahimi_age_stratified_2021] The Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) each showed that intensive systolic BP (SBP) control, defined by a SBP target < 120 mm Hg in SPRINT and < 130 mm Hg in STEP, reduced the risk for cardiovascular morbidity and mortality.
-[@sprint_research_group_randomized_2015; @zhang_trial_2021] However,
-only SPRINT showed a benefit of intensive treatment on all-cause mortality.
-Another contrasting result was that intensive treatment in SPRINT, but not STEP, increased the risk of acute kidney injury (AKI).
-While this may be attributable to the much lower prevalence of chronic kidney disease in STEP (2.3% vs 28.3%), the consistent association of AKI with increased mortality risk[@chawla2014aki; @rocco2018aki] raises questions about the longer term effect of intensive treatment on CVD and all-cause mortality.
-
-Because both trials were stopped early after a median follow-up of 3 years, the objective of the current study was to estimate the longer term, legacy effect of randomization to intensive treatment on cardiovascular and all-cause mortality risk for participants in SPRINT.
-A secondary objective, in a subset of trial participants, was to examine attained SBP following the conclusion of trial follow-up.
-To accomplish these objectives, we linked SPRINT participants to the National Death Index (NDI) from 2016 to 2020, and extracted longitudinal outpatient measurements of SBP available in the electronic health record (EHR) over the same time period.
-
-# METHODS
-
-__Trial Design__: The design and methods of SPRINT have been published previously.[@ambrosius_design_2014; @sprint_research_group_randomized_2015] Briefly, it was a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk of CVD.
-Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed.
-Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (CKD; defined by an estimated glomerular filtration rate of <60 mL/min/1.73 m^2^), or a Framingham Risk Score of 15% or greater or if they were aged 75 years or older.
-Individuals residing in a nursing home, persons with a diagnosis of dementia (based on medical record review), and those treated with medications primarily used for dementia therapy were excluded, as were persons with prevalent diabetes mellitus, history of stroke, proteinuria > 1 gram per day, or polycystic kidney disease.
-Individuals at 102 sites in the United States and Puerto Rico were randomized (1:1) to a SBP goal of less than 120 mm Hg (intensive treatment group, n = 4678) or a goal of less than 140 mm Hg (standard treatment group, n = 4683), using random permuted blocks with the randomization stratified by clinic site.
-The algorithms and formulary for the trial are listed in the published study protocol ( __citations needed__?
-).
-Trial enrollment began in November 2010 and ended in March 2013, with active follow-up through July 1, 2016.[@sprintFinal] The study was approved by the institutional review board at each participating site, and each participant provided written informed consent.
-The study is registered at ClinicalTrials.gov (NCT01206062).
-
-__Baseline Study Measurements__: Sociodemographic data were collected at baseline, with race or ethnicity information collected via self-report.
-The estimated glomerular filtration rate (eGFR) was calculated by the race-free 2021 CKD-EPI creatinine equation.[@inker_new_2021] Cognitive function was assessed using the Montreal Cognitive Assessment (MoCA) [@nasreddine_montreal_2005].
-Lower cognitive function was defined as scoring at or below the estimated age and education-specific normative 10th percentile from the Irish Longitudinal Study of Aging,[@kenny_normative_2013] after adding 3 points to the scores of non-White participants.[@sachsMoCA2021] We defined frailty status at baseline using a 36-item Frailty Index (FI) based upon the model of deficit accumulation.[@pajewski_characterizing_2016] The FI is calculated as the sum of the score for each deficit divided by the total number of nonmissing items.
-We categorized frailty status as fit (FI ≤ 0.10), less fit (0.10 < FI ≤ 0.21), or frail (FI > 0.21).
-
-__National Death Index Linkage__: Outcomes of interest included all-cause and CVD mortality.
-Methods of ascertainment and adjudication through the course of trial follow-up have been previously described.[@sprint_research_group_randomized_2015; @sprintFinal] Subsequently, mortality was ascertained through a US National Death Index (NDI) search.
-Possible matches were identified according to NDI guidelines.[@ndi_guide] Deaths were treated as confirmed if they were a Class 1 match, or a Class 2, 3, or 4 match with a probabilistic score above cutoffs recommended by the NDI.[@ndi_guide] NDI follow-up began in 2016 and ended on the date of death or December 31, 2020.
-Deaths ascertained in 2020 were based on the NDI preliminary data release.
-CVD mortality for NDI-based follow-up used the NDI Plus System, which automatically identifies underlying causes of death from death certificates, including conversion to ICD-10 codes.
-we defined CVD mortality as any death containing the ICD-10 codes of I00 to I99.[@regards_ndi]
-
-__EHR Ancillary Study__: We examined the trajectory of SBP following the conclusion of the trial using outpatient SBPs extracted from the EHR.
-Methods for the linkage of participants to their medical record number and the extraction of vital sign data have been previously described.[@drawz_concordance_2020] Because encounter type information was inconsistently available (i.e.
-outpatient, inpatient, observation, etc.), we defined a BP measurement as outpatient if there were was not a BP measurement on the preceding or following day, and if there were 2 or less BP measurements on a particular day.
-We averaged outpatient EHR BP readings when there were 2 on the same day.
-
-__Statistical Analysis__: Given the a priori expectation that treatment group differences may not be constant as a function of follow-up time (i.e.
-the proportional hazards assumption was likely to be invalid), we modeled treatment group differences as a function of time using two approaches.
-The first approach used Cox proportional hazards regression but split follow-up time into non-overlapping trial and cohort phases, estimating separate treatment group differences for each phase.[@zhang_time_varying_2018].
-The second approach estimated the effect of treatment group as a continuous time-dependent effect.[@martinussen2007dynamic; @therneau2017using] All analyses accounted for correlation within study sites,[@lin_robust_1989] and analyses of cardiovascular mortality accounted for the competing risk of non-cardiovascular mortality.[@scheike_flexible_2008].
-Mean between-group differences in outpatient SBP following the conclusion of trial follow-up were estimated using linear mixed models.
-Models included random effects for participant and clinic site and an interaction between treatment group and time since randomization, which was flexibly modeled using B-splines.
-All analyses were performed using SAS version 9.4 (SAS Institute Inc, Cary, NC) and R version `r paste(version$major, version$minor, sep = '.')` (R Project for Statistical Computing [http://www.r-project.org]) with multiple auxiliary R packages.[@r_language; @table.glue; @tidyverse; @survival_package; @timereg; @targets] All hypothesis tests were 2-sided, and P values less than 0.05 were considered statistically significant.
-No adjustments for multiple comparisons were made.
-
-# RESULTS
-
-___Study Participants___.
-A total of `r nrow(ndi_baseline)` participants were randomized between November 2010 and March 2013.
-The mean (standard deviation [SD]) age was `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years, with `r inline_text(tbl_characteristics, variable = 'age_cat', column = 'stat_0_1')`% of participants aged 75 years or older (__eTable 1__).
-Participants were `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% female and `r inline_text(tbl_characteristics, variable = 'race', column = 'stat_0_1')`% black.
-The mean (SD) SBP at baseline was `r inline_text(tbl_characteristics, variable = 'sbp', column = 'stat_0_1')` and `r inline_text(tbl_characteristics, variable = 'moca', column = 'stat_0_1')`% of participants MoCA scores were below the 10^th^ percentile.
-Participants included in the ancillary EHR study had greater age, lower SBP, less cognitive impairment, higher prevalence of CKD, and were more likely to be male or black than participants not included.
-
-```{r}
-
-# `r inline_text(tbl_characteristics, variable = 'moca', column = 'stat_0_1')`
-
-acm_overall_epochs <- fig |> 
-  filter(variable == 'overall') |> 
-  pull(eff_acm) |> 
-  unlist(recursive = FALSE) |> 
-  getElement('data') |> 
-  filter(find_signchange_lgl(hr_upr-1)) |> 
-  mutate(label = table_value(time))
-
-t_signif <- acm_overall_epochs$label[1] 
-t_atten <- acm_overall_epochs$label[2]
-
-```
-
-___All-cause Mortality___.
-In both treatment groups, median follow-up time was `r ap_b_o$fup_est` years.
-A total of `r ap_b_i$n_acm` and `r ap_b_s$n_acm` deaths occurred among participants randomized to intensive and standard treatment, respectively (__eTable 2__).
-The hazard ratio (HR) for all-cause mortality comparing intensive to standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial phase, and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during the cohort phase.
-The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 1__).
-In subgroups based on age, sex, race, CKD, cognitive function, and frailty, there was no evidence that intensive treatment during the trial phase had benefit for all-cause mortality during the cohort phase (__Figure 2__).
-
-```{r}
-
-cvd_overall_epochs <- fig |> 
-  filter(variable == 'overall') |> 
-  pull(eff_cvd) |> 
-  unlist(recursive = FALSE) |> 
-  getElement('data') |> 
-  slice(-1) |> 
-  filter(find_signchange_lgl(hr_upr-1)) |> 
-  mutate(label = table_value(time))
-
-t_signif <- cvd_overall_epochs$label[1] 
-t_atten  <- cvd_overall_epochs$label[2]
-
-```
-
-___Cardiovascular Mortality___.
-A total of `r ap_b_i$n_cvd` and `r ap_b_s$n_cvd` CVD mortality events occurred among participants randomized to intensive and standard treatment, respectively (__eTable 3__).
-The HR for CVD mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the cohort phase.
-The time-dependent effect of intensive versus standard treatment indicated a benefit for CVD mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort period (__Figure 3__).
-Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase had benefit for CVD mortality during the cohort phase (__Figure 4__).
-
-```{r}
-
 bp_viz_5yr <- bp_viz |> 
@@
 
-# a temporary function for inline reporting below.
+# a temporary function for inline reporting of BP values
 f <- function(d) with(d, .("{Estimate} ({Lower}, {Upper})"))
@@
 
-t_atten <- bp_viz |> 
+t_atten_sbp <- bp_viz |> 
   getElement('diff') |> 
@@
 
-___Blood Pressure___.
-The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization (__Figure 5__).
-For participants randomized to standard treatment, mean (95% CI) SBP was estimated to be `r f(bp_viz_5yr$Standard)` at 5 years and `r f(bp_viz_10yr$Standard)` post-randomization.
-The difference in mean SBP levels between participants randomized to intensive versus standard treatment, in mm Hg, was `r f(bp_diff_5yr)` at 5 years post-randomization, and was attenuated at `r t_atten` years post-randomization (__eFigure 1__).
+\newpage
 
-# DISCUSSION
+# ABSTRACT
 
-A striking aspect of our results is the quickly weakening level of BP control for participants randomized to intensive treatment.
-While an attenuation of the between-group BP delta subsequent to the trial was certainly expected, one hypothesis was that such an attenuation would be driven by participants randomized to standard treatment pursuing a lower BP goal, given the results of SPRINT and subsequent changes to hypertension guidelines.
-However, this is clearly not what occurred.
-While we do not have access to prescription records to know how participant medication regimes may have changed after the trial, these results likely show some contribution of clinician therapeutic inertia, which has been identified as a significant barrier to improving population level control of hypertension.
-Combined with evidence showing recent decreases in the population of prevalence of controlled hypertension in the US,[@muntner2020htn] our results highlight the sobering reality facing the hypertension community.
-Sustainability is a clear limiting factor, especially with trying to implement lower BP goals and interventions earlier in adulthood.
+__IMPORTANCE__.
+The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a systolic blood pressure (SBP) goal of <120 mmHg, reduced the risk of cardiovascular morbidity and mortality.
+However, the post-trial association of intensive treatment with mortality is unknown.
 
-This study has several limitations.
-First, while we restricted analyses to high quality NDI matches, some small degree of misclassification in linking participants to the NDI is likely.
-Second, while several studies have shown reasonable performance of using NDI diagnosis codes for defining CVD mortality, it is clearly not as robust as the adjudication process used in the primary follow-up for the trial.
-Third, information about BP control after the trial was limited to routine outpatient BPs extracted from the EHR, which are well known to poorly reflect the standardized BP measurement process used during the trial.[@drawz_concordance_2020]
+__OBJECTIVE__.
+To evaluate the association of receiving intensive treatment for approximately three years during the SPRINT trial with all-cause and cardiovascular mortality up to ten years post-randomization.
 
-# ACKNOWLEDGMENTS 
-The views expressed in this paper are those of the authors and do not represent the official position of the National Institutes of Health (NIH), the National Heart, Lung, and Blood Institute, the Department of Veterans Affairs, or the U.S.
-Government, or the SPRINT Research Group.
-This paper was not reviewed by the SPRINT Publications and Presentations Committee.
-The authors also wish to acknowledge computing support provided the Veterans Affairs Informatics and Computing Infrastructure (VINCI).
+__DESIGN, SETTING, AND PARTICIPANTS__.
+SPRINT, a randomized clinical trial of 9361 patients aged 50 years or older with hypertension and increased cardiovascular risk, but without diabetes or history of stroke.
+Randomization began on November 8, 2010, the trial intervention ended on August 20, 2015, and trial follow-up visits occurred through July 2016.
 
-# FINANCIAL DISCLOSURE 
-The work presented here was supported by R01HL136679, R01AG055606, and R01AG065805.
-The Systolic Blood Pressure Intervention Trial was funded by the National Institutes of Health (including the National Heart, Lung, and Blood Institute, the National Institute of Diabetes and Digestive and Kidney Diseases, the National Institute on Aging, and the National Institute of Neurological Disorders and Stroke) under contracts HHSN268200900040C, HHSN268200900046C, HHSN268200900047C, HHSN268200900048C, and HHSN268200900049C and interagency agreement A-HL-13-002-001.
-It was also supported in part with resources and use of facilities through the Department of Veterans Affairs.
-Azilsartan and chlorthalidone (combined with azilsartan) were provided by Takeda Pharmaceuticals International Inc.
+__INTERVENTIONS__. Randomization to SBP goal of either <120 mmHg (intensive treatment, N=4678) versus <140 mmHg (standard treatment, N=4683).
 
-Additional support was provided through the following National Center for Advancing Translational Sciences clinical and translational science awards: UL1TR000439 (awarded to Case Western Reserve University); UL1RR025755 (Ohio State University); UL1RR024134 and UL1TR000003 (University of Pennsylvania); UL1RR025771 (Boston University); UL1TR000093 (Stanford University); UL1RR025752, UL1TR000073, and UL1TR001064 (Tufts University); UL1TR000050 (University of Illinois); UL1TR000005 (University of Pittsburgh); U54TR000017-06 (University of Texas Southwestern Medical Center); UL1TR000105-05 (University of Utah); UL1 TR000445 (Vanderbilt University); UL1TR000075 (George Washington University); UL1 TR000002 (University of California, Davis); UL1 TR000064 (University of Florida); and UL1TR000433 (University of Michigan); and by National Institute of General Medical Sciences, Centers of Biomedical Research Excellence award NIGMS P30GM103337 (awarded to Tulane University).
+__MAIN OUTCOMES AND MEASURES__.
+All-cause and cardiovascular mortality assessed via the US National Death Index, beginning in 2016 through December 31st, 2020.
+In a subset of 3644 trial participants, outpatient SBP levels measured in routine clinical practice after the trial were examined.
 
-```{r tbl_footers}
+__RESULTS__.
+Among 9361 randomized participants (mean [standard deviation] age `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years; `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% women), the median intervention phase was 3.3 years.
+Over a median follow-up of `r ap_b_o$fup_est` years, intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% confidence interval [CI] {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial.
+However, there was no evidence of association during post-trial follow-up for cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
+The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization.
 
-ckd_defn <- as_paragraph("Chronic Kidney Disease was defined by an estimated glomerular filtration rate <60 ml/min/1.73 m", as_sup(2L), " based on the 2021 CKD-EPI creatinine equation.")
+__CONCLUSIONS AND RELEVANCE__.
+The beneficial effect of intensive treatment on cardiovascular and all-cause mortality was attenuated during post-trial observational follow-up.
+Given increasing SBP levels in participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension.
 
-```
+__CENTRAL ILLUSTRATION__
 
-```{r tbl_abbrvs}
-
-abbrvs <- list(
-  CI   = 'confidence interval',
-  FI   = 'frailty index',
-  SD   = 'standard deviation',
-  IQR  = 'interquartile range',
-  MoCA = 'Montreal cognitive assessment',
-  EHR  = 'electronic health records',
-  CVD  = 'cardiovascular disease'
-)
-
-abbrvs_paste <- function(strings){
-
-  front <- 'Abbreviations'
-
-  matter <- glue::glue_collapse(strings, sep = '; ', last = '; and ')
-
-  paste(front, matter, sep = ': ')
-
-}
-
-abbrvs_write <- function(abbr){
-
-  sorted <- abbr[sort(names(abbr))]
-
-  strings <- map2_chr(names(sorted), sorted, paste, sep = ' = ')
-
-  as_paragraph(abbrvs_paste(strings))
-
-}
-
-```
-
-<!-- new page not needed at the end of a block section -->
-```{r}
-
-margins_normal <- page_mar(bottom = 2.54 * 0.3937, 
-                           top    = 2.54 * 0.3937, 
-                           right  = 2.54 * 0.3937, 
-                           left   = 2.54 * 0.3937)
-
-block_section(
-  prop_section(
-    page_size = page_size(orient = "portrait"),
-    type = "continuous",
-    page_margins = margins_normal
-  )
-)
-
-```
-
-__Figure 1__: (A) Cumulative incidence of all-cause mortality by treatment group. (B) Time-dependent effect of randomization to intensive treatment for all-cause mortality. `r single_par`
+Panel A: Cumulative incidence of cardiovascular and non-cardiovascular mortality by treatment group.
+Panel B: Time-dependent effect of randomization to intensive treatment for cardiovascular mortality.
 
@@
 )
-
-plot_grid(
-  # row 1 (legend)
-  legend, 
-  blank_plot,
-  # row 2 (figures)
-  fig$inc_acm[[1]]$plot + theme(legend.position = ''), 
-  fig$eff_acm[[1]],
-  # row 3 (table)
-  fig$inc_acm[[1]]$table, 
-  blank_plot,
-  ncol = 2,
-  align = 'v',
-  axis = 'lr',
-  rel_heights = c(1,10,2),
-  labels = c("","","A","B","",""),
-  label_x = 0.1,
-  label_y = 1.05
-)
-
-```
-
-\newpage
-__Figure 2__: All-cause mortality hazard ratio for participants randomized to intensive versus standard treatment.
-
-```{r}
-
-long_viz_mort
-
-```
-
-\newpage
-__Figure 3__: (A) Cumulative incidence of cardiovascular and non-cardiovascular mortality by treatment group.
-(B) Time-dependent effect of randomization to intensive treatment for cardiovascular mortality.
-`r single_par`
-
-```{r}
 
@@
 ```
-
-\newpage
-__Figure 4__: Cardiovascular mortality hazard ratio for participants randomized to intensive versus standard treatment.
-
-```{r}
-
-long_viz_cvd
-
-```
-
-\newpage
-__Figure 5__: Mean systolic blood pressure over time by treatment group. Shaded areas indicate a 95% confidence interval for the mean. `r single_par`
-
-```{r}
-
-bp_viz$levels + 
-  theme(legend.key.size = unit(1, 'cm'))
-
-# fpath <- file.path(
-#   paste0(get_sasinet_drive(), ":"),
-#   'bjaeger',
-#   'sprint-mortality-ndi',
-#   'fig',
-#   'SPRINT_EHR_SBP_LongTerm_Comparison_122721.png'
-# )
-# 
-# knitr::include_graphics(
-#   path = fpath, 
-#   dpi = 600
-# )
-
-```
-
-```{r}
-block_section(
-  prop_section(
-    page_size = page_size(orient = "landscape"),
-    page_margins = page_mar(bottom = 1/8, top = 1/8, 
-                            right = 1/4, left = 1/4),
-    type = "continuous"
-  )
-)
-```
-
-__Online supplement only__
-
-Byron C.
-Jaeger, PhD,^1^ Adam P.
-Bress, PharmD, MS,^2,3^ Joshua D.
-Bundy, PhD,^4^ Alfred K.
-Cheung, MD,^5,6^ William C.
-Cushman, MD,^7^ Paul E.
-Drawz, MD,^8^ Lawrence J.
-Fine, MD,^9^ Karen C.
-Johnson, MD,^7^ Cora E.
-Lewis, MD,^10^ Suzanne Oparil, MD,^11^ Michael V.
-Rocco, MD,^12^ Stephen R.
-Rapp, PhD, ^13^ Joni K.
-Snyder, RN,BSN,^9^ Mark A.
-Supiano, MD,^14^ Paul K.
-Whelton, MD,^4^ Jeff D.
-Williamson, MD, MHS,^15^ Jackson T.
-Wright, Jr, MD, PhD,^16^ David M.
-Reboussin, PhD,^1^ and Nicholas M.
-Pajewski, PhD^1^
-
-^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC.
-^2^Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT.
-^3^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT.
-^4^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
-^5^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT.
-^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT.
-^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis.
-^8^Division of Renal Diseases & Hypertension, University of Minnesota, MN.
-^9^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD.
-^10^Department of Epidemiology, University of Alabama at Birmingham.
-^11^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham.
-^12^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^13^Departments of Psychiatry and Behavioral Medicine & Social Sciences and Health Policy, Wakfe Forest School of Medicine, Winston-Salem, NC.
-^14^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
-^15^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^16^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
-\
-
-```{r}
-block_section(
-  prop_section(
-    page_size = page_size(orient = "portrait"),
-    type = "continuous",
-    page_margins = margins_normal
-  )
-)
-```
-
-__eTable 1__: Characteristics of study participants.`r single_par`
-
-```{r}
-
-header_labels <- tbl_characteristics |> 
-  show_header_names(quiet = TRUE) |> 
-  mutate(
-    label = str_remove_all(label, fixed("*")),
-    label = if_else(column == 'stat_0_1', 
-                    paste0("Overall, ", label),
-                    label),
-    label = str_replace(label, ', ', '\n')
-  ) |> 
-  deframe()
-
-tbl_characteristics |> 
-  modify_footnote(update = everything() ~ NA) |> 
-  modify_spanning_header(update = everything() ~ NA) |> 
-  as_flex_table() |> 
-  set_header_labels(
-    label = "Characteristic",
-    stat_0_1 = "Overall\nN = 9,361",
-    stat_1_2 = "Standard\nN = 4,683",
-    stat_2_2 = "Intensive\nN = 4,678",
-    p.value_2 = "p-value",
-    stat_1_3 = "No\nN = 5,952",
-    stat_2_3 = "Yes\nN = 3,409",
-    p.value_3 = "p-value"
-  ) |> 
-  add_header_row(
-    values = c("Characteristic",
-               "Overall\nN = 9,361",
-               "Treatment",
-               header_labels['p.value_2'],
-               "Included in EHR ancillary study", 
-               header_labels['p.value_3']),
-    colwidths = c(1, 1, 2, 1, 2, 1)
-  ) |> 
-  theme_box() |> 
-  merge_v(part = 'header') |> 
-  align(align = 'center', part = 'all') |> 
-  align(j = 1, align = 'left', part = 'all') |> 
-  footnote(
-    i = 1, 
-    j = 1, 
-    ref_symbols = '',
-    value = as_paragraph(
-      "Table values are percentages unless otherwise specified"
-    )
-  ) |> 
-  footnote(
-    i = 1, 
-    j = ~p.value_2 + p.value_3,
-    part = 'header',
-    ref_symbols = 'a',
-    value = as_paragraph(
-      "P-values computed using Wilcoxon rank sum test for age in years, chi-square test for age \u226575 years, and chi-square test for all other variables"
-    )
-  ) |> 
-  footnote(
-    i = ~ label == "Chronic Kidney Disease",
-    j = 1, 
-    ref_symbols = 'b',
-    value = ckd_defn,
-    part = 'body'
-  ) |> 
-  footnote(
-    i = 1,
-    j = 1,
-    ref_symbols = '',
-    value = abbrvs_write(abbrvs[c('SD', 'EHR', 'IQR', 'MoCA', 'FI')]),
-    part = 'body'
-  ) |>
-  fontsize(size = 12, part = 'all') |> 
-  font(fontname = 'Times New Roman', part = 'all') |> 
-  width(j = ~ stat_0_1 + stat_1_2 + stat_2_2 + stat_1_3 + stat_2_3,
-        width = 1) |> 
-  width(j = 1, width = 1.75)
-
-```
-
-\newpage
-
-__eTable 2__: All-cause mortality by treatment group and subgroup `r single_par`
-
-```{r}
-
-tbl_inf <- long_inf |> 
-  mutate(mort = .("{mort_est}\n({mort_lwr}, {mort_upr})")) |> 
-  select(variable,
-         group, 
-         trial_phase,
-         mort) |> 
-  pivot_wider(names_from = trial_phase,
-              values_from = mort,
-              names_prefix = '.h_')
-
-tbl_data <- long_ds |>
-  mutate(
-    across(starts_with("n_"), as.integer),
-    .n = .("{n_acm} / {n_obs}"),
-    .i = .("{inc_mort_est}\n({inc_mort_lwr}, {inc_mort_upr})")
-  ) |>
-  select(-starts_with("inc_"),
-         -starts_with("n_")) |> 
-  pivot_wider(names_from = c(treatment, trial_phase),
-              values_from = c(.n, .i)) |> 
-  left_join(tbl_inf) |> 
-  mutate(
-    variable = replace(variable, group == variable, NA_character_),
-    variable = recode(variable, !!!recoders$variable),
-    group = recode(group, !!!recoders$group)
-  )
-
-tbl_object <- tbl_data |> 
-  select(variable,
-         group,
-         .n_Standard_trial,
-         .n_Intensive_trial,
-         .i_Standard_trial,
-         .i_Intensive_trial,
-         .h_trial,
-         .n_Standard_cohort,
-         .n_Intensive_cohort,
-         .i_Standard_cohort,
-         .i_Intensive_cohort,
-         .h_cohort) |> 
-  as_grouped_data(groups = 'variable') |> 
-  drop_na_rows() |> 
-  as_flextable(hide_grouplabel = TRUE) |> 
-  add_header_row(
-    values = c(" ", 
-               "N Events / N Total", 
-               "Incidence\n(95% CI)", 
-               "Hazard Ratio\n(95% CI)",
-               "N Events / N Total", 
-               "Incidence\n(95% CI)",
-               "Hazard Ratio\n(95% CI)"),
-    colwidths = c(1, 2, 2, 1, 2, 2, 1)
-  ) |> 
-  add_header_row(
-    values = c(" ", 
-               "Trial Follow-up Through Close-out Visits",
-               "Post-trial Follow-up"),
-    colwidths = c(1, 5, 5)
-  ) |> 
-  set_header_labels(
-    group = " ",
-    .n_Standard_trial   = "Standard",
-    .n_Intensive_trial  = "Intensive",
-    .i_Standard_trial   = "Standard",
-    .i_Intensive_trial  = "Intensive",
-    .h_trial            = "Hazard Ratio\n(95% CI)",
-    .n_Standard_cohort  = "Standard",
-    .n_Intensive_cohort = "Intensive",
-    .i_Standard_cohort  = "Standard",
-    .i_Intensive_cohort = "Intensive",
-    .h_cohort           = "Hazard Ratio\n(95% CI)"
-  ) |> 
-  merge_v(part = 'header') |>
-  theme_box() |> 
-  padding(j = 1, i = ~ !is.na(group), padding.left = 15, part = 'body') |> 
-  bg(i = ~ is.na(group), bg = 'grey80') |> 
-  italic(i = ~ is.na(group), italic = TRUE) |> 
-  width(width = 0.95) |> 
-  width(j=1, width = 1.5) |> 
-  align(align = 'center', part = 'all') |> 
-  align(j = 1, align = 'left', part = 'all') |> 
-  fontsize(size = 12, part = 'all') |> 
-  font(fontname = 'Times New Roman', part = 'all') |> 
-  footnote(i = ~ variable == 'Chronic Kidney Disease', 
-           j = 1,
-           part = 'body',
-           ref_symbols = 'a',
-           value = ckd_defn) |> 
-  footnote(i = 1, 
-           j = 1, 
-           ref_symbols = '',
-           value = abbrvs_write(abbrvs[c('CI', 'FI')]))
-
-tbl_object
-
-```
-
-\newpage
-__eTable 3__: Cardiovascular and non-cardiovascular mortality by treatment group and subgroup `r single_par`
-
-```{r}
-
-tbl_inf <- long_inf |> 
-  mutate(cvd = .("{cvd_est}\n({cvd_lwr}, {cvd_upr})"),
-         non = .("{other_est}\n({other_lwr}, {other_upr})")) |> 
-  select(variable,
-         group, 
-         trial_phase,
-         cvd,
-         non) |> 
-  pivot_wider(names_from = trial_phase,
-              values_from = c(cvd, non),
-              names_glue = ".h_{.value}_{trial_phase}")
-
-tbl_data <- long_ds |>
-  mutate(
-    across(starts_with("n_"), as.integer),
-    .n = .("{n_cvd} / {n_other} / {n_obs}")
-  ) |>
-  select(-starts_with("inc_"),
-         -starts_with("n_")) |> 
-  pivot_wider(names_from = c(treatment, trial_phase),
-              values_from = c(.n),
-              names_prefix = '.n_') |> 
-  left_join(tbl_inf) |> 
-  mutate(
-    variable = replace(variable, group == variable, NA_character_),
-    variable = recode(variable, !!!recoders$variable),
-    group = recode(group, !!!recoders$group)
-  )
-
-tbl_object <- tbl_data |> 
-  select(variable,
-         group,
-         .n_Standard_trial,
-         .n_Intensive_trial,
-         .h_cvd_trial,
-         .h_non_trial,
-         .n_Standard_cohort,
-         .n_Intensive_cohort,
-         .h_cvd_cohort,
-         .h_non_cohort) |> 
-  as_grouped_data(groups = 'variable') |> 
-  drop_na_rows() |> 
-  as_flextable(hide_grouplabel = TRUE) |> 
-  add_header_row(
-    values = c(" ", 
-               "N\nCVD / non-CVD / Total", 
-               "Hazard Ratio\n(95% CI)",
-               "N\nCVD / non-CVD / Total", 
-               "Hazard Ratio\n(95% CI)"),
-    colwidths = c(1, 2, 2, 2, 2)
-  ) |> 
-  add_header_row(
-    values = c(" ", 
-               "Trial Follow-up Through Close-out Visits",
-               "Post-trial Follow-up"),
-    colwidths = c(1, 4, 4)
-  ) |> 
-  set_header_labels(
-    group = " ",
-    .n_Standard_trial= "Standard",
-    .n_Intensive_trial= "Intensive",
-    .h_cvd_trial = "CVD",
-    .h_non_trial = "Non-CVD",
-    .n_Standard_cohort = "Standard",
-    .n_Intensive_cohort = "Intensive",
-    .h_cvd_cohort = "CVD",
-    .h_non_cohort = "Non-CVD"
-  ) |> 
-  merge_v(part = 'header') |>
-  theme_box() |> 
-  padding(j = 1, i = ~ !is.na(group), padding.left = 15, part = 'body') |> 
-  bg(i = ~ is.na(group), bg = 'grey80') |> 
-  italic(i = ~ is.na(group), italic = TRUE) |> 
-  width(width = 0.95) |> 
-  width(j=1, width = 1.5) |> 
-  align(align = 'center', part = 'all') |> 
-  align(j = 1, align = 'left', part = 'all') |> 
-  fontsize(size = 12, part = 'all') |> 
-  font(fontname = 'Times New Roman', part = 'all') |> 
-  footnote(i = ~ variable == 'Chronic Kidney Disease', 
-           j = 1,
-           part = 'body',
-           ref_symbols = 'a',
-           value = ckd_defn) |> 
-  footnote(i = 1, 
-           j = 1, 
-           ref_symbols = '',
-           value = abbrvs_write(abbrvs[c('CI', 'CVD', 'FI')]))
-
-tbl_object
-
-```
-
-\newpage
-
-__eFigure 1__: Mean difference in systolic blood pressure over time. Shaded areas indicate a 95% confidence interval for the mean. `r single_par`
-
-```{r}
-
-bp_viz$diff + 
-  theme(legend.key.size = unit(1, 'cm'))
-
-```
-
-```{r}
-block_section(
-  prop_section(
-    page_size = page_size(orient = "landscape"),
-    page_margins = page_mar(bottom = 1/8, top = 1/8, 
-                            right = 1/4, left = 1/4),
-    type = "continuous"
-  )
-)
-```
-
-# REFERENCES

### 2022-01-20 doc/manuscript.Rmd
first full draft!

@@
 
+bp_viz_5yr <- bp_viz |> 
+  getElement('levels') |> 
+  getElement('data') |> 
+  filter(source == 'EHR', 
+         time_months == 60)
+
+bp_viz_5yr <- split(bp_viz_5yr, bp_viz_5yr$treatment)
+
+bp_viz_10yr <- bp_viz |> 
+  getElement('levels') |> 
+  getElement('data') |> 
+  filter(source=='EHR',
+         time_months == 120)
+
+bp_viz_10yr <- split(bp_viz_10yr, bp_viz_10yr$treatment)
+
+# a temporary function for inline reporting of BP values
+f <- function(d) with(d, .("{Estimate} ({Lower}, {Upper})"))
+
+bp_diff_5yr <- bp_viz |> 
+  getElement('diff') |> 
+  getElement('data') |> 
+  filter(time_months == 60)
+
+t_atten_sbp <- bp_viz |> 
+  getElement('diff') |> 
+  getElement('data') |> 
+  filter(source == 'EHR',
+         find_signchange_lgl(Lower)) |> 
+  pull(time_months) |> 
+  divide_by(12)
+
 ```
@@
 __IMPORTANCE__.
-Both the Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) trial have shown benefits of intensive blood pressure control on cardiovascular morbidity and mortality.
-However, as both trials were ended after slightly more than 3 years of follow-up, the effect of intensive treatment on longer term mortality is unknown.
+The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a systolic blood pressure (SBP) goal of <120 mmHg, reduced the risk of cardiovascular morbidity and mortality.
+However, the post-trial association of intensive treatment with mortality is unknown.
 
-__OBJECTIVE__. To evaluate the legacy effect of intensive hypertension treatment on longer term all-cause and cardiovascular mortality.
+__OBJECTIVE__.
+To evaluate the association of receiving intensive treatment for approximately three years during the SPRINT trial with all-cause and cardiovascular mortality up to ten years post-randomization.
 
@@
 SPRINT, a randomized clinical trial of 9361 patients aged 50 years or older with hypertension and increased cardiovascular risk, but without diabetes or history of stroke.
-Randomization began on November 8, 2010, the trial intervention was stopped early on August 20, 2015, and trial follow-up visits occurred through July 2016.
+Randomization began on November 8, 2010, the trial intervention ended on August 20, 2015, and trial follow-up visits occurred through July 2016.
 
-__INTERVENTIONS__. Randomization to a systolic blood pressure goal of either less than 120 mm Hg (intensive treatment, n=4678) versus less than 140 mm Hg (standard treatment, N=4683).
+__INTERVENTIONS__. Randomization to SBP goal of either <120 mmHg (intensive treatment, N=4678) versus <140 mmHg (standard treatment, N=4683).
 
 __MAIN OUTCOMES AND MEASURES__.
-Cardiovascular and all-cause mortality assessed via the US National Death Index, beginning in 2016 through December 31st, 2020.
-Outpatient blood pressures measured in routine clinical practice after the trial were examined in a subset of trial participants (N=3644).
+All-cause and cardiovascular mortality assessed via the US National Death Index, beginning in 2016 through December 31st, 2020.
+In a subset of 3644 trial participants, outpatient SBP levels measured in routine clinical practice after the trial were examined.
 
 __RESULTS__.
-Among 9361 randomized participants (mean age, 67.9 years; 3332 women [35.6%]), the median intervention period was 3.34 years.
-Over a median follow-up of `r ap_b_o$fup_est` years, intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial.
-However, there was no indication of benefit during post-trial follow-up for either cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
-Results were similar for subgroups based on baseline age, cognitive function, and frailty status.
-Analyses of outpatient blood pressures indicated a steady decline in the mean between group difference following the trial, largely driven by increases in mean systolic blood pressure in participants randomized to intensive treatment, increasing from a mean of 131.7 mm Hg (95% CI 130.7 to 132.6 mm Hg) at 4.5 years of follow-up to 136.8 mm Hg (95% CI 135.7 to 137.8 mm Hg) at 9 years of follow-up.
+Among 9361 randomized participants (mean [standard deviation] age `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years; `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% women), the median intervention phase was 3.3 years.
+Over a median follow-up of `r ap_b_o$fup_est` years, intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% confidence interval [CI] {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial.
+However, there was no evidence of association during post-trial follow-up for cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
+The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization.
 
 __CONCLUSIONS AND RELEVANCE__.
-The observed benefit of intensive treatment on cardiovascular and all-cause mortality was largely attenuated during post-trial observational follow-up.
-Given increasing blood pressures in participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension.
+The beneficial effect of intensive treatment on cardiovascular and all-cause mortality was attenuated during post-trial observational follow-up.
+Given increasing SBP levels in participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension.
 
@@
 
-Hypertension is the leading modifiable risk factor for cardiovascular disease (CVD).[@gdbJACC2020] A meta analysis of pooled individual-level data from randomized trials has shown that pharmacological blood pressure (BP) reduction reduces the risk of major cardiovascular events across the spectrum of initial BP, with larger absolute risk reduction for older adults.[@rahimi_age_stratified_2021] The Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) each showed that intensive systolic BP (SBP) control, defined by a SBP target < 120 mm Hg in SPRINT and < 130 mm Hg in STEP, reduced the risk for cardiovascular morbidity and mortality.
-[@sprint_research_group_randomized_2015; @zhang_trial_2021] However,
-only SPRINT showed a benefit of intensive treatment on all-cause mortality.
-Another contrasting result was that intensive treatment in SPRINT, but not STEP, increased the risk of acute kidney injury (AKI).
-While this may be attributable to the much lower prevalence of chronic kidney disease in STEP (2.3% vs 28.3%), the consistent association of AKI with increased mortality risk[@chawla2014aki; @rocco2018aki] raises questions about the longer term effect of intensive treatment on CVD and all-cause mortality.
+<!-- NOTE: citations should go before the end punctuation. -->
 
-Because both trials were stopped early after a median follow-up of 3 years, the objective of the current study was to estimate the longer term, legacy effect of randomization to intensive treatment on cardiovascular and all-cause mortality risk for participants in SPRINT.
-A secondary objective, in a subset of trial participants, was to examine attained SBP following the conclusion of trial follow-up.
-To accomplish these objectives, we linked SPRINT participants to the National Death Index (NDI) from 2016 to 2020, and extracted longitudinal outpatient measurements of SBP available in the electronic health record (EHR) over the same time period.
+<!-- Rmd automatically puts the mark after the punctuation when the csl file indicates it should be there (which it does for jama.csl). -->
+
+Hypertension is the leading modifiable risk factor for cardiovascular disease (CVD) [@gdbJACC2020].
+Meta analyses of randomized trials have shown that pharmacological blood pressure (BP) lowering reduces the risk of major cardiovascular events across the spectrum of initial BP, with larger absolute risk reduction for older adults and lowest CVD risk when systolic BP (SBP) is 120 to 124 mm Hg [@rahimi_age_stratified_2021; @bundy_systolic_2017].
+The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive SBP control, defined by a SBP target < 120 mm Hg, reduced the risk for cardiovascular morbidity and mortality [@sprint_research_group_randomized_2015].
+Because a clear benefit was evident for intensive SBP lowering after a median follow-up of 3.3 years, SPRINT was stopped early.
+However, the association of receiving intensive treatment during the SPRINT trial with cardiovascular and all-cause mortality risk after the trial has not been evaluated.
+
+The objective of the current study was to evaluate the association of intensive SBP control during the SPRINT trial with post-trial all-cause and cardiovascular mortality.
+A secondary objective was to examine SBP following the conclusion of the SPRINT trial follow-up.
+To accomplish these objectives, we linked SPRINT participants to the National Death Index (NDI) from 2016 to 2020, and extracted longitudinal outpatient measurements of SBP available in the electronic health record (EHR) for a subset of trial participants over the same time period.
 
@@
 
-__Trial Design__: The design and methods of SPRINT have been published previously.[@ambrosius_design_2014; @sprint_research_group_randomized_2015] Briefly, it was a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk of CVD.
+__Trial Design__: The design and methods of SPRINT have been published previously [@ambrosius_design_2014; @sprint_research_group_randomized_2015].
+Briefly, it was a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk of CVD.
 Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed.
@@
 __Baseline Study Measurements__: Sociodemographic data were collected at baseline, with race or ethnicity information collected via self-report.
-The estimated glomerular filtration rate (eGFR) was calculated by the race-free 2021 CKD-EPI creatinine equation.[@inker_new_2021] Cognitive function was assessed using the Montreal Cognitive Assessment (MoCA) [@nasreddine_montreal_2005].
-Lower cognitive function was defined as scoring at or below the estimated age and education-specific normative 10th percentile from the Irish Longitudinal Study of Aging,[@kenny_normative_2013] after adding 3 points to the scores of non-White participants.[@sachsMoCA2021] We defined frailty status at baseline using a 36-item Frailty Index (FI) based upon the model of deficit accumulation.[@pajewski_characterizing_2016] The FI is calculated as the sum of the score for each deficit divided by the total number of nonmissing items.
+The estimated glomerular filtration rate (eGFR) was calculated by the race-free 2021 CKD-EPI creatinine equation [@inker_new_2021].
+Cognitive function was assessed using the Montreal Cognitive Assessment (MoCA) [@nasreddine_montreal_2005].
+Lower cognitive function was defined as scoring at or below the estimated age and education-specific normative 10th percentile from the Irish Longitudinal Study of Aging [@kenny_normative_2013], after adding three points to the scores of non-White participants [@sachsMoCA2021].
+We defined frailty status at baseline using a 36-item Frailty Index (FI) based upon the model of deficit accumulation [@pajewski_characterizing_2016].
+The FI is calculated as the sum of the score for each deficit divided by the total number of nonmissing items.
 We categorized frailty status as fit (FI ≤ 0.10), less fit (0.10 < FI ≤ 0.21), or frail (FI > 0.21).
 
-__National Death Index Linkage__: Outcomes of interest included all-cause and CVD mortality.
-Methods of ascertainment and adjudication through the course of trial follow-up have been previously described.[@sprint_research_group_randomized_2015; @sprintFinal] Subsequently, mortality was ascertained through a US National Death Index (NDI) search.
-Possible matches were identified according to NDI guidelines.[@ndi_guide] Deaths were treated as confirmed if they were a Class 1 match, or a Class 2, 3, or 4 match with a probabilistic score above cutoffs recommended by the NDI.[@ndi_guide] NDI follow-up began in 2016 and ended on the date of death or December 31, 2020.
+__National Death Index Linkage__: Outcomes of interest included all-cause and cardiovascular mortality.
+Methods of ascertainment and adjudication through the course of trial follow-up have been previously described [@sprint_research_group_randomized_2015; @sprintFinal].
+Subsequently, mortality was ascertained through a US National Death Index (NDI) search.
+Possible matches were identified according to NDI guidelines [@ndi_guide].
+Deaths were treated as confirmed if they were a Class 1 match, or a Class 2, 3, or 4 match with a probabilistic score above cutoffs recommended by the NDI [@ndi_guide].
+NDI follow-up began in 2016 and ended on the date of death or December 31, 2020.
 Deaths ascertained in 2020 were based on the NDI preliminary data release.
 CVD mortality for NDI-based follow-up used the NDI Plus System, which automatically identifies underlying causes of death from death certificates, including conversion to ICD-10 codes.
-we defined CVD mortality as any death containing the ICD-10 codes of I00 to I99.[@regards_ndi]
+we defined CVD mortality as any death containing the ICD-10 codes of I00 to I99 [@regards_ndi].
 
 __EHR Ancillary Study__: We examined the trajectory of SBP following the conclusion of the trial using outpatient SBPs extracted from the EHR.
-Methods for the linkage of participants to their medical record number and the extraction of vital sign data have been previously described.[@drawz_concordance_2020] Because encounter type information was inconsistently available (i.e.
+Methods for the linkage of participants to their medical record number and the extraction of vital sign data have been previously described [@drawz_concordance_2020].
+Because encounter type information was inconsistently available (i.e.
 outpatient, inpatient, observation, etc.), we defined a BP measurement as outpatient if there were was not a BP measurement on the preceding or following day, and if there were 2 or less BP measurements on a particular day.
@@
 the proportional hazards assumption was likely to be invalid), we modeled treatment group differences as a function of time using two approaches.
-The first approach used Cox proportional hazards regression but split follow-up time into non-overlapping trial and cohort phases, estimating separate treatment group differences for each phase.[@zhang_time_varying_2018].
-The second approach estimated the effect of treatment group as a continuous time-dependent effect.[@martinussen2007dynamic; @therneau2017using] All analyses accounted for correlation within study sites,[@lin_robust_1989] and analyses of cardiovascular mortality accounted for the competing risk of non-cardiovascular mortality.[@scheike_flexible_2008].
+The first approach split each participant's follow-up time into non-overlapping trial and cohort phases, and estimated regression coefficients for intensive treatment separately during each phase [@zhang_time_varying_2018].
+The second approach estimated a regression coefficient for intensive treatment as a continuous function of time since randomization [@martinussen2007dynamic; @therneau2017using].
+All analyses accounted for correlation within study sites [@lin_robust_1989], and analyses of cardiovascular mortality accounted for the competing risk of non-cardiovascular mortality [@scheike_flexible_2008].
 Mean between-group differences in outpatient SBP following the conclusion of trial follow-up were estimated using linear mixed models.
@@
 
-t_signif <- acm_overall_epochs$label[1] 
-t_atten <- acm_overall_epochs$label[2]
+t_signif_acm <- acm_overall_epochs$label[1] 
+t_atten_acm <- acm_overall_epochs$label[2]
 
@@
 The hazard ratio (HR) for all-cause mortality comparing intensive to standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial phase, and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during the cohort phase.
-The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 1__).
+The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif_acm` to `r t_atten_acm` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 1__).
 In subgroups based on age, sex, race, CKD, cognitive function, and frailty, there was no evidence that intensive treatment during the trial phase had benefit for all-cause mortality during the cohort phase (__Figure 2__).
@@
 
-t_signif <- cvd_overall_epochs$label[1] 
-t_atten  <- cvd_overall_epochs$label[2]
+t_signif_cvd <- cvd_overall_epochs$label[1] 
+t_atten_cvd  <- cvd_overall_epochs$label[2]
 
@@
 The HR for CVD mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the cohort phase.
-The time-dependent effect of intensive versus standard treatment indicated a benefit for CVD mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort period (__Figure 3__).
+The time-dependent effect of intensive versus standard treatment indicated a benefit for CVD mortality from `r t_signif_cvd` to `r t_atten_cvd` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 3__).
 Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase had benefit for CVD mortality during the cohort phase (__Figure 4__).
-
-```{r}
-
-bp_viz_5yr <- bp_viz |> 
-  getElement('levels') |> 
-  getElement('data') |> 
-  filter(source == 'EHR', 
-         time_months == 60)
-
-bp_viz_5yr <- split(bp_viz_5yr, bp_viz_5yr$treatment)
-
-bp_viz_10yr <- bp_viz |> 
-  getElement('levels') |> 
-  getElement('data') |> 
-  filter(source=='EHR',
-         time_months == 120)
-
-bp_viz_10yr <- split(bp_viz_10yr, bp_viz_10yr$treatment)
-
-# a temporary function for inline reporting below.
-f <- function(d) with(d, .("{Estimate} ({Lower}, {Upper})"))
-
-bp_diff_5yr <- bp_viz |> 
-  getElement('diff') |> 
-  getElement('data') |> 
-  filter(time_months == 60)
-
-t_atten <- bp_viz |> 
-  getElement('diff') |> 
-  getElement('data') |> 
-  filter(source == 'EHR',
-         find_signchange_lgl(Lower)) |> 
-  pull(time_months) |> 
-  divide_by(12)
-
-```
 
@@
 For participants randomized to standard treatment, mean (95% CI) SBP was estimated to be `r f(bp_viz_5yr$Standard)` at 5 years and `r f(bp_viz_10yr$Standard)` post-randomization.
-The difference in mean SBP levels between participants randomized to intensive versus standard treatment, in mm Hg, was `r f(bp_diff_5yr)` at 5 years post-randomization, and was attenuated at `r t_atten` years post-randomization (__eFigure 1__).
+The difference in mean SBP levels between participants randomized to intensive versus standard treatment, in mm Hg, was `r f(bp_diff_5yr)` at 5 years post-randomization, and was attenuated at `r t_atten_sbp` years post-randomization (__eFigure 1__).
 
@@
 
-A striking aspect of our results is the quickly weakening level of BP control for participants randomized to intensive treatment.
-While an attenuation of the between-group BP delta subsequent to the trial was certainly expected, one hypothesis was that such an attenuation would be driven by participants randomized to standard treatment pursuing a lower BP goal, given the results of SPRINT and subsequent changes to hypertension guidelines.
-However, this is clearly not what occurred.
-While we do not have access to prescription records to know how participant medication regimes may have changed after the trial, these results likely show some contribution of clinician therapeutic inertia, which has been identified as a significant barrier to improving population level control of hypertension.
-Combined with evidence showing recent decreases in the population of prevalence of controlled hypertension in the US,[@muntner2020htn] our results highlight the sobering reality facing the hypertension community.
-Sustainability is a clear limiting factor, especially with trying to implement lower BP goals and interventions earlier in adulthood.
+The current study analyzed all-cause and CVD-mortality among SPRINT participants up to 10 years following randomization and found no indication of benefit for randomization to intensive treatment after the trial had ended.
+Time-varying estimates of the benefit of intensive treatment for all-cause mortality were attenuated at `r t_atten_acm` years while the benefit for CVD-mortality was attenuated at `r t_atten_cvd` years post-randomization.
+Findings from our ancillary study of BP in the EHR indicated that the difference in SBP between treatment groups diminished over time and was attenuated approximately `r t_atten_sbp` years after randomization.
+These results in combination with the main findings of SPRINT indicate that the beneficial effect of BP control among adults with hypertension can be diminished if BP control is not sustained over time.
+
+The Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) trial enrolled 8511 Chinese patients 60 to 80 years of age with hypertension and randomized patients to a SBP target of 110 to less than 130 mm Hg (intensive treatment) or a target of 130 to less than 150 mm Hg (standard treatment) [@zhang_trial_2021].
+Similar to SPRINT, STEP found a HR of 0.72 (95% CI 0.39, 1.32) with intensive versus standard treatment for CVD mortality after a median follow-up of 3.34 years, but did not find evidence of a benefit for all-cause mortality.
+In the current study of SPRINT participants, the protective effect of intensive treatment for all-cause mortality was attenuated several years before the protective effect for CVD.
+These results in combination with findings from the STEP trial suggest weaker evidence for reduced all-cause versus CVD mortality risk with intensive BP control.
+
+<!-- I commenting this sentence out because I think it may be too speculative for this Discussion.
+Nick, if you like the idea of including the commented sentence in the paper, can we set up a call with Paul Whelton to discuss?
+-->
+
+<!-- Another contrasting result was that intensive treatment in SPRINT, but not STEP, increased the risk of acute kidney injury (AKI).
+While this may be attributable to the much lower prevalence of chronic kidney disease in STEP (2.3% vs 28.3%), the association of AKI with increased mortality risk may have been a contributing factor to the longer term effect of randomization to intensive treatment on all-cause mortality among SPRINT participants [@chawla2014aki; @rocco2018aki].
+-->
+
+Previous studies have found rising BP levels among US adults during the time period of the current study.
+General population studies of adults living in the US with hypertension found that the prevalence of uncontrolled BP (SBP ≥ 140 mm Hg or DBP ≥ 90 mm Hg) increased from 2013 to 2017 [@muntner2020htn].
+In addition, an analysis of data from 464,585 adults enrolled in a Quest Diagnostics wellness program found SBP was 1 mm Hg and 3 mm Hg higher, depending on age group and sex, in April through December of 2020 versus their corresponding values throughout 2019 [@laffin_rise_2022].
+The current study shows that even for adults who have maintained intense SBP control for 3 years, increasing SBP levels can quickly diminish the protective effect.
+Combined with previous findings on rising BP levels among US adults, data from the current study emphasize the need for implementation of population- and community-level strategies to improve BP control in the US.
+
+In October 2020, the US Surgeon General published a call to action to control hypertension [@us_surgeon_2020].
+Evidenced-based strategies to improve BP control addressed in the call to action include implementing treatment protocols, using integrated care teams, providing clinicians feedback on their performance, and promoting shared patient-provider management with self-measured BP monitoring.
+During the SPRINT trial, participants received team-based care consistent with strategies outlined in the 2020 call to action.
+After the trial phase, when these protocols were no longer followed, the incidence of all-cause mortality approximately doubled in both treatment groups.
+These data emphasize the potential benefit of implementing the goals and strategies of the 2020 US Surgeon General's call to action.
+Future research should continue to evaluate strategies for obtaining consistent BP control in clinical settings to reduce the burden of CVD, which remains the leading cause of death for US adults.
 
 This study has several limitations.
-First, while we restricted analyses to high quality NDI matches, some small degree of misclassification in linking participants to the NDI is likely.
-Second, while several studies have shown reasonable performance of using NDI diagnosis codes for defining CVD mortality, it is clearly not as robust as the adjudication process used in the primary follow-up for the trial.
-Third, information about BP control after the trial was limited to routine outpatient BPs extracted from the EHR, which are well known to poorly reflect the standardized BP measurement process used during the trial.[@drawz_concordance_2020]
+First, while we restricted analyses to high quality NDI matches, misclassification in linking participants to the NDI is possible.
+Second, while several studies have shown reasonable performance of using NDI diagnosis codes for defining CVD mortality, it is not as robust as the adjudication process used in the primary follow-up for the trial.
+Third, information about SBP control after the trial was limited to routine outpatient SBP values extracted from the EHR, which are known to poorly reflect the standardized BP measurement process used during the SPRINT trial.[@drawz_concordance_2020]
+
+In conclusion, the mean SBP levels of SPRINT participants randomized to intensive treatment increased after the trial and there was no evidence of association between receiving intensive treatment during the trial phase and post-trial all-cause or CVD-mortality up to 10 years following randomization.
+Therefore, maintaining controlled SBP may be essential for long-term CVD risk management.
 
@@
 
-\newpage
+The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par` \newpage
+
 __Figure 2__: All-cause mortality hazard ratio for participants randomized to intensive versus standard treatment.
@@
 
-\newpage
+The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par` \newpage
+
 __Figure 3__: (A) Cumulative incidence of cardiovascular and non-cardiovascular mortality by treatment group.
@@
 
-\newpage
+The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par` \newpage
+
 __Figure 4__: Cardiovascular mortality hazard ratio for participants randomized to intensive versus standard treatment.
@@
 ```
+The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par` \newpage
 
-\newpage
 __Figure 5__: Mean systolic blood pressure over time by treatment group. Shaded areas indicate a 95% confidence interval for the mean. `r single_par`
@@
 ```
+The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par`
 
@@
     stat_2_2 = "Intensive\nN = 4,678",
-    p.value_2 = "p-value",
     stat_1_3 = "No\nN = 5,952",
@@
                "Treatment",
-               header_labels['p.value_2'],
                "Included in EHR ancillary study", 
                header_labels['p.value_3']),
-    colwidths = c(1, 1, 2, 1, 2, 1)
+    colwidths = c(1, 1, 2, 2, 1)
   ) |> 
@@
     value = as_paragraph(
-      "Table values are percentages unless otherwise specified"
+      "Table values are mean (standard deviation) or percentage."
     )
@@
     i = 1, 
-    j = ~p.value_2 + p.value_3,
+    j = ~p.value_3,
     part = 'header',
-    ref_symbols = 'a',
+    ref_symbols = 'a ',
     value = as_paragraph(
-      "P-values computed using Wilcoxon rank sum test for age in years, chi-square test for age \u226575 years, and chi-square test for all other variables"
+      "P-values computed using Wilcoxon rank sum test for continuous variables and chi-square test for categorical variables."
+    )
+  ) |> 
+  footnote(
+    i = 1, 
+    j = ~p.value_3,
+    part = 'header',
+    ref_symbols = 'b ',
+    value = as_paragraph(
+      "Hypothesis tests compare the characteristics of participants included versus excluded from the ancillary study of blood pressure levels using data from the electronic health record."
     )
@@
     j = 1, 
-    ref_symbols = 'b',
+    ref_symbols = 'c ',
     value = ckd_defn,
@@
     ref_symbols = '',
-    value = abbrvs_write(abbrvs[c('SD', 'EHR', 'IQR', 'MoCA', 'FI')]),
+    value = abbrvs_write(abbrvs[c('EHR', 'MoCA', 'FI')]),
     part = 'body'
@@
         width = 1) |> 
+  width(j = ~ p.value_3, width = 1.01) |> 
   width(j = 1, width = 1.75)

### 2022-01-20 R/ndi_tabulate_characteristics.R

@@ -99,8 +99,7 @@ ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
                    ckd_2021 = recoders$levels$ckd_2021[2],
                    moca = recoders$levels$moca[2]),
       missing = 'no'
-    ) |>
-    add_p()
+    )
 
   tbl_overall <- ndi_tbl_data |>
     relocate(sbp, treatment, ehr_ancillary, .after = race) |>

### 2022-01-20 _targets.R

@@ -139,7 +139,9 @@ list(
       mutate(variable = str_remove(variable, '^base_viz_eff_'))
   ),
 
-  tar_render(manuscript, "doc/manuscript.Rmd")
+  tar_render(manuscript, "doc/manuscript.Rmd"),
+
+  tar_render(readme, "README.Rmd")
 
 )
 

### 2022-01-24 doc/manuscript.Rmd
NMP feedback; first draft

@@
 rspec <- round_spec() |>
-  round_using_magnitude(digits = c(2, 1, 0),
-                        breaks = c(10, 100, Inf))
+  round_using_magnitude(digits = c(2, 1),
+                        breaks = c(2, Inf))
+
+# rspec <- round_spec() |>
+#   round_using_decimal(digits = 1)
 
@@
 
+n_total <- nrow(ndi_baseline)
+n_ehr <- sum(ndi_baseline$ehr_ancillary == 'Yes')
+
 fig <- left_join(base_viz_inc, base_viz_eff)
@@
 
+bp_diff_10yr <- bp_viz |> 
+  getElement('diff') |> 
+  getElement('data') |> 
+  filter(time_months == 120)
+
 t_atten_sbp <- bp_viz |> 
@@
 __IMPORTANCE__.
-The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a systolic blood pressure (SBP) goal of <120 mmHg, reduced the risk of cardiovascular morbidity and mortality.
-However, the post-trial association of intensive treatment with mortality is unknown.
+The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a systolic blood pressure (SBP) goal of <120 mmHg, reduced the risk of cardiovascular and all-cause mortality.
+However, the legacy effect of intensive treatment on mortality, defined as the persistence of benefit after the trial, is unknown.
 
 __OBJECTIVE__.
-To evaluate the association of receiving intensive treatment for approximately three years during the SPRINT trial with all-cause and cardiovascular mortality up to ten years post-randomization.
+To evaluate the legacy effect of receiving intensive treatment for approximately three years during the SPRINT trial with all-cause and cardiovascular mortality up to ten years post-randomization.
 
 __DESIGN, SETTING, AND PARTICIPANTS__.
-SPRINT, a randomized clinical trial of 9361 patients aged 50 years or older with hypertension and increased cardiovascular risk, but without diabetes or history of stroke.
+Randomized clinical trial of 9361 patients aged 50 years or older with hypertension and increased cardiovascular risk, but without diabetes or history of stroke.
 Randomization began on November 8, 2010, the trial intervention ended on August 20, 2015, and trial follow-up visits occurred through July 2016.
@@
 __MAIN OUTCOMES AND MEASURES__.
-All-cause and cardiovascular mortality assessed via the US National Death Index, beginning in 2016 through December 31st, 2020.
-In a subset of 3644 trial participants, outpatient SBP levels measured in routine clinical practice after the trial were examined.
+Trial follow-up supplemented with mortality assessed via the US National Death Index, beginning in 2016 through December 31st, 2020.
+In a subset of `r n_ehr` trial participants, outpatient SBP levels measured in routine clinical practice after the trial were examined.
 
 __RESULTS__.
-Among 9361 randomized participants (mean [standard deviation] age `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years; `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% women), the median intervention phase was 3.3 years.
+Among `r n_total` randomized participants (mean [standard deviation] age `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years; `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% women), the median intervention phase was 3.3 years.
 Over a median follow-up of `r ap_b_o$fup_est` years, intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% confidence interval [CI] {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial.
-However, there was no evidence of association during post-trial follow-up for cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
-The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization.
+However, there was no evidence of benefit during post-trial follow-up for cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
+The estimated mean (95% CI) outpatient SBP among participants randomized to intensive treatment increased post-trial, from `r f(bp_viz_5yr$Intensive)` at five years to `r f(bp_viz_10yr$Intensive)` ten years post-randomization.
 
@@
 Meta analyses of randomized trials have shown that pharmacological blood pressure (BP) lowering reduces the risk of major cardiovascular events across the spectrum of initial BP, with larger absolute risk reduction for older adults and lowest CVD risk when systolic BP (SBP) is 120 to 124 mm Hg [@rahimi_age_stratified_2021; @bundy_systolic_2017].
-The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive SBP control, defined by a SBP target < 120 mm Hg, reduced the risk for cardiovascular morbidity and mortality [@sprint_research_group_randomized_2015].
-Because a clear benefit was evident for intensive SBP lowering after a median follow-up of 3.3 years, SPRINT was stopped early.
-However, the association of receiving intensive treatment during the SPRINT trial with cardiovascular and all-cause mortality risk after the trial has not been evaluated.
+The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a SBP target < 120 mm Hg, reduced the risk for cardiovascular and all-cause mortality [@sprint_research_group_randomized_2015].
+Because a clear benefit was evident for intensive treatment after a median follow-up of 3.3 years, SPRINT was stopped early.
+However, the persistence of benefit after the trial (i.e., the legacy effect) of intensive treatment on cardiovascular and all-cause mortality has not been evaluated [@wander_legacy_2018].
 
-The objective of the current study was to evaluate the association of intensive SBP control during the SPRINT trial with post-trial all-cause and cardiovascular mortality.
-A secondary objective was to examine SBP following the conclusion of the SPRINT trial follow-up.
-To accomplish these objectives, we linked SPRINT participants to the National Death Index (NDI) from 2016 to 2020, and extracted longitudinal outpatient measurements of SBP available in the electronic health record (EHR) for a subset of trial participants over the same time period.
+The objective of the current study was to evaluate the legacy effect of intensive treatment during the SPRINT trial by analyzing post-trial all-cause and cardiovascular mortality.
+A secondary objective was to examine SBP following the conclusion of the trial follow-up.
+To accomplish these objectives, we linked participants to the National Death Index (NDI) from 2016 to 2020 and extracted longitudinal outpatient measurements of SBP available in the electronic health record (EHR) for a subset of trial participants.
 
@@
 
-__Trial Design__: The design and methods of SPRINT have been published previously [@ambrosius_design_2014; @sprint_research_group_randomized_2015].
+__Trial Design__: The design and methods of the SPRINT trial have been published previously [@ambrosius_design_2014; @sprint_research_group_randomized_2015].
 Briefly, it was a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk of CVD.
@@
 Individuals at 102 sites in the United States and Puerto Rico were randomized (1:1) to a SBP goal of less than 120 mm Hg (intensive treatment group, n = 4678) or a goal of less than 140 mm Hg (standard treatment group, n = 4683), using random permuted blocks with the randomization stratified by clinic site.
-The algorithms and formulary for the trial are listed in the published study protocol ( __citations needed__?
-).
 Trial enrollment began in November 2010 and ended in March 2013, with active follow-up through July 1, 2016.[@sprintFinal] The study was approved by the institutional review board at each participating site, and each participant provided written informed consent.
@@
 
-__EHR Ancillary Study__: We examined the trajectory of SBP following the conclusion of the trial using outpatient SBPs extracted from the EHR.
-Methods for the linkage of participants to their medical record number and the extraction of vital sign data have been previously described [@drawz_concordance_2020].
+__EHR Ancillary Study__: Methods for the linkage of participants to their medical record number and the extraction of vital sign data have been previously described [@drawz_concordance_2020].
+We identified 3074 participants with 3 or more outpatient and trial BP measurements.
+After excluding `r 3074 - n_ehr` participants without EHR data following the conclusion of trial follow-up, a total of `r n_ehr` patients were included for the ancillary analysis.
 Because encounter type information was inconsistently available (i.e.
@@
 All analyses accounted for correlation within study sites [@lin_robust_1989], and analyses of cardiovascular mortality accounted for the competing risk of non-cardiovascular mortality [@scheike_flexible_2008].
+
+We examined the trajectory of SBP following the conclusion of the trial using outpatient SBPs extracted from the EHR.
 Mean between-group differences in outpatient SBP following the conclusion of trial follow-up were estimated using linear mixed models.
@@
 ___Study Participants___.
-A total of `r nrow(ndi_baseline)` participants were randomized between November 2010 and March 2013.
+A total of `r n_total` participants were randomized between November 2010 and March 2013.
 The mean (standard deviation [SD]) age was `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years, with `r inline_text(tbl_characteristics, variable = 'age_cat', column = 'stat_0_1')`% of participants aged 75 years or older (__eTable 1__).
@@
 The mean (SD) SBP at baseline was `r inline_text(tbl_characteristics, variable = 'sbp', column = 'stat_0_1')` and `r inline_text(tbl_characteristics, variable = 'moca', column = 'stat_0_1')`% of participants MoCA scores were below the 10^th^ percentile.
-Participants included in the ancillary EHR study had greater age, lower SBP, less cognitive impairment, higher prevalence of CKD, and were more likely to be male or black than participants not included.
+Compared to participants not included in the ancillary EHR study, participants included were more likely to be male, were older, had lower SBP and higher scores on the MoCA, and had a higher prevalence of CKD.
 
@@
 The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif_acm` to `r t_atten_acm` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 1__).
-In subgroups based on age, sex, race, CKD, cognitive function, and frailty, there was no evidence that intensive treatment during the trial phase had benefit for all-cause mortality during the cohort phase (__Figure 2__).
+In subgroups based on age, sex, race, CKD, cognitive function, and frailty status, there was no evidence that intensive treatment during the trial phase produced benefit for all-cause mortality during the cohort phase of follow-up (__Figure 2__).
 
@@
 ___Cardiovascular Mortality___.
-A total of `r ap_b_i$n_cvd` and `r ap_b_s$n_cvd` CVD mortality events occurred among participants randomized to intensive and standard treatment, respectively (__eTable 3__).
+A total of `r ap_b_i$n_cvd` and `r ap_b_s$n_cvd` CVD deaths occurred among participants randomized to intensive and standard treatment, respectively (__eTable 3__).
 The HR for CVD mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the cohort phase.
 The time-dependent effect of intensive versus standard treatment indicated a benefit for CVD mortality from `r t_signif_cvd` to `r t_atten_cvd` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 3__).
-Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase had benefit for CVD mortality during the cohort phase (__Figure 4__).
+Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase produced benefit for CVD mortality during the cohort phase (__Figure 4__).
 
 ___Blood Pressure___.
+Amongst a subset of `r n_ehr` trial participants, the median number of outpatient BP measurements extracted from the EHR during the cohort phase of follow-up was X (Interquartile range Y to Z; NMP to add this result).
 The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization (__Figure 5__).
 For participants randomized to standard treatment, mean (95% CI) SBP was estimated to be `r f(bp_viz_5yr$Standard)` at 5 years and `r f(bp_viz_10yr$Standard)` post-randomization.
-The difference in mean SBP levels between participants randomized to intensive versus standard treatment, in mm Hg, was `r f(bp_diff_5yr)` at 5 years post-randomization, and was attenuated at `r t_atten_sbp` years post-randomization (__eFigure 1__).
+The between-group difference in mean SBP levels (intensive minus standard)  was `r f(bp_diff_5yr)` mm Hg at 5 years post-randomization and was reduced to `r f(bp_diff_5yr)` at 10 years post-randomization (__eFigure 1__).
 
@@
 
-The current study analyzed all-cause and CVD-mortality among SPRINT participants up to 10 years following randomization and found no indication of benefit for randomization to intensive treatment after the trial had ended.
+The current study analyzed all-cause and CVD-mortality among SPRINT participants up to 10 years following randomization, finding that the benefits associated with intensive treatment quickly attenuated after the trial ended.
 Time-varying estimates of the benefit of intensive treatment for all-cause mortality were attenuated at `r t_atten_acm` years while the benefit for CVD-mortality was attenuated at `r t_atten_cvd` years post-randomization.
-Findings from our ancillary study of BP in the EHR indicated that the difference in SBP between treatment groups diminished over time and was attenuated approximately `r t_atten_sbp` years after randomization.
-These results in combination with the main findings of SPRINT indicate that the beneficial effect of BP control among adults with hypertension can be diminished if BP control is not sustained over time.
+Findings from our ancillary study of outpatient SBP measured in routine clinical practice indicated that the difference in SBP between treatment groups diminished over time, with no detectable difference in SBP approximately `r t_atten_sbp` years after randomization.
+These results in combination with the primary findings of SPRINT indicate that the beneficial effect of intensive treatment among adults with hypertension can be diminished if BP control is not sustained.
 
 The Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) trial enrolled 8511 Chinese patients 60 to 80 years of age with hypertension and randomized patients to a SBP target of 110 to less than 130 mm Hg (intensive treatment) or a target of 130 to less than 150 mm Hg (standard treatment) [@zhang_trial_2021].
-Similar to SPRINT, STEP found a HR of 0.72 (95% CI 0.39, 1.32) with intensive versus standard treatment for CVD mortality after a median follow-up of 3.34 years, but did not find evidence of a benefit for all-cause mortality.
+STEP found a HR of 0.72 (95% CI 0.39, 1.32) with intensive versus standard treatment for CVD mortality after a median follow-up of 3.34 years, but did not find evidence of a benefit for all-cause mortality.
 In the current study of SPRINT participants, the protective effect of intensive treatment for all-cause mortality was attenuated several years before the protective effect for CVD.
@@
 Second, while several studies have shown reasonable performance of using NDI diagnosis codes for defining CVD mortality, it is not as robust as the adjudication process used in the primary follow-up for the trial.
-Third, information about SBP control after the trial was limited to routine outpatient SBP values extracted from the EHR, which are known to poorly reflect the standardized BP measurement process used during the SPRINT trial.[@drawz_concordance_2020]
+Third, information about SBP control after the trial was limited to routine outpatient SBP values extracted from the EHR, which are known to poorly reflect the standardized BP measurement process used during the SPRINT trial.[@drawz_concordance_2020] While this prohibits definite conclusions about the absolute level of BP in both treatment groups, as well as pin-pointing when the between-group difference may have completely attenuated, the general observation of steadily increasing BPs for participants in the intensive treatment following the trial is likely still valid.
 
-In conclusion, the mean SBP levels of SPRINT participants randomized to intensive treatment increased after the trial and there was no evidence of association between receiving intensive treatment during the trial phase and post-trial all-cause or CVD-mortality up to 10 years following randomization.
-Therefore, maintaining controlled SBP may be essential for long-term CVD risk management.
+In conclusion, while intensive treatment produced beneficial effects on mortality during the trial, there was no evidence that this produced sustained benefits on cardiovascular and all-cause mortality subsequent to the conclusion of trial follow-up.
+Given steadily increasing mean SBP levels in participants randomized to intensive treatment after the trial, these results demonstrate that maintaining more intensive BP targets throughout adulthood will likely be essential for long-term CVD risk management.
+
+\newpage
+
+# REFERENCES
+
+<div id="refs"></div>
+
+\newpage
 
@@
 
-__Figure 1__: (A) Cumulative incidence of all-cause mortality by treatment group. (B) Time-dependent effect of randomization to intensive treatment for all-cause mortality. `r single_par`
+__Figure 1__: Incidence of All-Cause Mortality by Treatment Group. `r single_par`
 
@@
 
-The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par` \newpage
+(A) Cumulative incidence of all-cause mortality by treatment group. (B) Time-dependent effect of randomization to intensive treatment for all-cause mortality.`r single_par`
+
+The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par`
+
+\newpage
 
@@
 
-The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par` \newpage
+The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par`
 
-__Figure 3__: (A) Cumulative incidence of cardiovascular and non-cardiovascular mortality by treatment group.
-(B) Time-dependent effect of randomization to intensive treatment for cardiovascular mortality.
-`r single_par`
+\newpage
+
+__Figure 3__: Incidence of cardiovascular versus non-cardiovascular mortality by treatment group. `r single_par`
 
@@
 
-The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par` \newpage
+(A) Cumulative incidence of cardiovascular and non-cardiovascular mortality by treatment group.
+(B) Time-dependent effect of randomization to intensive treatment for cardiovascular mortality.`r single_par`
+
+The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par`
+
+\newpage
 
@@
 ```
-The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par` \newpage
+The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par`
 
-__Figure 5__: Mean systolic blood pressure over time by treatment group. Shaded areas indicate a 95% confidence interval for the mean. `r single_par`
+\newpage
+
+__Figure 5__: Mean systolic blood pressure over time by treatment group. `r single_par`
 
@@
 ```
+
+EHR denotes electronic health record. `r single_par`
+Shaded areas indicate a 95% confidence interval for the mean.`r single_par`
 The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par`
@@
     stat_2_2 = "Intensive\nN = 4,678",
-    stat_1_3 = "No\nN = 5,952",
-    stat_2_3 = "Yes\nN = 3,409",
+    stat_1_3 = "No\nN = 6,417",
+    stat_2_3 = "Yes\nN = 2,944",
     p.value_3 = "p-value"
@@
     values = c(" ", 
-               "Trial Follow-up Through Close-out Visits",
-               "Post-trial Follow-up"),
+               "Trial Phase",
+               "Cohort Phase"),
     colwidths = c(1, 5, 5)
@@
     values = c(" ", 
-               "Trial Follow-up Through Close-out Visits",
-               "Post-trial Follow-up"),
+               "Trial Phase",
+               "Cohort Phase"),
     colwidths = c(1, 4, 4)
@@
 
-__eFigure 1__: Mean difference in systolic blood pressure over time. Shaded areas indicate a 95% confidence interval for the mean. `r single_par`
+__eFigure 1__: Mean difference in systolic blood pressure measured in the trial and as part of routine clinical practice over time. `r single_par`
 
@@
 ```
+
+Shaded areas indicate a 95% confidence interval for the mean. `r single_par`
+
+EHR denotes electronic health record.`r single_par`
 
@@
 ```
-
-# REFERENCES

### 2022-01-24 R/ndi_load.R

@@ -60,7 +60,9 @@ ndi_load <- function(sasinet_drive = "Z", fname, ...) {
   # indicate ancillary study membership for baseline data
   if(fname == 'longterm_death.csv'){
 
-    ancillary <- read_csv('ehr/SPRINT_EHR_SBP_LongTerm_PID_010622.csv')
+    ancillary <- read_csv('ehr/SPRINT_EHR_SBP_LongTerm_PID_011522.csv')
+
+    names(ancillary) <- tolower(names(ancillary))
 
     data_in$ehr_ancillary <- if_else(
       condition = data_in$pid %in% ancillary$pid,

### 2022-02-08 doc/manuscript.Rmd
incorporate feedback from BL

@@
 Johnson, MD,^7^ Cora E.
-Lewis, MD,^10^ Suzanne Oparil, MD,^11^ Michael V.
+Lewis, MD, MSPH^10^ Suzanne Oparil, MD,^11^ Michael V.
 Rocco, MD,^12^ Stephen R.
@@
 
+<!-- TODO: INKSCAPE A TIMELINE WITH TRIAL PHASES -->
+
 # ABSTRACT
@@
 
-__OBJECTIVE__.
-To evaluate the legacy effect of receiving intensive treatment for approximately three years during the SPRINT trial with all-cause and cardiovascular mortality up to ten years post-randomization.
+__OBJECTIVE__. To evaluate the legacy effect of receiving intensive treatment during the SPRINT trial with all-cause and cardiovascular mortality up to ten years post-randomization.
 
@@
 Randomized clinical trial of 9361 patients aged 50 years or older with hypertension and increased cardiovascular risk, but without diabetes or history of stroke.
-Randomization began on November 8, 2010, the trial intervention ended on August 20, 2015, and trial follow-up visits occurred through July 2016.
+Randomization began on November 8, 2010, the trial intervention ended on August 20, 2015, and observational trial follow-up visits occurred through July 2016.
 
@@
 __MAIN OUTCOMES AND MEASURES__.
-Trial follow-up supplemented with mortality assessed via the US National Death Index, beginning in 2016 through December 31st, 2020.
+Extended trial follow-up assessed mortality via the US National Death Index, beginning in 2016 through December 31st, 2020.
 In a subset of `r n_ehr` trial participants, outpatient SBP levels measured in routine clinical practice after the trial were examined.
@@
 Among `r n_total` randomized participants (mean [standard deviation] age `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years; `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% women), the median intervention phase was 3.3 years.
-Over a median follow-up of `r ap_b_o$fup_est` years, intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% confidence interval [CI] {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial.
-However, there was no evidence of benefit during post-trial follow-up for cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
-The estimated mean (95% CI) outpatient SBP among participants randomized to intensive treatment increased post-trial, from `r f(bp_viz_5yr$Intensive)` at five years to `r f(bp_viz_10yr$Intensive)` ten years post-randomization.
+Intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% confidence interval [CI] {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through trial close-out visits.
+However, there was no evidence of benefit during extended post-trial follow-up for cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
+The estimated mean (95% CI) outpatient SBP among participants randomized to intensive treatment increased from `r f(bp_viz_5yr$Intensive)` at five years to `r f(bp_viz_10yr$Intensive)` ten years post-randomization.
 
 __CONCLUSIONS AND RELEVANCE__.
-The beneficial effect of intensive treatment on cardiovascular and all-cause mortality was attenuated during post-trial observational follow-up.
+The beneficial effect of intensive treatment on cardiovascular and all-cause mortality was attenuated during extended post-trial follow-up.
 Given increasing SBP levels in participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension.
@@
 Hypertension is the leading modifiable risk factor for cardiovascular disease (CVD) [@gdbJACC2020].
-Meta analyses of randomized trials have shown that pharmacological blood pressure (BP) lowering reduces the risk of major cardiovascular events across the spectrum of initial BP, with larger absolute risk reduction for older adults and lowest CVD risk when systolic BP (SBP) is 120 to 124 mm Hg [@rahimi_age_stratified_2021; @bundy_systolic_2017].
 The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a SBP target < 120 mm Hg, reduced the risk for cardiovascular and all-cause mortality [@sprint_research_group_randomized_2015].
 Because a clear benefit was evident for intensive treatment after a median follow-up of 3.3 years, SPRINT was stopped early.
-However, the persistence of benefit after the trial (i.e., the legacy effect) of intensive treatment on cardiovascular and all-cause mortality has not been evaluated [@wander_legacy_2018].
+However, the long-term persistence of benefit after the trial (i.e., the legacy effect) of intensive treatment on cardiovascular and all-cause mortality has not been evaluated [@wander_legacy_2018].
 
-The objective of the current study was to evaluate the legacy effect of intensive treatment during the SPRINT trial by analyzing post-trial all-cause and cardiovascular mortality.
+The objective of the current study was to evaluate the long-term legacy effect of intensive treatment during the SPRINT trial by analyzing post-trial all-cause and cardiovascular mortality.
 A secondary objective was to examine SBP following the conclusion of the trial follow-up.
-To accomplish these objectives, we linked participants to the National Death Index (NDI) from 2016 to 2020 and extracted longitudinal outpatient measurements of SBP available in the electronic health record (EHR) for a subset of trial participants.
+To accomplish these objectives, we linked participants to the National Death Index (NDI) from 2016 through 2020 and extracted longitudinal outpatient measurements of SBP available in the electronic health record (EHR) for a subset of trial participants.
 
@@
 Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed.
-Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (CKD; defined by an estimated glomerular filtration rate of <60 mL/min/1.73 m^2^), or a Framingham Risk Score of 15% or greater or if they were aged 75 years or older.
+Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (CKD; defined by an estimated glomerular filtration rate of <60 mL/min/1.73 m^2^) [@levey_new_2009], or a Framingham Risk Score of 15% or greater or if they were aged 75 years or older.
 Individuals residing in a nursing home, persons with a diagnosis of dementia (based on medical record review), and those treated with medications primarily used for dementia therapy were excluded, as were persons with prevalent diabetes mellitus, history of stroke, proteinuria > 1 gram per day, or polycystic kidney disease.
 Individuals at 102 sites in the United States and Puerto Rico were randomized (1:1) to a SBP goal of less than 120 mm Hg (intensive treatment group, n = 4678) or a goal of less than 140 mm Hg (standard treatment group, n = 4683), using random permuted blocks with the randomization stratified by clinic site.
-Trial enrollment began in November 2010 and ended in March 2013, with active follow-up through July 1, 2016.[@sprintFinal] The study was approved by the institutional review board at each participating site, and each participant provided written informed consent.
+Randomization began on November 8, 2010 and ended in March 2013.
+The trial intervention ended on August 20, 2015, and observational trial follow-up visits occurred through July 2016.[@sprintFinal] The study was approved by the institutional review board at each participating site, and each participant provided written informed consent.
 The study is registered at ClinicalTrials.gov (NCT01206062).
@@
 -->
+
+Meta analyses of randomized trials have shown that pharmacological blood pressure (BP) lowering reduces the risk of major CVD events across the spectrum of initial BP [@rahimi_age_stratified_2021; @bundy_systolic_2017].
 
@@
   font(fontname = 'Times New Roman', part = 'all') |>
+  footnote(i = 2, j = 4,
+           part = 'header',
+           ref_symbols = 'a',
+           value = as_paragraph('Incidence is presented as the expected rate of events per 1,000 person-years.')) |> 
   footnote(i = ~ variable == 'Chronic Kidney Disease', 
@@
            part = 'body',
-           ref_symbols = 'a',
+           ref_symbols = 'b',
            value = ckd_defn) |> 

### 2022-02-10 _targets.R
name bug
Somewhere out there, manuscript.docx is open, and that means I cannot
write over it! So I just renamed manuscript.Rmd to make the markdown doc
have a different output name.

@@ -139,7 +139,7 @@ list(
       mutate(variable = str_remove(variable, '^base_viz_eff_'))
   ),
 
-  tar_render(manuscript, "doc/manuscript.Rmd"),
+  tar_render(manuscript, "doc/manuscript-rmd.Rmd"),
 
   tar_render(readme, "README.Rmd")
 

### 2022-02-15 R/add_annotations.R
updating viz for observational period

@@ -11,6 +11,7 @@ add_annotations <- function(fig,
                             xmax = NULL,
                             xmin = NULL,
                             ymult = 0.90,
+                            mid_label = 'Trial And\nObservational\nPhase',
                             x_cohort_phase = 6,
                             textsize = 4.5) {
 
@@ -53,11 +54,11 @@ add_annotations <- function(fig,
              x = 3.51,
              y = fig_ymax * ymult,
              size = textsize,
-             label = 'Trial And\nCohort\nPhase') +
+             label = mid_label) +
     annotate(geom = 'text',
              x = x_cohort_phase,
              y = fig_ymax * ymult,
              size = textsize,
-             label = 'Cohort Phase')
+             label = 'Observational Phase')
 
 }

### 2022-02-15 R/bp_long_viz.R

@@ -5,20 +5,23 @@
 #' @title
 #' @param date
 
-bp_long_viz <- function(date = "01-15-22") {
+bp_long_viz <- function(date_ehr = "02-08-22",
+                        date_trial = "01-15-22") {
 
   cols_bg <- viridis(3)
   cols_tx <- pal_lancet()(2)
   cols_diff <- c("maroon4","black")
 
-  .date <- str_remove_all(date, fixed('-'))
+  .date_ehr <- str_remove_all(date_ehr, fixed('-'))
+  .date_trial <- str_remove_all(date_trial, fixed('-'))
 
-  fname <- paste0("SPRINT_EHR_SBP_GLIMMIX_LongTerm_", .date, ".csv")
-  fname2 <- paste0("SPRINT_EHR_SBP_GLIMMIX_JAMA_IM_REDO_", .date, ".csv")
+
+  fname_ehr <- paste0("SPRINT_EHR_SBP_LongTerm_EHRonly_", .date_ehr, ".csv")
+  fname_trial <- paste0("SPRINT_EHR_SBP_GLIMMIX_JAMA_IM_REDO_", .date_trial, ".csv")
 
   # Blood pressure levels over time ----
 
-  bp <- read_csv(file.path('ehr', fname)) |>
+  bp_ehr <- read_csv(file.path('ehr', fname_ehr)) |>
     select(-Statement, -Alpha, -tValue, -DF, -StdErr) |>
     separate(Label, into = c('group', 'time_months'), sep = ';') |>
     mutate(time_months = as.numeric(time_months),
@@ -27,33 +30,19 @@ bp_long_viz <- function(date = "01-15-22") {
                           "INT" = "Intensive (EHR)",
                           "DIFF" = "Difference (EHR)"))
 
-  bp_trial <- read_csv(file.path('ehr', fname2)) |>
+  bp_trial <- read_csv(file.path('ehr', fname_trial)) |>
     select(-Statement, -Alpha, -tValue, -DF, -StdErr) |>
     separate(Label, into = c('group', 'source', 'time_months'), sep = ';') |>
-    mutate(time_months = as.numeric(time_months))
-
-  bp_levels <- filter(bp, group %in% c("Standard (EHR)",
-                                       "Intensive (EHR)") & time_months>54)
-
-  bp_levels_trial <- bp_trial |>
-    filter(group %in% c("STD", "INT") & source=="EHR") |>
-    select(-source) |>
-    mutate(group = recode(group,
-                          "STD" = "Standard (EHR)",
[…]

### 2022-02-15 R/ndi_base_viz_incidence.R

@@ -46,9 +46,13 @@ ndi_base_viz_incidence <- function(ndi_data_list) {
         tables.height = 1/8
       )
 
-      fig_kap$plot <- add_annotations(fig_kap$plot,
-                                      cols = cols_bg,
-                                      ymax = ymax)
+      fig_kap$plot <- add_annotations(
+        fig_kap$plot,
+        cols = cols_bg,
+        ymax = ymax,
+        mid_label = '\nTrial\nAnd\nObservational\nPhase',
+        x_cohort_phase = 7
+      )
 
 
       fit_cr <- cuminc(ftime   = ndi_data$acm_years,
@@ -116,11 +120,14 @@ ndi_base_viz_incidence <- function(ndi_data_list) {
         group_by(event) |>
         summarize(est = mean(est))
 
-      fig_cr <- add_annotations(fig_cr,
-                                cols = cols_bg,
-                                ymax = ymax,
-                                xmax = 13,
-                                x_cohort_phase = 6.25) +
+      fig_cr <- add_annotations(
+        fig_cr,
+        cols = cols_bg,
+        ymax = ymax,
+        xmax = 13,
+        mid_label = '\nTrial\nAnd\nObservational\nPhase',
+        x_cohort_phase = 7
+      ) +
         annotate(geom = 'text',
                  x = 11.5,
                  y = y_text$est[y_text$event == 'Non-CVD mortality'],

### 2022-02-15 R/ndi_base_viz_tv_effect.R

@@ -63,6 +63,8 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
                                      ymax = 2.5,
                                      xmax = 10,
                                      ymin = 0.35,
+                                     mid_label = '\nTrial\nAnd\nObservational\nPhase',
+                                     x_cohort_phase = 7,
                                      ymult = 0.75)
 
       fit_cvd <- comp.risk(
@@ -109,6 +111,8 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
                                      ymax = 2.5,
                                      xmax = 10,
                                      ymin = 0.15,
+                                     mid_label = '\nTrial\nAnd\nObservational\nPhase',
+                                     x_cohort_phase = 7,
                                      ymult = 0.75)
 
       tibble(eff_acm = list(fig_eff_acm),

### 2022-02-15 R/ndi_long_viz_forest.R

@@ -36,13 +36,13 @@ ndi_long_viz_mort <- function(long_inf, recoders) {
       trial_phase = factor(
         trial_phase,
         levels = c('trial', 'cohort'),
-        labels = c("Trial Phase", "Cohort Phase")
+        labels = c("Trial Phase", "Observational Phase")
       ),
       fill = 1 +
         as.numeric(variable == 'Overall') +
-        as.numeric(variable == 'Overall' & trial_phase == 'Cohort Phase') +
+        as.numeric(variable == 'Overall' & trial_phase == 'Observational Phase') +
         as.numeric(variable != group) +
-        as.numeric(variable != group & trial_phase == 'Cohort Phase'),
+        as.numeric(variable != group & trial_phase == 'Observational Phase'),
       fill = factor(fill),
       label = table_value(mort_est, rspec = rspec),
       label = replace(label, label == 'NA', "")
@@ -73,14 +73,14 @@ ndi_long_viz_mort <- function(long_inf, recoders) {
                       group = '',
                       label = 'HR',
                       fill = '1',
-                      trial_phase = c('Trial Phase', 'Cohort Phase'))
+                      trial_phase = c('Trial Phase', 'Observational Phase'))
 
 
   data_gg <- bind_rows(data_bind, data_gg) |>
     mutate(
       trial_phase = factor(
         trial_phase,
-        levels = c("Trial Phase", "Cohort Phase")
+        levels = c("Trial Phase", "Observational Phase")
       )
     )
 
@@ -161,13 +161,13 @@ ndi_long_viz_cvd <- function(long_inf, recoders) {
       trial_phase = factor(
         trial_phase,
         levels = c('trial', 'cohort'),
-        labels = c("Trial Phase", "Cohort Phase")
+        labels = c("Trial Phase", "Observational Phase")
       ),
       fill = 1 +
         as.numeric(variable == 'Overall') +
-        as.numeric(variable == 'Overall' & trial_phase == 'Cohort Phase') +
+        as.numeric(variable == 'Overall' & trial_phase == 'Observational Phase') +
         as.numeric(variable != group) +
-        as.numeric(variable != group & trial_phase == 'Cohort Phase'),
+        as.numeric(variable != group & trial_phase == 'Observational Phase'),
       fill = factor(fill),
       label = table_value(cvd_est, rspec = rspec),
       label = replace(label, label == 'NA', "")
[…]

### 2022-02-15 doc/manuscript-rmd.Rmd
name bug
Somewhere out there, manuscript.docx is open, and that means I cannot
write over it! So I just renamed manuscript.Rmd to make the markdown doc
have a different output name.
copyedits from JW/PD/AB

--- 2022-02-08 doc/manuscript.Rmd
@@
                       fig.width = 11, 
-                      fig.height = 6.25)
+                      fig.height = 6)
 
@@
 ^5^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT.
-^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT.
+^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake City, UT.
 ^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis.
@@
 ^12^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^13^Departments of Psychiatry and Behavioral Medicine & Social Sciences and Health Policy, Wakfe Forest School of Medicine, Winston-Salem, NC.
+^13^Departments of Psychiatry and Behavioral Medicine & Social Sciences and Health Policy, Wake Forest School of Medicine, Winston-Salem, NC.
 ^14^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
@@
 
-<!-- TODO: INKSCAPE A TIMELINE WITH TRIAL PHASES -->
-
 # ABSTRACT
@@
 __IMPORTANCE__.
-The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a systolic blood pressure (SBP) goal of <120 mmHg, reduced the risk of cardiovascular and all-cause mortality.
-However, the legacy effect of intensive treatment on mortality, defined as the persistence of benefit after the trial, is unknown.
+The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a systolic blood pressure (SBP) goal of <120mmHg, reduced the risk of cardiovascular and all-cause mortality.
+However, the legacy effect of intensive treatment on mortality, defined as the persistence of benefit after stopping intensive treatment, is unknown.
+
+<!-- The edit to defn of legacy effect was recommended by JW -->
 
@@
 
-__INTERVENTIONS__. Randomization to SBP goal of either <120 mmHg (intensive treatment, N=4678) versus <140 mmHg (standard treatment, N=4683).
+__INTERVENTIONS__. Randomization to SBP goal of <120mmHg (intensive, N=4678) versus <140mmHg (standard, N=4683).
 
@@
 __RESULTS__.
-Among `r n_total` randomized participants (mean [standard deviation] age `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years; `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% women), the median intervention phase was 3.3 years.
-Intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% confidence interval [CI] {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through trial close-out visits.
-However, there was no evidence of benefit during extended post-trial follow-up for cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
+Among `r n_total` randomized participants, the mean (standard deviation) age was `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years and `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% were women.
+Over the median intervention phase of 3.3 years, intensive treatment was beneficial for both cardiovascular mortality (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% confidence interval [CI] {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
+However, at the median total follow-up of `r ap_b_o$fup_est` years, there was no longer evidence of benefit for cardiovascular mortality (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
 The estimated mean (95% CI) outpatient SBP among participants randomized to intensive treatment increased from `r f(bp_viz_5yr$Intensive)` at five years to `r f(bp_viz_10yr$Intensive)` ten years post-randomization.
@@
 __CONCLUSIONS AND RELEVANCE__.
-The beneficial effect of intensive treatment on cardiovascular and all-cause mortality was attenuated during extended post-trial follow-up.
+The beneficial effect of intensive treatment on cardiovascular and all-cause mortality did not persist after the trial was stopped.
 Given increasing SBP levels in participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension.
@@
 Hypertension is the leading modifiable risk factor for cardiovascular disease (CVD) [@gdbJACC2020].
-The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a SBP target < 120 mm Hg, reduced the risk for cardiovascular and all-cause mortality [@sprint_research_group_randomized_2015].
+The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a SBP target < 120 mm Hg, reduced the risk for cardiovascular and all-cause mortality [@sprintFinal].
 Because a clear benefit was evident for intensive treatment after a median follow-up of 3.3 years, SPRINT was stopped early.
-However, the long-term persistence of benefit after the trial (i.e., the legacy effect) of intensive treatment on cardiovascular and all-cause mortality has not been evaluated [@wander_legacy_2018].
+However, the persistence of benefit after the trial was stopped (i.e., "legacy" effect) of intensive treatment on cardiovascular and all-cause mortality has not been evaluated [@wander_legacy_2018].
 
-The objective of the current study was to evaluate the long-term legacy effect of intensive treatment during the SPRINT trial by analyzing post-trial all-cause and cardiovascular mortality.
-A secondary objective was to examine SBP following the conclusion of the trial follow-up.
-To accomplish these objectives, we linked participants to the National Death Index (NDI) from 2016 through 2020 and extracted longitudinal outpatient measurements of SBP available in the electronic health record (EHR) for a subset of trial participants.
+The objective of the current study was to evaluate the effect of stopping intensive treatment as specified by the SPRINT trial protocol by analyzing post-trial all-cause and cardiovascular mortality.
+A secondary objective was to examine SBP following the discontinuation of the trial intervention.
+To accomplish these objectives, we linked participants to the National Death Index (NDI) from 2016 through 2020 and extracted longitudinal outpatient measurements of SBP from 20xx to 2020 available in the electronic health record (EHR) for a subset of trial participants.
 
@@
 
-__Trial Design__: The design and methods of the SPRINT trial have been published previously [@ambrosius_design_2014; @sprint_research_group_randomized_2015].
+__Trial Design__: The design and methods of SPRINT have been published previously [@ambrosius_design_2014; @sprintFinal].
 Briefly, it was a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk of CVD.
 Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed.
-Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (CKD; defined by an estimated glomerular filtration rate of <60 mL/min/1.73 m^2^) [@levey_new_2009], or a Framingham Risk Score of 15% or greater or if they were aged 75 years or older.
+Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (CKD), or a Framingham Risk Score of 15% or greater or if they were aged 75 years or older.
 Individuals residing in a nursing home, persons with a diagnosis of dementia (based on medical record review), and those treated with medications primarily used for dementia therapy were excluded, as were persons with prevalent diabetes mellitus, history of stroke, proteinuria > 1 gram per day, or polycystic kidney disease.
@@
 Randomization began on November 8, 2010 and ended in March 2013.
-The trial intervention ended on August 20, 2015, and observational trial follow-up visits occurred through July 2016.[@sprintFinal] The study was approved by the institutional review board at each participating site, and each participant provided written informed consent.
+On August 20, 2015, the SPRINT trial prematurely concluded due to the clear benefit of intensive treatment.
+As trial follow-up visits and provision of antihypertensive medication continued through July 2016 [@sprintFinal], the current study considers a trial phase spanning from November 8, 2010 through July 2016 and an observational phase spanning from August 2017 through December, 2020 (__eFigure 1__).
+The study was approved by the institutional review board at each participating site, and each participant provided written informed consent.
 The study is registered at ClinicalTrials.gov (NCT01206062).
@@
 __National Death Index Linkage__: Outcomes of interest included all-cause and cardiovascular mortality.
-Methods of ascertainment and adjudication through the course of trial follow-up have been previously described [@sprint_research_group_randomized_2015; @sprintFinal].
-Subsequently, mortality was ascertained through a US National Death Index (NDI) search.
+Methods of ascertainment and adjudication through the course of trial follow-up have been previously described [@sprintFinal].
+In the final report of the SPRINT trial, mortality was ascertained through a US National Death Index (NDI) search completed in December 2016 [@sprintFinal].
+For the current analysis, we completed an NDI search including deaths through 2020.
+As the NDI is continuously updated, the current analysis includes some deaths in 2016 that were not identified at the time of the final SPRINT report.
 Possible matches were identified according to NDI guidelines [@ndi_guide].
 Deaths were treated as confirmed if they were a Class 1 match, or a Class 2, 3, or 4 match with a probabilistic score above cutoffs recommended by the NDI [@ndi_guide].
-NDI follow-up began in 2016 and ended on the date of death or December 31, 2020.
 Deaths ascertained in 2020 were based on the NDI preliminary data release.
@@
 We identified 3074 participants with 3 or more outpatient and trial BP measurements.
-After excluding `r 3074 - n_ehr` participants without EHR data following the conclusion of trial follow-up, a total of `r n_ehr` patients were included for the ancillary analysis.
+After excluding `r 3074 - n_ehr` participants without EHR data following July 2016 (i.e., conclusion of the trial phase), a total of `r n_ehr` patients were included for the ancillary analysis.
 Because encounter type information was inconsistently available (i.e.
@@
 
-__Statistical Analysis__: Given the a priori expectation that treatment group differences may not be constant as a function of follow-up time (i.e.
+__Statistical Analysis__: Given the _a priori_ expectation that treatment group differences may not be constant as a function of follow-up time (i.e.
 the proportional hazards assumption was likely to be invalid), we modeled treatment group differences as a function of time using two approaches.
-The first approach split each participant's follow-up time into non-overlapping trial and cohort phases, and estimated regression coefficients for intensive treatment separately during each phase [@zhang_time_varying_2018].
+The first approach split each participant's follow-up time into non-overlapping trial and observational phases, and estimated regression coefficients for intensive treatment separately during each phase [@zhang_time_varying_2018].
 The second approach estimated a regression coefficient for intensive treatment as a continuous function of time since randomization [@martinussen2007dynamic; @therneau2017using].
@@
 We examined the trajectory of SBP following the conclusion of the trial using outpatient SBPs extracted from the EHR.
-Mean between-group differences in outpatient SBP following the conclusion of trial follow-up were estimated using linear mixed models.
-Models included random effects for participant and clinic site and an interaction between treatment group and time since randomization, which was flexibly modeled using B-splines.
+Mean between-group differences in outpatient SBP were estimated using linear mixed models.
+Models included random effects for participant and clinic site and an interaction between treatment group and time since randomization, which was modeled using B-splines.
 All analyses were performed using SAS version 9.4 (SAS Institute Inc, Cary, NC) and R version `r paste(version$major, version$minor, sep = '.')` (R Project for Statistical Computing [http://www.r-project.org]) with multiple auxiliary R packages.[@r_language; @table.glue; @tidyverse; @survival_package; @timereg; @targets] All hypothesis tests were 2-sided, and P values less than 0.05 were considered statistically significant.
@@
 ___All-cause Mortality___.
-In both treatment groups, median follow-up time was `r ap_b_o$fup_est` years.
+In both treatment groups, the total median follow-up time was `r ap_b_o$fup_est` years.
 A total of `r ap_b_i$n_acm` and `r ap_b_s$n_acm` deaths occurred among participants randomized to intensive and standard treatment, respectively (__eTable 2__).
-The hazard ratio (HR) for all-cause mortality comparing intensive to standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial phase, and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during the cohort phase.
-The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif_acm` to `r t_atten_acm` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 1__).
-In subgroups based on age, sex, race, CKD, cognitive function, and frailty status, there was no evidence that intensive treatment during the trial phase produced benefit for all-cause mortality during the cohort phase of follow-up (__Figure 2__).
+The hazard ratio (HR) for all-cause mortality comparing intensive to standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial phase, and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during the observational phase.
+The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif_acm` to `r t_atten_acm` years from randomization, and was attenuated throughout the remainder of the observational phase (__Figure 1__).
+In subgroups based on age, sex, race, CKD, cognitive function, and frailty status, there was no evidence that intensive treatment during the trial phase produced benefit for all-cause mortality during the observational phase of follow-up (__Figure 2__).
 
@@
 A total of `r ap_b_i$n_cvd` and `r ap_b_s$n_cvd` CVD deaths occurred among participants randomized to intensive and standard treatment, respectively (__eTable 3__).
-The HR for CVD mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the cohort phase.
-The time-dependent effect of intensive versus standard treatment indicated a benefit for CVD mortality from `r t_signif_cvd` to `r t_atten_cvd` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 3__).
-Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase produced benefit for CVD mortality during the cohort phase (__Figure 4__).
+The HR for CVD mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the observational phase.
+The time-dependent effect of intensive versus standard treatment indicated a benefit for CVD mortality from `r t_signif_cvd` to `r t_atten_cvd` years from randomization, and was attenuated throughout the remainder of the observational phase (__Figure 3__).
+Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase produced benefit for CVD mortality during the observational phase (__Figure 4__).
 
 ___Blood Pressure___.
-Amongst a subset of `r n_ehr` trial participants, the median number of outpatient BP measurements extracted from the EHR during the cohort phase of follow-up was X (Interquartile range Y to Z; NMP to add this result).
+Amongst a subset of `r n_ehr` trial participants, the median number of outpatient BP measurements extracted from the EHR during the observational phase of follow-up was 20 (IQR 10 to 34).
 The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization (__Figure 5__).
 For participants randomized to standard treatment, mean (95% CI) SBP was estimated to be `r f(bp_viz_5yr$Standard)` at 5 years and `r f(bp_viz_10yr$Standard)` post-randomization.
-The between-group difference in mean SBP levels (intensive minus standard)  was `r f(bp_diff_5yr)` mm Hg at 5 years post-randomization and was reduced to `r f(bp_diff_5yr)` at 10 years post-randomization (__eFigure 1__).
+The between-group difference in mean SBP levels (intensive minus standard)  was `r f(bp_diff_5yr)` mm Hg at 5 years post-randomization and was reduced to `r f(bp_diff_5yr)` at 10 years post-randomization (__eFigure 2__).
 
@@
 
-The current study analyzed all-cause and CVD-mortality among SPRINT participants up to 10 years following randomization, finding that the benefits associated with intensive treatment quickly attenuated after the trial ended.
+The current study analyzed all-cause and CVD-mortality among SPRINT participants up to 10 years following randomization, finding that the benefits associated with intensive treatment quickly attenuated after the trial intervention was discontinued.
 Time-varying estimates of the benefit of intensive treatment for all-cause mortality were attenuated at `r t_atten_acm` years while the benefit for CVD-mortality was attenuated at `r t_atten_cvd` years post-randomization.
 Findings from our ancillary study of outpatient SBP measured in routine clinical practice indicated that the difference in SBP between treatment groups diminished over time, with no detectable difference in SBP approximately `r t_atten_sbp` years after randomization.
-These results in combination with the primary findings of SPRINT indicate that the beneficial effect of intensive treatment among adults with hypertension can be diminished if BP control is not sustained.
+These results in combination with the primary findings of SPRINT indicate that the beneficial effect of intensive treatment among adults with hypertension appears to diminish if BP control is not sustained.
 
@@
 Evidenced-based strategies to improve BP control addressed in the call to action include implementing treatment protocols, using integrated care teams, providing clinicians feedback on their performance, and promoting shared patient-provider management with self-measured BP monitoring.
-During the SPRINT trial, participants received team-based care consistent with strategies outlined in the 2020 call to action.
+During the SPRINT trial, participants received team-based care consistent with strategies outlined in the 2020 call to action and providers received real time feedback on individual and aggregate participant blood pressure control.
 After the trial phase, when these protocols were no longer followed, the incidence of all-cause mortality approximately doubled in both treatment groups.
-These data emphasize the potential benefit of implementing the goals and strategies of the 2020 US Surgeon General's call to action.
+These data emphasize the benefit that can be realized by implementing the goals and strategies of the 2020 US Surgeon General's call to action.
 Future research should continue to evaluate strategies for obtaining consistent BP control in clinical settings to reduce the burden of CVD, which remains the leading cause of death for US adults.
@@
 Second, while several studies have shown reasonable performance of using NDI diagnosis codes for defining CVD mortality, it is not as robust as the adjudication process used in the primary follow-up for the trial.
-Third, information about SBP control after the trial was limited to routine outpatient SBP values extracted from the EHR, which are known to poorly reflect the standardized BP measurement process used during the SPRINT trial.[@drawz_concordance_2020] While this prohibits definite conclusions about the absolute level of BP in both treatment groups, as well as pin-pointing when the between-group difference may have completely attenuated, the general observation of steadily increasing BPs for participants in the intensive treatment following the trial is likely still valid.
+Third, information about SBP control after the trial was limited to routine outpatient SBP values extracted from the EHR, which are known to poorly reflect the standardized BP measurement process used during the SPRINT trial [@drawz_concordance_2020].
+While this prohibits definite conclusions about the absolute level of BP in both treatment groups, as well as pin-pointing when the between-group difference may have completely attenuated, the observation of steadily increasing SBP for participants in the intensive treatment and relatively stable SBP in the standard treatment group following the trial is likely still valid.
 
-In conclusion, while intensive treatment produced beneficial effects on mortality during the trial, there was no evidence that this produced sustained benefits on cardiovascular and all-cause mortality subsequent to the conclusion of trial follow-up.
+In conclusion, while intensive treatment produced beneficial effects on mortality during the trial, there was no evidence that this produced sustained benefits on cardiovascular and all-cause mortality subsequent to discontinuing the intervention protocol.
 Given steadily increasing mean SBP levels in participants randomized to intensive treatment after the trial, these results demonstrate that maintaining more intensive BP targets throughout adulthood will likely be essential for long-term CVD risk management.
@@
                            right  = 2.54 * 0.3937, 
-                           left   = 2.54 * 0.3937)
+                           left   = 2.54 * 0.3937,
+                           gutter = 0)
 
@@
 
-The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par`
+The trial phase and observational phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and observational phase for the population when time is measured relative to the date of randomization.`r single_par`
 
@@
 ```
-The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par`
+The trial phase and observational phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and observational phase for the population when time is measured relative to the date of randomization.`r single_par`
 
@@
 Shaded areas indicate a 95% confidence interval for the mean.`r single_par`
-The trial phase and cohort phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and cohort phase for the population when time is measured relative to the date of randomization.`r single_par`
+The trial phase and observational phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and observational phase for the population when time is measured relative to the date of randomization.`r single_par`
 
@@
                "Trial Phase",
-               "Cohort Phase"),
+               "Observational Phase"),
     colwidths = c(1, 5, 5)
@@
                "Trial Phase",
-               "Cohort Phase"),
+               "Observational Phase"),
     colwidths = c(1, 4, 4)
@@
 
-__eFigure 1__: Mean difference in systolic blood pressure measured in the trial and as part of routine clinical practice over time. `r single_par`
+__eFigure 1__: Timeline of landmark events and definition of 'trial' and 'observational' periods for the current analysis `r single_par`
+
+```{r}
+
+fpath <- file.path(
+  get_sasinet_drive(),
+  'bjaeger',
+  'sprint-mortality-ndi',
+  'fig',
+  'sprint-ndi-timeline.png'
+)
+
+knitr::include_graphics(
+  path = fpath,
+  dpi = 600
+)
+
+```
+
+\newpage
+
+__eFigure 2__: Mean difference in systolic blood pressure measured in the trial and as part of routine clinical practice over time. `r single_par`
 
@@
     page_size = page_size(orient = "landscape"),
-    page_margins = page_mar(bottom = 1/8, top = 1/8, 
-                            right = 1/4, left = 1/4),
+    page_margins = page_mar(bottom = 1/12, top = 1/12, 
+                            right = 1/8, left = 1/8),
     type = "continuous"

### 2022-02-15 _targets.R
more flexible bp viz for ehr only vs trial only updates

@@ -107,7 +107,8 @@ list(
     tar_target(base_viz_eff, ndi_base_viz_tv_effect(base_sub))
   ),
 
-  tar_target(bp_viz, bp_long_viz(date='01-15-22')),
+  tar_target(bp_viz, bp_long_viz(date_trial = '01-15-22',
+                                 date_ehr = '02-08-22')),
 
   # ndi_long_describe(ndi_data_list)
 

### 2022-03-02 doc/manuscript-rmd.Rmd
prep for conf slides

@@
 Fine, MD,^9^ Karen C.
-Johnson, MD,^7^ Cora E.
+Johnson, MD, MPH^7^ Cora E.
 Lewis, MD, MSPH^10^ Suzanne Oparil, MD,^11^ Michael V.
@@
 ^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake City, UT.
-^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis.
+^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis TN.
 ^8^Division of Renal Diseases & Hypertension, University of Minnesota, MN.
@@
                    base_viz_eff,
-                   long_viz_mort,
-                   long_viz_cvd,
+                   long_viz,
                    bp_viz,
@@
 CVD mortality for NDI-based follow-up used the NDI Plus System, which automatically identifies underlying causes of death from death certificates, including conversion to ICD-10 codes.
-we defined CVD mortality as any death containing the ICD-10 codes of I00 to I99 [@regards_ndi].
+We defined CVD mortality as any death containing the ICD-10 codes of I00 to I99 [@regards_ndi].
 
@@
 We identified 3074 participants with 3 or more outpatient and trial BP measurements.
-After excluding `r 3074 - n_ehr` participants without EHR data following July 2016 (i.e., conclusion of the trial phase), a total of `r n_ehr` patients were included for the ancillary analysis.
+After excluding `r 3074 - n_ehr` participants without EHR data following July 2016 (i.e., conclusion of the trial phase), a total of `r n_ehr` patients were included for the ancillary BP analysis.
 Because encounter type information was inconsistently available (i.e.
-outpatient, inpatient, observation, etc.), we defined a BP measurement as outpatient if there were was not a BP measurement on the preceding or following day, and if there were 2 or less BP measurements on a particular day.
+outpatient, inpatient, observation, etc.), we defined a BP measurement as outpatient if there was not a BP measurement on the preceding or following day, and if there were 2 or less BP measurements on a particular day.
 We averaged outpatient EHR BP readings when there were 2 on the same day.
@@
 
-The current study analyzed all-cause and CVD-mortality among SPRINT participants up to 10 years following randomization, finding that the benefits associated with intensive treatment quickly attenuated after the trial intervention was discontinued.
+The current study analyzed all-cause and CVD-mortality among SPRINT participants up to 10 years following randomization, finding that the benefits associated with intensive treatment quickly attenuated as SBP levels increased after the trial intervention was discontinued.
 Time-varying estimates of the benefit of intensive treatment for all-cause mortality were attenuated at `r t_atten_acm` years while the benefit for CVD-mortality was attenuated at `r t_atten_cvd` years post-randomization.
@@
 During the SPRINT trial, participants received team-based care consistent with strategies outlined in the 2020 call to action and providers received real time feedback on individual and aggregate participant blood pressure control.
-After the trial phase, when these protocols were no longer followed, the incidence of all-cause mortality approximately doubled in both treatment groups.
+After the trial phase, when these protocols were no longer followed by the primary care providers, the incidence of all-cause mortality approximately doubled in both treatment groups.
 These data emphasize the benefit that can be realized by implementing the goals and strategies of the 2020 US Surgeon General's call to action.
@@
 First, while we restricted analyses to high quality NDI matches, misclassification in linking participants to the NDI is possible.
-Second, while several studies have shown reasonable performance of using NDI diagnosis codes for defining CVD mortality, it is not as robust as the adjudication process used in the primary follow-up for the trial.
+Second, while several studies have shown reasonable performance of using NDI diagnosis codes for defining CVD mortality [@regards_ndi], it is not as robust as the adjudication process used in the primary follow-up for the trial.
 Third, information about SBP control after the trial was limited to routine outpatient SBP values extracted from the EHR, which are known to poorly reflect the standardized BP measurement process used during the SPRINT trial [@drawz_concordance_2020].
@@
 
-__Figure 2__: All-cause mortality hazard ratio for participants randomized to intensive versus standard treatment.
+__Figure 2__: All-cause mortality hazard ratio subgroup analysis for participants randomized to intensive versus standard treatment.
 
@@
 
-long_viz_mort
+plot_grid(long_viz$acm$trial,
+          long_viz$acm$cohort,
+          align = 'v',
+          nrow = 1)
 
@@
 
-The trial phase and observational phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and observational phase for the population when time is measured relative to the date of randomization.`r single_par`
+P-values test for heterogeneity in the treatment effect among subgroups and were not adjusted for multiple testing.`r single_par`
 
@@
 
-__Figure 4__: Cardiovascular mortality hazard ratio for participants randomized to intensive versus standard treatment.
+__Figure 4__: Cardiovascular mortality hazard ratio subgroup analysis for participants randomized to intensive versus standard treatment.
 
@@
 
-long_viz_cvd
+plot_grid(long_viz$cvd$trial,
+          long_viz$cvd$cohort,
+          align = 'v',
+          nrow = 1)
 
 ```
-The trial phase and observational phase do not overlap for individual participants, but due to participants being randomized on different days, there is an overlap in the trial and observational phase for the population when time is measured relative to the date of randomization.`r single_par`
+P-values test for heterogeneity in the treatment effect among subgroups and were not adjusted for multiple testing.`r single_par`
 

### 2022-03-02 index.Rmd
prep for conf slides

---
title: "Longer term All-Cause and Cardiovascular Mortality with Intensive Blood Pressure Control"
subtitle: "A Secondary Analysis of SPRINT"
author: "Nicholas Pajewski, PhD"
institute: "Wake Forest School of Medicine"
date: "2016/12/12 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

background-image: url(https://upload.wikimedia.org/wikipedia/commons/b/be/Sharingan_triple.svg)

```{r setup, include=FALSE}

options(htmltools.dir.version = FALSE)

source("packages.R")

knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE,
                      warning = FALSE,
                      dpi = 600, 
                      fig.width = 10, 
                      fig.height = 8)

tar_load(names = c(base_viz_inc,
                   base_viz_eff,
                   long_viz,
                   bp_viz,
                   ndi_baseline))

fig <- left_join(base_viz_inc, base_viz_eff)

```

???

Notes for you go here

---

```{r}

fig$inc_acm[[1]]

# If you wanted to make multiple figs on the same slide
# blank_plot <- ggplot(data.frame()) + aes() + theme_void()
# 
# legend <- get_legend(
#   fig$inc_acm[[1]]$plot + 
#     theme(legend.position = c(.55, .2),
#           legend.direction = 'horizontal')
# )
# 
# plot_grid(
#   # row 1 (legend)
#   legend, 
#   blank_plot,
#   # row 2 (figures)
#   fig$inc_acm[[1]]$plot + theme(legend.position = ''), 
#   fig$eff_acm[[1]],
#   # row 3 (table)
#   fig$inc_acm[[1]]$table, 
#   blank_plot,
#   ncol = 2,
#   align = 'v',
#   axis = 'lr',
#   rel_heights = c(1,10,2),
#   labels = c("","","A","B","",""),
#   label_x = 0.1,
#   label_y = 1.05
# )

```

---

### 2022-03-02 R/ndi_load.R

@@ -7,13 +7,14 @@
 ndi_load <- function(sasinet_drive = "Z", fname, ...) {
 
   path <- file.path(
-    paste0(sasinet_drive, ""),
+    paste0(sasinet_drive, ":"),
     "npajewski",
     "NDI",
     "Data",
     fname
   )
 
+  # sub_cvd = history of cvd variable.
 
   data_in <- read_csv(path) |>
     transmute(

### 2022-03-02 R/ndi_long_intr.R

@@ -1,0 +1,77 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param ndi_longitudinal
+#' @param group
+ndi_long_intr <- function(ndi_longitudinal, group) {
+
+  # treatment heterogeneity is assessed in each period separately.
+
+  if(group == 'overall'){
+    return(tibble(outcome = c('cvd', 'acm'),
+                  period = c('trial', 'cohort'),
+                  p.value = NA_real_))
+  }
+
+  ndi_split <- ndi_longitudinal |>
+    mutate(time = if_else(trial_phase == 'trial',
+                          end_yrs,
+                          end_yrs - start_yrs),
+           .before = start_yrs) |>
+    drop_na(all_of(group))
+
+  ndi_split <- ndi_split |>
+    split(ndi_split$trial_phase)
+
+  frmla <-
+    as.formula(
+      glue(
+        "Surv(time, mortality) ~
+          treatment * {group} +
+          strata(randSite)"
+      )
+    )
+
+  acm <- ndi_split |>
+    map(cph_fit, frmla) |>
+    map(anova) |>
+    map_dfr(tidy, .id = 'period') |>
+    filter(term == paste("treatment", group, sep = ':')) |>
+    transmute(period, p.value)
+
+  frmla <- as.formula(
+    glue(
+      "Event(time, cvd_mortality) ~
+          const(treatment) * const({group}) +
+          cluster(randSite)"
+    )
+  )
+
+  cvd <- ndi_split |>
+    map_dfr(
+      ~ {
+
+        fit <- comp.risk(formula = frmla,
+                         data = .x,
+                         cause = 1,
+                         model = 'prop',
+                         n.sim = 1000)
+
+        wald_index <- which(str_detect(rownames(coef(fit)), ':'))
+
+        wald_pval <- fit |>
+          wald.test(coef.null = wald_index) |>
+          getElement('p.value')
+
+        tibble(p.value = wald_pval)
+
+      },
+      .id = 'period'
+    ) |>
+    transmute(period, p.value)
+
+  bind_rows(acm = acm, cvd = cvd, .id = 'outcome')
+
+}

### 2022-03-02 R/ndi_long_viz_forest.R

@@ -2,252 +2,305 @@
 #'
 #' .. content for \details{} ..
 #'
-#' @title
+#' @param trt_hetero
+#' @param long_ds
+#' @param recoders
 #' @param long_inf
-ndi_long_viz_mort <- function(long_inf, recoders) {
+ndi_long_viz <- function(long_inf,
+                         trt_hetero,
+                         long_ds,
+                         recoders) {
+
 
   rspec <- round_spec() |>
-    round_using_decimal(digits = 2)
+    round_using_magnitude(digits = c(2, 2, 1),
+                          breaks = c(1, 10, Inf))
 
-  data_init <- long_inf |>
-    mutate(variable = str_remove(variable, 'long_inf_'),
-           variable = recode(variable, !!!recoders$variable),
-           group = recode(group, !!!recoders$group))
+  recoders$levels$frail <- c("Fit", "Pre-frail", "Frail")
+  recoders$group[c('Fit', "Pre_frail", "Frail")] <- c("Fit", "Pre-frail", "Frail")
 
-  data_bind <- tibble(variable = unique(data_init$variable),
-                      group = recode(variable, !!!recoders$variable)) |>
-    expand_grid(tibble(trial_phase = c("cohort", "trial")))
+  gg_spanners <- tibble(
+    variable = recoders$variable,
+    group = variable
+  )
 
-  data_gg <- bind_rows(data_bind, data_init) |>
-    mutate(variable = fct_inorder(variable),
-           group = fct_inorder(group)) |>
-    arrange(variable, group, trial_phase) |>
-    slice(-c(1,3)) |>
-    group_by(trial_phase) |>
-    mutate(
-      x = seq(n(), 1),
-      .before = 1,
-      across(where(is.factor), as.character),
-      group = if_else(
-        condition = group == variable | group == 'Overall',
-        true = group,
-        false = paste(" ", group)
-      ),
-      trial_phase = factor(
-        trial_phase,
-        levels = c('trial', 'cohort'),
-        labels = c("Trial Phase", "Observational Phase")
-      ),
-      fill = 1 +
-        as.numeric(variable == 'Overall') +
-        as.numeric(variable == 'Overall' & trial_phase == 'Observational Phase') +
-        as.numeric(variable != group) +
-        as.numeric(variable != group & trial_phase == 'Observational Phase'),
-      fill = factor(fill),
-      label = table_value(mort_est, rspec = rspec),
[…]

### 2022-03-02 _targets.R

@@ -13,6 +13,8 @@ if(user == 'nmpieyeskey'){
 }
 
 
+# TODO: consider history of CVD as a subgroup.
+
 ## Load your R files
 lapply(list.files("./R", full.names = TRUE), source)
 
@@ -64,6 +66,18 @@ list(
     ndi_tabulate_characteristics(ndi_baseline, recoders)
   ),
 
+  trt_intr <- tar_map(
+    values = subgroups,
+    tar_target(long_trt_intr, ndi_long_intr(ndi_longitudinal, group))
+  ),
+
+  tar_combine(
+    trt_hetero,
+    trt_intr[[1]],
+    command = bind_rows(!!!.x, .id = 'variable') |>
+      mutate(variable = str_remove(variable, '^long_trt_intr_'))
+  ),
+
   # arrow assignment is necessary to make 'long' an object that
   # will be recognized in downstream tar_ functions
   long <- tar_map(
@@ -90,8 +104,10 @@ list(
       mutate(variable = str_remove(variable, '^long_inf_'))
   ),
 
-  tar_target(long_viz_mort, ndi_long_viz_mort(long_inf, recoders)),
-  tar_target(long_viz_cvd, ndi_long_viz_cvd(long_inf, recoders)),
+  tar_target(long_viz, ndi_long_viz(long_inf,
+                                    trt_hetero,
+                                    long_ds,
+                                    recoders)),
 
   base <- tar_map(
     values = subgroups,

### 2022-03-02 packages.R

@@ -22,6 +22,10 @@ library(officer)
 library(cowplot)
 library(gtsummary)
 
+library(grid)
+library(glue)
+library(broom)
+
 conflict_prefer("filter", "dplyr")
 conflict_prefer("select", "dplyr")
 conflict_prefer("border", "flextable")

### 2022-03-02 index.Rmd
putting slides in the pipe

@@
 
+# IN CASE YOU NEED THIS IN PDF: Run this code in console:
+# pagedown::chrome_print('index.html')
+# But avoid using the pdf version when you present b/c you 
+# can access keyboard shortcuts if you present from html.
+# e.g., presenter mode by pressing 'p'
+
 options(htmltools.dir.version = FALSE)
-
-source("packages.R")
 
@@
                       warning = FALSE,
-                      dpi = 600, 
-                      fig.width = 10, 
-                      fig.height = 8)
+                      dpi = 600)
 
@@
 
-```{r}
+```{r fig.width = 10, fig.height = 8}
 
-fig$inc_acm[[1]]
+fig$inc_cvd[[1]] + 
+  theme(legend.key.size = unit(1.2, 'cm'), #change legend key size
+        legend.key.height = unit(1.2, 'cm'), #change legend key height
+        legend.key.width = unit(1.2, 'cm'), #change legend key width
+        legend.title = element_text(size=16), #change legend title font size
+        legend.text = element_text(size=16)) #change legend text font size
 

### 2022-03-02 R/ndi_load.R

@@ -4,10 +4,10 @@
 #'
 #' @title
 #' @param sasinet_drive
-ndi_load <- function(sasinet_drive = "Z", fname, ...) {
+ndi_load <- function(sasinet_drive = "Z:", fname, ...) {
 
   path <- file.path(
-    paste0(sasinet_drive, ":"),
+    paste0(sasinet_drive),
     "npajewski",
     "NDI",
     "Data",

### 2022-03-02 _targets.R

@@ -158,7 +158,9 @@ list(
 
   tar_render(manuscript, "doc/manuscript-rmd.Rmd"),
 
-  tar_render(readme, "README.Rmd")
+  tar_render(readme, "README.Rmd"),
+
+  tar_render(slides, "index.Rmd")
 
 )
 

### 2022-03-14 doc/manuscript-rmd.Rmd
slides first draft, PK&WC feedback

@@
 Randomization began on November 8, 2010 and ended in March 2013.
-On August 20, 2015, the SPRINT trial prematurely concluded due to the clear benefit of intensive treatment.
+In addition to the trial’s intervention phase, which spanned November 8, 2010 through August 20, 2015, when the administrative decision to stop the trial was made, the current study includes as part of the “trial phase” the additional period of study-provided antihypertensive medications prior to the final closeout visit on July 1, 2016 (eFigure 1).
+Observational follow-up continued through December 2020
+
 As trial follow-up visits and provision of antihypertensive medication continued through July 2016 [@sprintFinal], the current study considers a trial phase spanning from November 8, 2010 through July 2016 and an observational phase spanning from August 2017 through December, 2020 (__eFigure 1__).
@@
 
-Meta analyses of randomized trials have shown that pharmacological blood pressure (BP) lowering reduces the risk of major CVD events across the spectrum of initial BP [@rahimi_age_stratified_2021; @bundy_systolic_2017].
+<!-- Meta analyses of randomized trials have shown that pharmacological blood pressure (BP) lowering reduces the risk of major CVD events across the spectrum of initial BP [@rahimi_age_stratified_2021; @bundy_systolic_2017].
+-->
 
@@
 ```
+
+\newpage
 

### 2022-03-14 index.Rmd
rdy to discuss slides with NMP
slides first draft, PK&WC feedback

@@
 subtitle: "A Secondary Analysis of SPRINT"
-author: "Nicholas Pajewski, PhD"
+author: "Nicholas M. Pajewski, PhD & Byron C. Jaeger, PhD"
 institute: "Wake Forest School of Medicine"
-date: "2016/12/12 (updated: `r Sys.Date()`)"
+date: "`r Sys.Date()`"
 output:
@@
 ---
-
-background-image: url(https://upload.wikimedia.org/wikipedia/commons/b/be/Sharingan_triple.svg)
 
@@
 
-tar_load(names = c(base_viz_inc,
+tar_load(names = c(recoders,
+                   base_viz_inc,
                    base_viz_eff,
-                   long_viz,
                    bp_viz,
+                   tbl_characteristics,
+                   base_ds, 
+                   long_ds,
+                   long_inf,
+                   trt_hetero,
+                   base_inf,
                    ndi_baseline))
+
+long_viz <-
+  ndi_long_viz(long_inf, trt_hetero, long_ds, recoders, size_text = 4.5)
+
+n_total <- nrow(ndi_baseline)
+
+n_ehr <- sum(ndi_baseline$ehr_ancillary == 'Yes')
 
@@
 
+## Importance and Background
+
+- SPRINT showed that intensive treatment reduced the risk of cardiovascular and all-cause mortality.
+
+- The trial began in 2010 and the last study closeout visit occurred in 2016.
+
+## Objective and Methods
+
+- We assessed whether the benefit of intensive treatment persisted after the final study closeout visit in 2016.
+
+- We tracked cardiovascular and all-cause mortality events from the US National Death Index from 2016 to 2020.
+
+- In a subset of participants, we examined outpatient systolic blood pressure levels measured in routine clinical practice after the trial.
+
 ???
 
-Notes for you go here
+intensive treatment: Systolic blood pressure goal of < 120 mm Hg
+
+legacy effect: the persistence of benefit after the trial
 
@@
 fig$inc_cvd[[1]] +
-  theme(legend.key.size = unit(1.2, 'cm'), #change legend key size
-        legend.key.height = unit(1.2, 'cm'), #change legend key height
-        legend.key.width = unit(1.2, 'cm'), #change legend key width
-        legend.title = element_text(size=16), #change legend title font size
-        legend.text = element_text(size=16)) #change legend text font size
-
-# If you wanted to make multiple figs on the same slide
-# blank_plot <- ggplot(data.frame()) + aes() + theme_void()
-# 
-# legend <- get_legend(
-#   fig$inc_acm[[1]]$plot + 
-#     theme(legend.position = c(.55, .2),
-#           legend.direction = 'horizontal')
-# )
-# 
-# plot_grid(
-#   # row 1 (legend)
-#   legend, 
-#   blank_plot,
-#   # row 2 (figures)
-#   fig$inc_acm[[1]]$plot + theme(legend.position = ''), 
-#   fig$eff_acm[[1]],
-#   # row 3 (table)
-#   fig$inc_acm[[1]]$table, 
-#   blank_plot,
-#   ncol = 2,
-#   align = 'v',
-#   axis = 'lr',
-#   rel_heights = c(1,10,2),
-#   labels = c("","","A","B","",""),
-#   label_x = 0.1,
-#   label_y = 1.05
-# )
+  theme(legend.key.size = unit(1.2, 'cm'),    
+        legend.key.height = unit(1.2, 'cm'),  
+        legend.key.width = unit(1.2, 'cm'),   
+        legend.title = element_text(size=16), 
+        legend.text = element_text(size=16))
 
@@
 ---
+
+```{r fig.width = 10, fig.height = 8}
+
+bp_viz$levels + 
+  theme(legend.position = 'top',
+        legend.key.size = unit(1.2, 'cm'), #change legend key size
+        legend.key.height = unit(1.2, 'cm'), #change legend key height
+        legend.key.width = unit(1.2, 'cm'), #change legend key width
+        legend.title = element_text(size=16), #change legend title font size
+        legend.text = element_text(size=16))
+
+```
+
+---
+
+```{r fig.width = 10, fig.height = 8}
+
+long_viz$cvd$cohort
+
+```
+
+---
+
+## Key findings
+
+- Intensive treatment produced beneficial effects on mortality during the trial.
+
+- The benefits associated with intensive treatment quickly attenuated as systolic blood pressure levels increased among participants who underwent intensive treatment after the trial.
+
+- There was no evidence of sustained benefits after discontinuing the intervention protocol.
+
+## Conclusion
+
+Maintaining intensive BP targets throughout adulthood is likely an essential component of long-term CVD risk management.

### 2022-03-14 R/ndi_long_viz_forest.R

@@ -9,7 +9,8 @@
 ndi_long_viz <- function(long_inf,
                          trt_hetero,
                          long_ds,
-                         recoders) {
+                         recoders,
+                         size_text = 3) {
 
 
   rspec <- round_spec() |>
@@ -35,6 +36,17 @@ ndi_long_viz <- function(long_inf,
     pivot_wider(values_from = mort_inc,
                 names_from = treatment)
 
+  gg_acm_counts <- long_ds |>
+    transmute(
+      variable, group, trial_phase, treatment,
+      mort_label = table_glue(
+        "{as.integer(n_acm)}/{n_obs} ({as.integer(100*n_acm)/n_obs})",
+        rspec = rspec
+      )
+    ) |>
+    pivot_wider(values_from = mort_label,
+                names_from = treatment)
+
   gg_cvd_counts <- long_ds |>
     transmute(
       variable, group, trial_phase, treatment,
@@ -54,7 +66,7 @@ ndi_long_viz <- function(long_inf,
 
   gg_acm <- long_inf |>
     left_join(trt_hetero_merge_in) |>
-    left_join(gg_acm_incidence) |>
+    left_join(gg_acm_counts) |>
     transmute(
       variable = recode(variable, !!!recoders$variable),
       group = recode(group, !!!recoders$group),
@@ -150,7 +162,7 @@ ndi_long_viz <- function(long_inf,
   y_col_2 <- exp(-1.5)
   y_col_3 <- exp(1.5)
   y_col_4 <- exp(2.5)
-  size_text <- 3
+
   size_arrow <- 0.15
 
   ndi_forest_worker <- function(gg_data,
@@ -275,7 +287,7 @@ ndi_long_viz <- function(long_inf,
     .x = gg$acm,
     .y = c("Observational phase", "Trial phase"),
     .f = ndi_forest_worker,
-    header_ds = "Incidence rate (95% CI) per 1000 person-years",
+    header_ds = "no. of all-cause deaths / no. at risk (%)",
     y_col_0 = y_col_0,
     y_col_1 = y_col_1,
     y_col_2 = y_col_2,

### 2022-03-18 doc/manuscript-rmd.Rmd
NHLBI co-authors withdrew

@@
 
-Byron C.
-Jaeger, PhD,^1^ Adam P.
-Bress, PharmD, MS,^2,3^ Joshua D.
-Bundy, PhD,^4^ Alfred K.
-Cheung, MD,^5,6^ William C.
-Cushman, MD,^7^ Paul E.
-Drawz, MD,^8^ Lawrence J.
-Fine, MD,^9^ Karen C.
-Johnson, MD, MPH^7^ Cora E.
-Lewis, MD, MSPH^10^ Suzanne Oparil, MD,^11^ Michael V.
-Rocco, MD,^12^ Stephen R.
-Rapp, PhD, ^13^ Joni K.
-Snyder, RN,BSN,^9^ Mark A.
-Supiano, MD,^14^ Paul K.
-Whelton, MD,^4^ Jeff D.
-Williamson, MD, MHS,^15^ Jackson T.
-Wright, Jr, MD, PhD,^16^ David M.
-Reboussin, PhD,^1^ and Nicholas M.
-Pajewski, PhD^1^
+Byron C. Jaeger, PhD,^1^ 
+Adam P. Bress, PharmD, MS,^2,3^ 
+Joshua D. Bundy, PhD,^4^ 
+Alfred K. Cheung, MD,^5,6^ 
+William C. Cushman, MD,^7^ 
+Paul E. Drawz, MD,^8^ 
+Karen C. Johnson, MD, MPH^7^ 
+Cora E. Lewis, MD, MSPH^9^ 
+Suzanne Oparil, MD,^10^ 
+Michael V. Rocco, MD,^11^ 
+Stephen R. Rapp, PhD, ^12^ 
+Mark A. Supiano, MD,^13^ 
+Paul K. Whelton, MD,^4^ 
+Jeff D. Williamson, MD, MHS,^14^ 
+Jackson T. Wright, Jr, MD, PhD,^15^ 
+David M. Reboussin, PhD,^1^ and 
+Nicholas M. Pajewski, PhD^1^
 
-^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC.
-^2^Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT.
-^3^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT.
-^4^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA.
-^5^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT.
-^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake City, UT.
-^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis TN.
-^8^Division of Renal Diseases & Hypertension, University of Minnesota, MN.
-^9^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD.
-^10^Department of Epidemiology, University of Alabama at Birmingham.
-^11^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham.
-^12^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^13^Departments of Psychiatry and Behavioral Medicine & Social Sciences and Health Policy, Wake Forest School of Medicine, Winston-Salem, NC.
-^14^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City.
-^15^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC.
-^16^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
+^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC. `r single_par`
+^2^Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT. `r single_par`
+^3^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT. `r single_par`
+^4^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA. `r single_par`
+^5^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT. `r single_par`
+^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake City, UT. `r single_par`
+^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis, TN. `r single_par`
+^8^Division of Renal Diseases & Hypertension, University of Minnesota, MN. `r single_par`
+^9^Department of Epidemiology, University of Alabama at Birmingham. `r single_par`
+^10^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham. `r single_par`
+^11^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC. `r single_par`
+^11^Departments of Psychiatry and Behavioral Medicine & Social Sciences and Health Policy, Wake Forest School of Medicine, Winston-Salem, NC. `r single_par`
+^13^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City. `r single_par`
+^14^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC. `r single_par`
+^15^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
 \
@@
 __IMPORTANCE__.
-The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a systolic blood pressure (SBP) goal of <120mmHg, reduced the risk of cardiovascular and all-cause mortality.
-However, the legacy effect of intensive treatment on mortality, defined as the persistence of benefit after stopping intensive treatment, is unknown.
+The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a systolic blood pressure (SBP) goal of <120mmHg, reduced cardiovascular and all-cause mortality risk.
+However, the legacy effect of intensive treatment, defined as the persistence of benefit after stopping intensive treatment, is unknown.
 
-<!-- The edit to defn of legacy effect was recommended by JW -->
-
-__OBJECTIVE__. To evaluate the legacy effect of receiving intensive treatment during the SPRINT trial with all-cause and cardiovascular mortality up to ten years post-randomization.
+__OBJECTIVE__. To evaluate the effect of receiving intensive treatment during the SPRINT trial, which began in 2010 and ended in 2016, with cardiovascular and all-cause mortality through 2020.
 
 __DESIGN, SETTING, AND PARTICIPANTS__.
-Randomized clinical trial of 9361 patients aged 50 years or older with hypertension and increased cardiovascular risk, but without diabetes or history of stroke.
+Randomized clinical trial of 9361 patients aged ≥50 years with hypertension and increased cardiovascular risk, but without diabetes or history of stroke.
 Randomization began on November 8, 2010, the trial intervention ended on August 20, 2015, and observational trial follow-up visits occurred through July 2016.
@@
 __MAIN OUTCOMES AND MEASURES__.
-Extended trial follow-up assessed mortality via the US National Death Index, beginning in 2016 through December 31st, 2020.
+Extended trial follow-up assessed mortality via the US National Death Index from 2016 through 2020.
 In a subset of `r n_ehr` trial participants, outpatient SBP levels measured in routine clinical practice after the trial were examined.
@@
 However, at the median total follow-up of `r ap_b_o$fup_est` years, there was no longer evidence of benefit for cardiovascular mortality (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
-The estimated mean (95% CI) outpatient SBP among participants randomized to intensive treatment increased from `r f(bp_viz_5yr$Intensive)` at five years to `r f(bp_viz_10yr$Intensive)` ten years post-randomization.
+In a subgroup of `r n_ehr` participants, the estimated mean (95% CI) outpatient SBP among participants randomized to intensive treatment increased from `r f(bp_viz_5yr$Intensive)` at five years to `r f(bp_viz_10yr$Intensive)` ten years post-randomization.
 
 __CONCLUSIONS AND RELEVANCE__.
-The beneficial effect of intensive treatment on cardiovascular and all-cause mortality did not persist after the trial was stopped.
+The beneficial effect of intensive treatment on cardiovascular and all-cause mortality did not persist after the trial.
 Given increasing SBP levels in participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension.
@@
 Hypertension is the leading modifiable risk factor for cardiovascular disease (CVD) [@gdbJACC2020].
-The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a SBP target < 120 mm Hg, reduced the risk for cardiovascular and all-cause mortality [@sprintFinal].
-Because a clear benefit was evident for intensive treatment after a median follow-up of 3.3 years, SPRINT was stopped early.
-However, the persistence of benefit after the trial was stopped (i.e., "legacy" effect) of intensive treatment on cardiovascular and all-cause mortality has not been evaluated [@wander_legacy_2018].
+The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a SBP target < 120 mm Hg, reduced the risk for cardiovascular and all-cause mortality compared with treatment to an SBP target of < 140 mm Hg [@sprintFinal].
+Because a clear benefit was evident for intensive antihypertensive drug treatment after a median follow-up of 3.3 years, the trial was stopped early.
+However, the longer-term effect of randomization to intensive treatment on cardiovascular and all-cause mortality after the trial (i.e., the legacy effect [@wander_legacy_2018]) has not been evaluated.
 
-The objective of the current study was to evaluate the effect of stopping intensive treatment as specified by the SPRINT trial protocol by analyzing post-trial all-cause and cardiovascular mortality.
-A secondary objective was to examine SBP following the discontinuation of the trial intervention.
-To accomplish these objectives, we linked participants to the National Death Index (NDI) from 2016 through 2020 and extracted longitudinal outpatient measurements of SBP from 20xx to 2020 available in the electronic health record (EHR) for a subset of trial participants.
+The objective of the current study was to evaluate the long-term legacy effect of intensive treatment on mortality by passive follow-up using administrative data sources.
+We linked participants to the National Death Index (NDI) from 2016 through 2020, adding 4 years of follow-up after the conclusion of trial visits.
+A secondary objective was to examine change in attained BP levels following the discontinuation of the trial intervention and study visits.
+To examine this issue, we extracted longitudinal outpatient measurements of SBP from 2010 to 2020 available in the electronic health record (EHR) for a subset of trial participants.
 

### 2022-03-18 doc/manuscript-rmd.Rmd
draft rdy for NMP review; time to submit?

@@
 However, at the median total follow-up of `r ap_b_o$fup_est` years, there was no longer evidence of benefit for cardiovascular mortality (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`).
-In a subgroup of `r n_ehr` participants, the estimated mean (95% CI) outpatient SBP among participants randomized to intensive treatment increased from `r f(bp_viz_5yr$Intensive)` at five years to `r f(bp_viz_10yr$Intensive)` ten years post-randomization.
+In a subgroup of `r n_ehr` participants, the estimated mean (95% CI) outpatient SBP among participants randomized to intensive treatment increased from `r f(bp_viz_5yr$Intensive)` at five years to `r f(bp_viz_10yr$Intensive)` ten years following randomization.
 
@@
 __Trial Design__: The design and methods of SPRINT have been published previously [@ambrosius_design_2014; @sprintFinal].
-Briefly, it was a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk of CVD.
-Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed.
-Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (CKD), or a Framingham Risk Score of 15% or greater or if they were aged 75 years or older.
-Individuals residing in a nursing home, persons with a diagnosis of dementia (based on medical record review), and those treated with medications primarily used for dementia therapy were excluded, as were persons with prevalent diabetes mellitus, history of stroke, proteinuria > 1 gram per day, or polycystic kidney disease.
+Briefly, it was a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk for CVD.
+Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the use and number of anti-hypertensive agents prescribed.
+Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (CKD), a Framingham Risk Score of 15% or greater, or if they were aged 75 years or older.
+Individuals residing in a nursing home, with a diagnosis of dementia (based on medical record review), and those treated with medications prescribed for dementia were excluded, as were persons with prevalent diabetes mellitus, history of stroke, proteinuria > 1 gram per day, or polycystic kidney disease.
 Individuals at 102 sites in the United States and Puerto Rico were randomized (1:1) to a SBP goal of less than 120 mm Hg (intensive treatment group, n = 4678) or a goal of less than 140 mm Hg (standard treatment group, n = 4683), using random permuted blocks with the randomization stratified by clinic site.
 Randomization began on November 8, 2010 and ended in March 2013.
+On August 20, 2015, the Director of the National Heart, Lung, and Blood Institute accepted the data and safety monitoring board’s recommendation to inform the investigators and participants of the cardiovascular results, and decided to stop the trial, early, for benefit.
 In addition to the trial’s intervention phase, which spanned November 8, 2010 through August 20, 2015, when the administrative decision to stop the trial was made, the current study includes as part of the “trial phase” the additional period of study-provided antihypertensive medications prior to the final closeout visit on July 1, 2016 (eFigure 1).
-Observational follow-up continued through December 2020
-
-As trial follow-up visits and provision of antihypertensive medication continued through July 2016 [@sprintFinal], the current study considers a trial phase spanning from November 8, 2010 through July 2016 and an observational phase spanning from August 2017 through December, 2020 (__eFigure 1__).
+Observational follow-up continued through December 2020.
 The study was approved by the institutional review board at each participating site, and each participant provided written informed consent.
@@
 __Baseline Study Measurements__: Sociodemographic data were collected at baseline, with race or ethnicity information collected via self-report.
-The estimated glomerular filtration rate (eGFR) was calculated by the race-free 2021 CKD-EPI creatinine equation [@inker_new_2021].
+In the current study, the estimated glomerular filtration rate (eGFR) was calculated by the race-free 2021 CKD-EPI creatinine equation [@inker_new_2021].
 Cognitive function was assessed using the Montreal Cognitive Assessment (MoCA) [@nasreddine_montreal_2005].
@@
 In the final report of the SPRINT trial, mortality was ascertained through a US National Death Index (NDI) search completed in December 2016 [@sprintFinal].
-For the current analysis, we completed an NDI search including deaths through 2020.
-As the NDI is continuously updated, the current analysis includes some deaths in 2016 that were not identified at the time of the final SPRINT report.
+For the current analysis, we completed an NDI search including deaths through December 2020.
 Possible matches were identified according to NDI guidelines [@ndi_guide].
@@
 __EHR Ancillary Study__: Methods for the linkage of participants to their medical record number and the extraction of vital sign data have been previously described [@drawz_concordance_2020].
-We identified 3074 participants with 3 or more outpatient and trial BP measurements.
+We identified 3074 participants with 3 or more electronic health record reports of outpatient BP measurements during the trial.
 After excluding `r 3074 - n_ehr` participants without EHR data following July 2016 (i.e., conclusion of the trial phase), a total of `r n_ehr` patients were included for the ancillary BP analysis.
@@
 Models included random effects for participant and clinic site and an interaction between treatment group and time since randomization, which was modeled using B-splines.
-All analyses were performed using SAS version 9.4 (SAS Institute Inc, Cary, NC) and R version `r paste(version$major, version$minor, sep = '.')` (R Project for Statistical Computing [http://www.r-project.org]) with multiple auxiliary R packages.[@r_language; @table.glue; @tidyverse; @survival_package; @timereg; @targets] All hypothesis tests were 2-sided, and P values less than 0.05 were considered statistically significant.
+All analyses were performed using SAS version 9.4 (SAS Institute Inc, Cary, NC) and R version `r paste(version$major, version$minor, sep = '.')` (R Project for Statistical Computing [http://www.r-project.org]) with assistance from multiple R packages.[@r_language; @table.glue; @tidyverse; @survival_package; @timereg; @targets] All hypothesis tests were 2-sided, and P values less than 0.05 were considered statistically significant.
 No adjustments for multiple comparisons were made.
@@
 The mean (SD) SBP at baseline was `r inline_text(tbl_characteristics, variable = 'sbp', column = 'stat_0_1')` and `r inline_text(tbl_characteristics, variable = 'moca', column = 'stat_0_1')`% of participants MoCA scores were below the 10^th^ percentile.
-Compared to participants not included in the ancillary EHR study, participants included were more likely to be male, were older, had lower SBP and higher scores on the MoCA, and had a higher prevalence of CKD.
+Compared to participants not included in the ancillary EHR study, participants included were more likely to be male, older, with lower SBP and higher scores on the MoCA, and a higher prevalence of CKD.
 
@@
 The hazard ratio (HR) for all-cause mortality comparing intensive to standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial phase, and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during the observational phase.
-The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif_acm` to `r t_atten_acm` years from randomization, and was attenuated throughout the remainder of the observational phase (__Figure 1__).
+The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif_acm` to `r t_atten_acm` years following randomization, and was attenuated throughout the remainder of the observational phase (__Figure 1__).
 In subgroups based on age, sex, race, CKD, cognitive function, and frailty status, there was no evidence that intensive treatment during the trial phase produced benefit for all-cause mortality during the observational phase of follow-up (__Figure 2__).
@@
 ___Cardiovascular Mortality___.
-A total of `r ap_b_i$n_cvd` and `r ap_b_s$n_cvd` CVD deaths occurred among participants randomized to intensive and standard treatment, respectively (__eTable 3__).
+During the `r with(ap_b_o, table_glue("{fup_est}"))` years of follow-up, a total of `r ap_b_i$n_cvd` and `r ap_b_s$n_cvd` CVD deaths occurred among participants randomized to intensive and standard treatment, respectively (__eTable 3__).
 The HR for CVD mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the observational phase.
 The time-dependent effect of intensive versus standard treatment indicated a benefit for CVD mortality from `r t_signif_cvd` to `r t_atten_cvd` years from randomization, and was attenuated throughout the remainder of the observational phase (__Figure 3__).
-Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase produced benefit for CVD mortality during the observational phase (__Figure 4__).
+Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had a lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase produced benefit for CVD mortality during the observational phase (__Figure 4__).
 
 ___Blood Pressure___.
-Amongst a subset of `r n_ehr` trial participants, the median number of outpatient BP measurements extracted from the EHR during the observational phase of follow-up was 20 (IQR 10 to 34).
-The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization (__Figure 5__).
-For participants randomized to standard treatment, mean (95% CI) SBP was estimated to be `r f(bp_viz_5yr$Standard)` at 5 years and `r f(bp_viz_10yr$Standard)` post-randomization.
-The between-group difference in mean SBP levels (intensive minus standard)  was `r f(bp_diff_5yr)` mm Hg at 5 years post-randomization and was reduced to `r f(bp_diff_5yr)` at 10 years post-randomization (__eFigure 2__).
+Amongst a subset of `r n_ehr` trial participants, the median number of outpatient BP measurements extracted from the EHR during the observational phase of follow-up was 20 (interquartile range: 10 to 34).
+The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years following randomization and `r f(bp_viz_10yr$Intensive)` at 10 years following randomization (__Figure 5__).
+For participants randomized to standard treatment, mean (95% CI) SBP was estimated to be `r f(bp_viz_5yr$Standard)` at 5 years following randomization and `r f(bp_viz_10yr$Standard)` at 10 years following randomization.
+The between-group difference in mean SBP levels (intensive minus standard)  was `r f(bp_diff_5yr)` mm Hg at 5 years following randomization and was reduced to `r f(bp_diff_10yr)` at 10 years following randomization (__eFigure 2__).
 
@@
 
-The current study analyzed all-cause and CVD-mortality among SPRINT participants up to 10 years following randomization, finding that the benefits associated with intensive treatment quickly attenuated as SBP levels increased after the trial intervention was discontinued.
-Time-varying estimates of the benefit of intensive treatment for all-cause mortality were attenuated at `r t_atten_acm` years while the benefit for CVD-mortality was attenuated at `r t_atten_cvd` years post-randomization.
-Findings from our ancillary study of outpatient SBP measured in routine clinical practice indicated that the difference in SBP between treatment groups diminished over time, with no detectable difference in SBP approximately `r t_atten_sbp` years after randomization.
-These results in combination with the primary findings of SPRINT indicate that the beneficial effect of intensive treatment among adults with hypertension appears to diminish if BP control is not sustained.
+The current study analyzed all-cause and CVD-mortality among trial participants up to 10 years following randomization, finding that the benefits associated with intensive treatment quickly attenuated after the trial intervention was discontinued.
+Time-varying estimates of the benefit of intensive treatment for all-cause mortality were attenuated at `r t_atten_acm` years while the benefit for CVD-mortality was attenuated at `r t_atten_cvd` years following randomization.
+Findings from our ancillary study of outpatient SBP measured in routine clinical practice indicated that the difference in SBP between treatment groups diminished steadily over time, with no detectable difference in SBP approximately 9 years after randomization.
+These results in combination with the primary findings of the trial indicate that the beneficial effect of intensive treatment among adults with hypertension appears to diminish if BP control is not sustained.
+
+The current study's results in the "trial phase" are similar but not identical to results from the "intervention period" of the SPRINT final report, which considered August 20, 2015 to July 29, 2016 as an observational post-intervention period [@sprintFinal].
+The current study included August 20, 2015 to July 1, 2016 in the “trial phase” because antihypertensive medications continued to be provided until the final closeout visit on July 1, 2016.
+Also, as the NDI is continuously updated, the current analysis includes some deaths in 2016 that were not identified at the time of the final SPRINT report.
 
@@
 STEP found a HR of 0.72 (95% CI 0.39, 1.32) with intensive versus standard treatment for CVD mortality after a median follow-up of 3.34 years, but did not find evidence of a benefit for all-cause mortality.
-In the current study of SPRINT participants, the protective effect of intensive treatment for all-cause mortality was attenuated several years before the protective effect for CVD.
+In the current study, the protective effect of the SPRINT intensive treatment for all-cause mortality was attenuated several years before attenuation of the protective effect for CVD mortality.
 These results in combination with findings from the STEP trial suggest weaker evidence for reduced all-cause versus CVD mortality risk with intensive BP control.
-
-<!-- I commenting this sentence out because I think it may be too speculative for this Discussion.
-Nick, if you like the idea of including the commented sentence in the paper, can we set up a call with Paul Whelton to discuss?
--->
-
-<!-- Another contrasting result was that intensive treatment in SPRINT, but not STEP, increased the risk of acute kidney injury (AKI).
-While this may be attributable to the much lower prevalence of chronic kidney disease in STEP (2.3% vs 28.3%), the association of AKI with increased mortality risk may have been a contributing factor to the longer term effect of randomization to intensive treatment on all-cause mortality among SPRINT participants [@chawla2014aki; @rocco2018aki].
--->
-
-<!-- Meta analyses of randomized trials have shown that pharmacological blood pressure (BP) lowering reduces the risk of major CVD events across the spectrum of initial BP [@rahimi_age_stratified_2021; @bundy_systolic_2017].
--->
 
@@
 General population studies of adults living in the US with hypertension found that the prevalence of uncontrolled BP (SBP ≥ 140 mm Hg or DBP ≥ 90 mm Hg) increased from 2013 to 2017 [@muntner2020htn].
-In addition, an analysis of data from 464,585 adults enrolled in a Quest Diagnostics wellness program found SBP was 1 mm Hg and 3 mm Hg higher, depending on age group and sex, in April through December of 2020 versus their corresponding values throughout 2019 [@laffin_rise_2022].
-The current study shows that even for adults who have maintained intense SBP control for 3 years, increasing SBP levels can quickly diminish the protective effect.
+In addition, an analysis of data from 464,585 adults enrolled in a Quest Diagnostics wellness program found SBP was between 1 mm Hg and 3 mm Hg higher, depending on age group and sex, in April through December of 2020 versus their corresponding values throughout 2019 [@laffin_rise_2022].
+The current study shows that even for adults who have maintained intense SBP control for 3 years, increasing SBP levels may quickly diminish the protective effect.
 Combined with previous findings on rising BP levels among US adults, data from the current study emphasize the need for implementation of population- and community-level strategies to improve BP control in the US.
@@
 Second, while several studies have shown reasonable performance of using NDI diagnosis codes for defining CVD mortality [@regards_ndi], it is not as robust as the adjudication process used in the primary follow-up for the trial.
-Third, information about SBP control after the trial was limited to routine outpatient SBP values extracted from the EHR, which are known to poorly reflect the standardized BP measurement process used during the SPRINT trial [@drawz_concordance_2020].
+Third, information about SBP control after the trial was limited to routine outpatient SBP values in a subgroup of the SPRINT participants and was extracted from their EHR record, which is known to poorly reflect the standardized BP measurement process used during the trial [@drawz_concordance_2020].
 While this prohibits definite conclusions about the absolute level of BP in both treatment groups, as well as pin-pointing when the between-group difference may have completely attenuated, the observation of steadily increasing SBP for participants in the intensive treatment and relatively stable SBP in the standard treatment group following the trial is likely still valid.
@@
 In conclusion, while intensive treatment produced beneficial effects on mortality during the trial, there was no evidence that this produced sustained benefits on cardiovascular and all-cause mortality subsequent to discontinuing the intervention protocol.
-Given steadily increasing mean SBP levels in participants randomized to intensive treatment after the trial, these results demonstrate that maintaining more intensive BP targets throughout adulthood will likely be essential for long-term CVD risk management.
+Given steadily increasing mean SBP levels in participants randomized to intensive treatment after the trial, these results suggest that maintaining more intensive BP targets throughout adulthood will likely be essential for long-term CVD risk management.
 

### 2024-07-05 doc/manuscript-rmd.Rmd
updated ndi

@@
     reference_docx: style_manuscript_times_new_roman.docx
+    keep_md: true
 ---
@@
 
+ggsave(filename = 'fig/figure1.pdf', 
+       device = 'pdf', 
+       dpi = knitr::opts_chunk$get("dpi"),
+       width = knitr::opts_chunk$get("fig.width"), 
+       height = knitr::opts_chunk$get("fig.height"))
+
 ```
@@
           nrow = 1)
+
+ggsave(filename = 'fig/figure2.pdf', 
+       device = 'pdf', 
+       dpi = knitr::opts_chunk$get("dpi"),
+       width = knitr::opts_chunk$get("fig.width"), 
+       height = knitr::opts_chunk$get("fig.height"))
 
@@
 
+ggsave(filename = 'fig/figure3.pdf', 
+       device = 'pdf', 
+       dpi = knitr::opts_chunk$get("dpi"),
+       width = knitr::opts_chunk$get("fig.width"), 
+       height = knitr::opts_chunk$get("fig.height"))
+
 ```
@@
 
+ggsave(filename = 'fig/figure4.pdf', 
+       device = 'pdf', 
+       dpi = knitr::opts_chunk$get("dpi"),
+       width = knitr::opts_chunk$get("fig.width"), 
+       height = knitr::opts_chunk$get("fig.height"))
+
 ```
@@
 
-# fpath <- file.path(
-#   paste0(get_sasinet_drive(), ":"),
-#   'bjaeger',
-#   'sprint-mortality-ndi',
-#   'fig',
-#   'SPRINT_EHR_SBP_LongTerm_Comparison_122721.png'
-# )
-# 
-# knitr::include_graphics(
-#   path = fpath, 
-#   dpi = 600
-# )
+ggsave(filename = 'fig/figure5.pdf', 
+       device = 'pdf', 
+       dpi = knitr::opts_chunk$get("dpi"),
+       width = knitr::opts_chunk$get("fig.width"), 
+       height = knitr::opts_chunk$get("fig.height"))
 
@@
 
-fpath <- file.path(
-  get_sasinet_drive(),
-  'bjaeger',
-  'sprint-mortality-ndi',
-  'fig',
-  'sprint-ndi-timeline.png'
-)
-
-knitr::include_graphics(
-  path = fpath,
-  dpi = 600
-)
+# fpath <- file.path(
+#   # get_sasinet_drive(),
+#   # 'bjaeger',
+#   # 'sprint-mortality-ndi',
+#   # 'fig',
+#   'doc',
+#   'sprint-ndi-timeline.png'
+# )
+# 
+# knitr::include_graphics(
+#   path = fpath,
+#   dpi = 600
+# )
 
@@
   theme(legend.key.size = unit(1, 'cm'))
+
+ggsave(filename = 'fig/figure6.pdf', 
+       device = 'pdf', 
+       dpi = knitr::opts_chunk$get("dpi"),
+       width = knitr::opts_chunk$get("fig.width"), 
+       height = knitr::opts_chunk$get("fig.height"))
 

### 2024-07-05 index.qmd
updated ndi

---
title: "Longer Term All-Cause and Cardiovascular Mortality With Intensive Blood Pressure Control"
subtitle: "A Secondary Analysis of a Randomized Clinical Trial"
author: "Byron C. Jaeger, PhD"
institute: "Wake Forest University School of Medicine"
format: 
  revealjs:
    slide-number: true
    footer: Slides available at
    width: 1280
    height: 720
    fig-dpi: 300
    fig-width: 8
    fig-height: 6
    fig-align: "center"
    message: false
    bibliography: doc/refs.bib
    reference-location: document
knitr:
  opts_chunk:
    out.width: "100%"
---

```{r}

withr::with_dir(
  here::here(),
  code = {
    source("packages.R")
    targets::tar_load(
      names = c(long_viz,
                bp_viz,
                base_viz_inc,
                base_sub_overall)
    )
  }
)

ltd_2020 <- read_csv("T:/sprint/npajewski/NDI/Data/longterm_death.csv")
ltd_2023 <- read_csv("T:/sprint/npajewski/NDI/Data/longterm_death_2023.csv")

n_cvd_2020 <- sum(ltd_2020$cvd_event_cr == 1, na.rm = TRUE)
n_cvd_2023 <- sum(ltd_2023$cvd_event_cr == 1, na.rm = TRUE)

n_acm_2020 <- sum(ltd_2020$acm_event, na.rm = TRUE)
n_acm_2023 <- sum(ltd_2023$acm_event, na.rm = TRUE)

theme_fig <- function(x, ...){
  theme_bw() +
    theme(text = element_text(size = 18),
          axis.text = element_text(size = 18),
          axis.title = element_text(size = 18),
          panel.grid.major.x = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          legend.text = element_text(size = 18),
          legend.title = element_text(size = 18),
          legend.background = element_blank(),
          ...)
}

theme_tbl <- function(x, ...) {
  theme_void() +
    theme(text = element_text(size = 18),
          plot.caption = element_text(hjust = 0, size = 2),
          plot.title.position = "plot",
          plot.caption.position =  "plot",
          ...)
}

cols_bg <- viridis(3)
cols_tx <- pal_lancet()(2)

```

## Overview

- Background: SPRINT

- Why is longer term follow-up relevant?

- How did SPRINT do longer term follow-up on *all-cause and CVD mortality*?

- How did we analyze the longer term follow-up?

- Results

- Conclusion

## SPRINT

Two treatments with different systolic blood pressure (SBP) targets:

- *Standard*: SBP target of < 140 mm Hg
- *Intensive*: SBP target of < 120 mm Hg

Participants:

- aged 50 years or older 
- with hypertension and increased cardiovascular risk
- without diabetes or history of stroke

## Why do longer term follow-up?

::: {.column width="39%"}

- Investigate the legacy effect of intensive treatment

- Does targeting SBP < 120 mm Hg have long term benefits or harms compared to targeting SBP < 140 mm Hg?

:::

::: {.column width="59%"}

```{r}

ndi_data <- base_sub_overall$overall

fit_cr <- cuminc(ftime   = ndi_data$acm_years,
                 fstatus = factor(ndi_data$cvd_event_cr,
                                  levels = c(0, 1, 2),
                                  labels = c("- Censored",
                                             "- CVD mortality",
                                             "- Non-CVD mortality")),
                 group   = ndi_data$treatment,
                 cencode = '- Censored')

data_cr <-
  map_dfr(.x = fit_cr[1:4],
          .f = as_tibble,
          .id = 'fit') |>
  separate(fit, into = c('treatment','event'), sep = ' - ') |>
  mutate(
    treatment = factor(
      x = treatment,
      levels = c("Standard", "Intensive")
    ),
    lwr = est - 1.96 * sqrt(var),
    upr = est + 1.96 * sqrt(var)
  )

ymax <- 0.10

fig_cr <- ggplot(filter(data_cr, time < 4.48)) +
  aes(x = time,
      y = est,
      col = treatment,
      group = interaction(treatment, event)) +
  geom_line(linewidth = 1.5) +
  geom_ribbon(
    aes(x = time,
        ymin = est - sqrt(var) * 1.96,
        ymax = est + sqrt(var) * 1.96,
        fill = treatment,
        group = interaction(treatment, event)),
    alpha = 0.10,
    inherit.aes = FALSE
  ) +
  theme_fig(legend.position = 'inside',
            legend.position.inside = c(0.15, 0.75)) +
  labs(linetype = 'Event',
       color = 'Treatment',
       fill = 'Treatment') +
  ylab('Cumulative incidence') +
  xlab('Years since randomization') +
  scale_y_continuous(labels = scales::percent,
                     breaks = seq(0, ymax, by = 0.05),
                     limits = c(0, ymax)) +
  scale_x_continuous(breaks = c(0:12),
                     expand = c(0, 0, 0, 0),
                     limits = c(0, 13)) +
  scale_color_manual(values = cols_tx) +
  scale_fill_manual(values = cols_tx) + 
  annotate(geom = "text", hjust = 0, size = 7,
           x = 5, y = .045, label = "Non-CVD mortality") + 
  annotate(geom = "text", hjust = 0, size = 7,
           x = 5, y = 0.0175, 
           label = "CVD mortality") + 
  annotate(geom = "text", hjust = 0, size = 7,
           x = 11, y = 0.07, 
           label = "What\nhappens\nhere??")

fig_cr

```

:::

## Longer term all-cause/CVD mortality

::: {.column width="49%"}

- We assessed all-cause^[Deaths were treated as confirmed if they were a Class 1 match, or a Class 2, 3, or 4 match with a probabilistic score above cutoffs recommended by the NDI.] and cardiovascular disease^[Cardiovascular mortality for NDI-based follow-up used the NDI Plus System, which automatically identifies underlying causes of death from death certificates, including conversion to International Statistical Classification of Diseases and Related Health Problems, Tenth Revision (ICD-10) codes. We defined CVD mortality as any death containing the ICD-10 codes of I00 to I99] (CVD) mortality post-trial via the US National Death Index (NDI) from 2016-2020 [@jaeger2022longer].

- Among 2944 trial participants,^[We identified 3074 participants with 3 or more electronic health record reports of outpatient blood pressure measurements during the trial. After excluding 130 participants without electronic health record data following July 2016 (ie, conclusion of the trial), a total of 2944 patients were included for the ancillary blood pressure analysis.] post-trial SBP levels were analyzed [@drawz_concordance_2020].

:::

::: {.column width="49%"}
![](img/jama_cardio_cover.png)
:::

## How did we analyze longer term follow-up?

![](img/ph-assumption.gif){.absolute top="120" right="0"}

::: {.column width="50%"}
- assumed treatment group differences would not be constant over time.

- split each participant’s follow-up time into non-overlapping trial and observational phases

- estimated regression coefficients for intensive treatment separately during each phase.

:::

::: {.column width="49%"}
:::

# Results

## What's changed?

::: {.column width="49%"}
**2016-2020**

- `r n_acm_2020` all-cause mortality events
- `r n_cvd_2020` CVD mortality events
:::

::: {.column width="49%"}
**2016-2023**

- `r n_acm_2023` all-cause mortality events
- `r n_cvd_2023` CVD mortality events
:::

**Difference**:

- `r n_acm_2023 - n_acm_2020` additional all-cause mortality events
- `r n_cvd_2023 - n_cvd_2020` additional CVD mortality events
- No additional analyses on BP

## Systolic BP levels

::: {.column width="15%"}

SBP levels among the intensive group increased after the trial.

:::

::: {.column width="4%"}

:::

::: {.column width="79%"}
```{r}
#| fig-width: 10

bp_viz$levels + 
  theme_fig(legend.position = 'top')

```
:::

## Systolic BP difference

::: {.column width="15%"}
By year 10, the estimated difference was about 0
:::

::: {.column width="4%"}

:::

::: {.column width="79%"}
```{r}
#| fig-width: 10

bp_viz$diff + 
  theme_fig(legend.position = 'top') + 
  ylab("Between-group difference in SBP, mm Hg")

```
:::

## All-cause mortality (2016-2020)

::: {.column width="15%"}
Our original analysis showed the benefit for all-cause mortality attenuated quickly
:::

::: {.column width="4%"}
:::

::: {.column width="79%"}

```{r}
#| fig-width: 10

base_viz_inc$inc_acm[[1]]$plot + 
  annotate(geom = 'rect',
           xmin = 9.8,
           ymin = .18,
           xmax = Inf,
           ymax = .35,
           fill = 'black',
           alpha = 1) + 
  theme_fig(legend.position = 'inside',
            legend.position.inside = c(0.75, 0.20))

```
:::

## All-cause mortality (2016-2023)

::: {.column width="15%"}
This update shows an unexpected separation after several years of overlap
:::

::: {.column width="4%"}
:::

::: {.column width="79%"}

```{r}
#| fig-width: 10

base_viz_inc$inc_acm[[1]]$plot + 
  theme_fig(legend.position = 'inside',
            legend.position.inside = c(0.75, 0.20))

```

:::

## CVD mortality (2016-2020)

::: {.column width="15%"}
Our original analysis showed benefit for CVD mortality attenuated.
:::

::: {.column width="4%"}
:::

::: {.column width="79%"}

```{r}
#| fig-width: 10

base_viz_inc$inc_cvd[[1]] + 
  annotate(geom = 'rect',
           xmin = 9.8,
           ymin = .02,
           xmax = 13.2,
           ymax = .25,
           fill = 'black',
           alpha = 1) + 
  theme_fig(legend.position = 'inside',
            legend.position.inside = c(0.5, 0.69))

```

:::

## CVD mortality (2016-2023)

::: {.column width="15%"}
This update is a bit of a head-scratcher.
:::

::: {.column width="4%"}
:::

::: {.column width="79%"}
```{r}
#| fig-width: 10

base_viz_inc$inc_cvd[[1]] +
  theme_fig(legend.position = 'inside',
            legend.position.inside = c(0.5, 0.69))

```
:::

## All-cause mortalty (time-split analysis)

```{r}
#| fig-width: 14

library(cowplot)

plot_grid(
  long_viz$acm$trial,
  long_viz$acm$cohort,
  nrow = 1
)

```

## CVD mortalty (time-split analysis)

```{r}
#| fig-width: 14

plot_grid(
  long_viz$cvd$trial,
  long_viz$cvd$cohort,
  nrow = 1
)

```

## References

::: {#refs}
:::

### 2024-07-05 R/get_sasinet_drive.R

@@ -11,7 +11,7 @@ get_sasinet_drive <- function(username = NULL) {
 
   switch(
     username,
-    'bcjaeger' = 'Z:',
+    'bcjaeger' = 'T:/sprint',
     'nmpieyeskey' = 'O:/sprint'
   )
 

### 2024-07-05 R/ndi_base_describe.R

@@ -16,7 +16,8 @@ ndi_base_describe <- function(ndi_data_list) {
         group_map(
           .f = ~ prodlim(Hist(acm_years, 1 - acm_event) ~ 1, data = .x) |>
             quantile() |>
-            getElement('quantiles.survival') |>
+            unclass() %>%
+            as_tibble() %>%
             filter(q == 0.50) |>
             bind_cols(.y) # .y is the group info
         )
@@ -25,7 +26,8 @@ ndi_base_describe <- function(ndi_data_list) {
         prodlim(Hist(acm_years, 1 - acm_event) ~ 1,
                 data = ndi_data) |>
         quantile() |>
-        getElement('quantiles.survival') |>
+        unclass() %>%
+        as_tibble() %>%
         filter(q == 0.50) |>
         mutate(treatment = 'overall')
 

### 2024-07-05 R/ndi_base_infer.R

@@ -26,15 +26,11 @@ ndi_base_infer <- function(ndi_data_list) {
           treatment + treatment:time_factor + strata(randSite)
       )
 
+      k_mat <- diag(length(acm_fit_timevarying$coefficients))
+      k_mat[, 1] <- 1
 
       cph_estimate(acm_fit_timevarying,
-                   k_mat = rbind(c(1,0,0,0,0,0,0),
-                                 c(1,1,0,0,0,0,0),
-                                 c(1,0,1,0,0,0,0),
-                                 c(1,0,0,1,0,0,0),
-                                 c(1,0,0,0,1,0,0),
-                                 c(1,0,0,0,0,1,0),
-                                 c(1,0,0,0,0,0,1))) |>
+                   k_mat = k_mat) |>
         mutate(years = levels(ndi_intervals$time_factor),
                .before = 1)
 

### 2024-07-05 R/ndi_base_viz_incidence.R

@@ -1,9 +1,8 @@
-#' .. content for \description{} (no empty lines) ..
-#'
-#' .. content for \details{} ..
-#'
-#' @title
-#' @param base_sub
+
+# tar_load(base_sub_overall)
+# ndi_data_list <- base_sub_overall
+# ndi_data <- ndi_data_list$overall
+
 ndi_base_viz_incidence <- function(ndi_data_list) {
 
   cols_bg <- viridis(3)
@@ -82,12 +81,14 @@ ndi_base_viz_incidence <- function(ndi_data_list) {
 
       ymax <- ymax_candidates[min(which(ymax_candidates > ymax_true))]
 
+      xmax <- max(data_cr$time)
+
       fig_cr <- ggplot(data_cr) +
         aes(x = time,
             y = est,
             col = treatment,
             group = interaction(treatment, event)) +
-        geom_line(size = 1.5) +
+        geom_line(linewidth = 1.5) +
         geom_ribbon(
           aes(x = time,
               ymin = est - sqrt(var) * 1.96,
@@ -106,36 +107,34 @@ ndi_base_viz_incidence <- function(ndi_data_list) {
         scale_y_continuous(labels = scales::percent,
                            breaks = seq(0, ymax, by = 0.05),
                            limits = c(0, ymax)) +
-        scale_x_continuous(breaks = c(0:10),
-                           expand = c(0, 0, 0, 0),
-                           limits = c(0, 13)) +
+        scale_x_continuous(breaks = c(0:13),
+                           expand = c(0, 0, 0, 0)) +
         scale_color_manual(values = cols_tx) +
         scale_fill_manual(values = cols_tx)
-
-
 
       y_text <- data_cr |>
         group_by(treatment, event) |>
         filter(time == max(time)) |>
         group_by(event) |>
-        summarize(est = mean(est))
+        summarize(est_max = max(est) + 0.02,
+                  est_min = min(est) - 0.04)
 
       fig_cr <- add_annotations(
         fig_cr,
         cols = cols_bg,
         ymax = ymax,
-        xmax = 13,
+        xmax = xmax,
         mid_label = '\nTrial\nAnd\nObservational\nPhase',
         x_cohort_phase = 7
       ) +
         annotate(geom = 'text',
                  x = 11.5,
-                 y = y_text$est[y_text$event == 'Non-CVD mortality'],
+                 y = y_text$est_max[y_text$event == 'Non-CVD mortality'],
                  size = 5,
[…]

### 2024-07-05 R/ndi_base_viz_tv_effect.R

@@ -24,7 +24,10 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
         data = ndi_data
       )
 
-      plot_times <- seq(0, 10, length.out = 1000)
+      maxtime <- 10 # for original ndi
+      maxtime <- 13.5 # for updated ndi
+
+      plot_times <- seq(0, maxtime, length.out = 1000)
 
       spline_x <- cbind(1, ns(x = plot_times, df=6))
 
@@ -50,7 +53,7 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
         geom_line(color = cols_tx[2]) +
         scale_y_log10(limits = c(0.35, 2.5),
                       breaks = c(0.5, 1)) +
-        scale_x_continuous(breaks = seq(0,10)) +
+        scale_x_continuous(breaks = seq(0,maxtime)) +
         theme_fig() +
         geom_segment(aes(x = 0, xend = max(time),
                          y = 1, yend = 1),
@@ -61,7 +64,7 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
 
       fig_eff_acm <- add_annotations(fig_eff_acm, cols = cols_bg,
                                      ymax = 2.5,
-                                     xmax = 10,
+                                     xmax = maxtime,
                                      ymin = 0.35,
                                      mid_label = '\nTrial\nAnd\nObservational\nPhase',
                                      x_cohort_phase = 7,
@@ -73,7 +76,7 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
         model = 'prop',
         n.sim = 1000,
         # times = times,
-        data = ndi_data
+        data = drop_na(ndi_data, cvd_event_cr)
       )
 
       var_cvd <- as_tibble(fit_cvd$var.cum) |>
@@ -98,18 +101,18 @@ ndi_base_viz_tv_effect <- function(ndi_data_list) {
         geom_ribbon(alpha = 0.2) +
         geom_line(color = cols_tx[2]) +
         scale_y_log10(limits = c(0.15, 2.5), breaks = c(0.15, 0.25, 0.5, 1)) +
-        scale_x_continuous(breaks = seq(0,10)) +
+        scale_x_continuous(breaks = seq(0,maxtime)) +
         theme_fig() +
         geom_segment(aes(x = 0, xend = max(time),
                          y = 1, yend = 1),
                      linetype = 2,
                      color = 'black') +
         labs(x = 'Years since randomization',
-             y = 'Hazard ratio (95% CI)')
[…]

### 2024-07-05 R/ndi_load.R

@@ -4,7 +4,7 @@
 #'
 #' @title
 #' @param sasinet_drive
-ndi_load <- function(sasinet_drive = "Z:", fname, ...) {
+ndi_load <- function(sasinet_drive = "T:", fname, ...) {
 
   path <- file.path(
     paste0(sasinet_drive),
@@ -51,6 +51,7 @@ ndi_load <- function(sasinet_drive = "Z:", fname, ...) {
                      labels = c("Fit",
                                 "Pre_frail",
                                 "Frail")),
+      hdl = HDL,
       # treatment indicator with factor structure is useful for tables
       treatment = factor(intensive,
                          levels = c(0,1),

### 2024-07-05 R/ndi_long_infer.R

@@ -1,9 +1,6 @@
-#' .. content for \description{} (no empty lines) ..
-#'
-#' .. content for \details{} ..
-#'
-#' @title
-#' @param ndi_data_list
+
+# tar_load(long_sub_overall)
+# ndi_data_list <- long_sub_overall
 
 ndi_long_infer <- function(ndi_data_list, n_sim = 1000) {
 

### 2024-07-05 R/ndi_long_viz_forest.R

@@ -165,6 +165,12 @@ ndi_long_viz <- function(long_inf,
 
   size_arrow <- 0.15
 
+  gg$acm$cohort$pval <- gg$acm$cohort$p_acm
+  gg$acm$trial$pval <- gg$acm$trial$p_acm
+
+  gg$cvd$cohort$pval <- gg$cvd$cohort$p_cvd
+  gg$cvd$trial$pval <- gg$cvd$trial$p_cvd
+
   ndi_forest_worker <- function(gg_data,
                                 header_top,
                                 header_ds,
@@ -239,7 +245,7 @@ ndi_long_viz <- function(long_inf,
                 hjust = 0.5, size = size_text) +
       geom_text(aes(label = hr, y = y_col_3),
                 hjust = 0.5, size = size_text) +
-      geom_text(aes(label = p_acm, y = y_col_4),
+      geom_text(aes(label = pval, y = y_col_4),
                 hjust = 1, size = size_text) +
       geom_text(data = gg_header,
                 aes(x = x,
@@ -275,6 +281,7 @@ ndi_long_viz <- function(long_inf,
             axis.ticks.y = element_blank(),
             axis.text.y = element_blank(),
             axis.line = element_blank())
+
 
     cowplot::plot_grid(fig_main,
                        fig_bottom,

### 2024-07-05 R/ndi_tabulate_characteristics.R

@@ -35,7 +35,7 @@ ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
   }
 
   tbl_by_anci <- ndi_tbl_data |>
-    relocate(sbp, treatment, .after = race) |>
+    relocate(sbp, treatment, n_agents, .after = race) |>
     tbl_summary(
       by = 'ehr_ancillary',
       statistic = list(
@@ -52,6 +52,7 @@ ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
       type = list(age_yrs ~ "continuous",
                   age_cat ~ "dichotomous",
                   sbp ~ 'continuous',
+                  n_agents ~ 'continuous',
                   treatment ~ "dichotomous",
                   sex ~ "dichotomous",
                   race ~ "dichotomous",
@@ -69,7 +70,7 @@ ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
     add_p()
 
   tbl_by_tx <- ndi_tbl_data |>
-    relocate(sbp, treatment, ehr_ancillary, .after = race) |>
+    relocate(sbp, treatment, n_agents, ehr_ancillary, .after = race) |>
     tbl_summary(
       by = 'treatment',
       statistic = list(
@@ -87,6 +88,7 @@ ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
                   age_cat ~ "dichotomous",
                   ehr_ancillary ~ "dichotomous",
                   sbp ~ 'continuous',
+                  n_agents ~ 'continuous',
                   sex ~ "dichotomous",
                   race ~ "dichotomous",
                   ckd_2021 ~ "dichotomous",
@@ -120,6 +122,7 @@ ndi_tabulate_characteristics <- function(ndi_baseline, recoders) {
       type = list(age_yrs ~ "continuous",
                   age_cat ~ "dichotomous",
                   sbp ~ 'continuous',
+                  n_agents ~ 'continuous',
                   treatment ~ "dichotomous",
                   ehr_ancillary ~ "dichotomous",
                   sex ~ "dichotomous",

### 2024-07-05 R/ndi_update.R

@@ -1,0 +1,101 @@
+# Pull in the 2023 ndi data
+
+ndi_baseline_update <- function(ndi_baseline) {
+
+  ndi_2023 <-
+    read_csv("T:/sprint/npajewski/NDI/Data/longterm_death_2023.csv") %>%
+    select(pid, matches("^acm|^cvd")) %>%
+    mutate(cvd_event = if_else(cvd_event_cr == 2, 0, cvd_event_cr))
+
+  ndi_baseline %>%
+    select(-matches("^acm|^cvd")) %>%
+    left_join(ndi_2023)
+
+}
+
+ndi_longitudinal_update <- function(ndi_longitudinal){
+
+  ndi_2023 <-
+    read_csv("T:/sprint/npajewski/NDI/Data/longterm_death_2023.csv") %>%
+    select(pid, intensive, years_clo_end, matches("^acm|^cvd")) %>%
+    mutate(cvd_event = if_else(cvd_event_cr == 2, 0, cvd_event_cr)) %>%
+    group_by(pid) %>%
+    group_split() %>%
+    map_dfr(expand_row) %>%
+    mutate(trial_phase = factor(trial_phase,
+                                levels = c(0, 1),
+                                labels = c("trial", "cohort")))
+
+  ndi_longitudinal %>%
+    select(-start_yrs, -end_yrs, -mortality, -cvd_mortality) %>%
+    left_join(ndi_2023)
+
+}
+
+ndi_longitudinal_assert_replication_nmp <- function(){
+
+  attempt <- read_csv('T:/sprint/npajewski/NDI/Data/longterm_death.csv') %>%
+    select(pid,
+           years_clo_end,
+           intensive,
+           acm_years,
+           acm_event,
+           cvd_event_cr) %>%
+    group_by(pid) %>%
+    group_split() %>%
+    map_dfr(expand_row)
+
+  target <-
+    read_csv('T:/sprint/npajewski/NDI/Data/longterm_death_td_subgroup.csv') %>%
+    select(pid, intensive1:trial_phase)
+
+  testthat::expect_equal(target, attempt)
+
+}
+
+expand_row <- function(x){
+
+  out <- list()
+
+  if (x$acm_years < x$years_clo_end){
+
+    out[[1]] <- tibble(
+      pid = x$pid,
+      intensive1 = x$intensive,
+      intensive2 = 0,
+      start_yrs = 0,
+      end_yrs = x$acm_years,
[…]

### 2024-07-05 R/recoders_make.R

@@ -12,6 +12,7 @@ recoders_make <- function() {
                  "age_cat" = "Age group",
                  "age_yrs" = "Age, years",
                  "sbp" = "Systolic blood pressure",
+                 "n_agents" = "Number of blood pressure medications prescribed",
                  "sex" = "Sex",
                  "race" = "Race",
                  "ckd_2021" = "Chronic Kidney Disease",

### 2024-07-05 _targets.R

@@ -13,10 +13,9 @@ if(user == 'nmpieyeskey'){
 }
 
 
-# TODO: consider history of CVD as a subgroup.
-
 ## Load your R files
 lapply(list.files("./R", full.names = TRUE), source)
+
 
 subgroups <- tibble(group = c("overall",
                               "age_cat",
@@ -37,6 +36,10 @@ list(
              fname = 'longterm_death.csv',
              # participant identifier
              pid,
+             # Number of Medications Prescribed
+             n_agents,
+             # Number of Medication Classes Prescribed
+             n_classes,
              # systolic blood pressure at baseline
              sbp,
              # study site
@@ -45,7 +48,16 @@ list(
              acm_years, acm_event,
              # cvd mortality status
              cvd_event = as.numeric(cvd_event_cr == 1),
-             cvd_event_cr)
+             cvd_event_cr) %>%
+      ndi_baseline_update() %>%
+      # todo: ask nick about this
+      filter(pid != 10055290)
+  ),
+
+  tar_target(
+    # assert my data splitting replicates nick's on older data.
+    test_longi_update_code,
+    ndi_longitudinal_assert_replication_nmp()
   ),
 
   tar_target(
@@ -58,7 +70,10 @@ list(
                                   levels = c(0,1),
                                   labels = c("trial", "cohort")),
              start_yrs, end_yrs,
-             mortality, cvd_mortality)
+             mortality, cvd_mortality) %>%
+      ndi_longitudinal_update() %>%
+      # todo: ask nick about this
+      filter(pid != 10055290)
   ),
 
   tar_target(
@@ -123,8 +138,10 @@ list(
     tar_target(base_viz_eff, ndi_base_viz_tv_effect(base_sub))
   ),
 
-  tar_target(bp_viz, bp_long_viz(date_trial = '01-15-22',
-                                 date_ehr = '02-08-22')),
+  tar_target(bp_viz,
+             cue = tar_cue('always'),
+             bp_long_viz(date_trial = '01-15-22',
+                         date_ehr = '02-08-22')),
 
   # ndi_long_describe(ndi_data_list)
 
@@ -158,15 +175,17 @@ list(
 
   tar_render(manuscript, "doc/manuscript-rmd.Rmd"),
 
-  tar_render(readme, "README.Rmd"),
+  tar_render(readme, "README.Rmd")
 
-  tar_render(slides, "index.Rmd")
[…]

### 2024-07-05 packages.R

@@ -8,10 +8,10 @@ library(ggsci)
 library(tidyverse)
 library(survival)
 library(splines)
-library(Greg)
+library(Greg)           # for timeSplitter
 library(timereg)
 library(multcomp)
-library(riskRegression)
+library(riskRegression) # for Score
 library(prodlim)
 library(viridis)
 library(table.glue)
@@ -21,7 +21,7 @@ library(cmprsk)
 library(officer)
 library(cowplot)
 library(gtsummary)
-
+library(magrittr)
 library(grid)
 library(glue)
 library(broom)

### 2024-07-18 index.qmd
DR's feedback
dropping some of the more tedious things from my slides so that this 12
minute presentation doesn't try to do too much.

@@
     slide-number: true
-    footer: Slides available at
+    footer: Slides available at https://www.byronjaeger.com/talk/ (google byron jaeger biostats talks)
     width: 1280
@@
 
+## Disclosures
+
+I have nothing to disclose.
+
+## TL;DR
+
+- After SPRINT, from 2016-2020, the mean systolic blood pressure levels of participants returned to about 140 mm Hg, and the benefit of intensive blood pressure control attenuated.
+
+- Additional follow-up from 2020-2023 provides an opportunity to think carefully about interventions that had a beneficial effect.
+
 ## Overview
@@
 - How did SPRINT do longer term follow-up on *all-cause and CVD mortality*?
-
-- How did we analyze the longer term follow-up?
 
@@
 
-## How did we analyze longer term follow-up?
-
-![](img/ph-assumption.gif){.absolute top="120" right="0"}
-
-::: {.column width="50%"}
-- assumed treatment group differences would not be constant over time.
-
-- split each participant’s follow-up time into non-overlapping trial and observational phases
-
-- estimated regression coefficients for intensive treatment separately during each phase.
-
-:::
-
-::: {.column width="49%"}
-:::
-
 # Results
 
-## What's changed?
+## Number of events
 
@@
 
-bp_viz$levels + 
+bp_viz$levels_ehr_only + 
   theme_fig(legend.position = 'top')
@@
 
-bp_viz$diff + 
+bp_viz$diff_ehr_only + 
   theme_fig(legend.position = 'top') + 
@@
 ::: {.column width="15%"}
-This update is a bit of a head-scratcher.
+A bit of a head-scratcher.
 :::
@@
 ```
+:::
+
+## Conclusion
+
+- After SPRINT, from 2016-2020, SBP returned to about 140 mm Hg, and the benefit of intensive blood pressure control attenuated.
+
+- The latest update shows what appears to be a resurfaced benefit in terms of CVD mortality. This could be attributed to:
+
+    + participants in the intensive blood pressure control group who maintained control after the trial
+
+    + intensive blood pressure control has higher estimated benefit for younger adults.
+
+## References
+
+::: {#refs}
+:::
+
+::: {style="font-size: 2em"}
+# BONUS ROUND
+:::
+
+## How did we analyze longer term follow-up?
+
+![](img/ph-assumption.gif){.absolute top="120" right="0"}
+
+::: {.column width="50%"}
+- assumed treatment group differences would not be constant over time.
+
+- split each participant’s follow-up time into non-overlapping trial and observational phases
+
+- estimated regression coefficients for intensive treatment separately during each phase.
+
+:::
+
+::: {.column width="49%"}
 :::
@@
 ```
-
-## References
-
-::: {#refs}
-:::

### 2024-07-18 R/bp_long_viz.R

@@ -91,8 +91,46 @@ bp_long_viz <- function(date_ehr = "02-08-22",
                               group = interaction(treatment, source)),
                 inherit.aes = FALSE)
 
+  fig_levels_ehr_only <-
+    ggplot(filter(bp_levels, source == "EHR")) +
+    aes(x = time_months/12,
+        y = Estimate,
+        color = treatment) +
+    geom_line(size = 0.85) +
+    labs(y = 'Systolic blood pressure, mm Hg',
+         x = 'Years since randomization',
+         fill = 'Treatment',
+         color = 'Treatment',
+         linetype = 'Data Source') +
+    scale_x_continuous(limits = c(0,10),
+                       breaks = seq(0,10)) +
+    scale_y_continuous(limits = c(110, 152),
+                       breaks = seq(110, 150, by=5)) +
+    scale_color_manual(values = cols_tx) +
+    scale_fill_manual(values = cols_tx) +
+    scale_linetype_manual(values = c(1,2)) +
+    theme_fig(legend.position = 'right') +
+    geom_ribbon(alpha = 0.2,
+                mapping = aes(x = time_months/12,
+                              ymin = Lower,
+                              ymax = Upper,
+                              fill = treatment,
+                              group = interaction(treatment, source)),
+                inherit.aes = FALSE)
+
+
+
   fig_levels <-
     add_annotations(fig_levels,
+                    cols = cols_bg,
+                    ymax = 152,
+                    xmax = 10,
+                    ymin = 110,
+                    ymult = 0.965,
+                    x_cohort_phase = 6)
+
+  fig_levels_ehr_only <-
+    add_annotations(fig_levels_ehr_only,
                     cols = cols_bg,
                     ymax = 152,
                     xmax = 10,
@@ -120,6 +158,24 @@ bp_long_viz <- function(date_ehr = "02-08-22",
                        breaks = seq(-5,25, by=5)) +
     theme_fig(legend.position = 'right')
 
+  fig_diff_ehr_only <-
+    ggplot(filter(bp_diff, source == "EHR")) +
+    aes(x = time_months/12,
+        y = Estimate) +
+    geom_line(color = cols_tx[2], size = 0.85) +
+    geom_ribbon(aes(ymin = Lower, ymax = Upper),
+                fill = cols_tx[2],
+                alpha = 0.2) +
+    labs(y = 'Between-group difference in systolic blood pressure, mm Hg',
[…]

### 2024-07-28 README.Rmd
adjust legend

@@
 # specification for tables
-rspec <- round_spec() |>
+rspec <- round_spec() %>%
   round_using_magnitude(digits = c(2, 1, 0),
@@
 
+withr::with_dir(
+  here::here(),
+  code = {
     tar_load(names = c(recoders,
@@
                        ndi_baseline))
+  }
+)
+
+```
+
+```{r make-results, include=FALSE}
 
@@
 
-inline$base_ds <- base_ds |>
+inline$base_ds <- base_ds %>%
   mutate(across(matches("est$|lwr$|upr"), table_value),
-         across(matches("^n_"), ~table_value(as.integer(.x)))) |>
-  as_inline(tbl_variables = tbl_variables,
-            tbl_values = setdiff(names(base_ds), tbl_variables))
+         across(matches("^n_"), ~table_value(as.integer(.x))))
+
+inline$base_ds <- as_inline(inline$base_ds,
+                            tbl_variables = tbl_variables,
+                            tbl_values = setdiff(names(base_ds), 
+                                                 tbl_variables))
 
@@
 
-inline$long_inf <- long_inf |>
+inline$long_inf <- long_inf %>%
   mutate(across(matches("est$|lwr$|upr"), table_value),
-         across(matches("^n_"), ~table_value(as.integer(.x)))) |>
-  as_inline(tbl_variables = tbl_variables,
-            tbl_values = setdiff(names(long_inf), tbl_variables))
+         across(matches("^n_"), ~table_value(as.integer(.x))))
+
+inline$long_inf <- as_inline(inline$long_inf,
+                             tbl_variables = tbl_variables,
+                             tbl_values = setdiff(names(long_inf),
+                                                  tbl_variables))
 
@@
 
-bp_viz_5yr <- bp_viz |> 
-  getElement('levels') |> 
-  getElement('data') |> 
+bp_viz_5yr <- bp_viz %>% 
+  getElement('levels') %>% 
+  getElement('data') %>% 
   filter(source == 'EHR', 
@@
 
-bp_viz_10yr <- bp_viz |> 
-  getElement('levels') |> 
-  getElement('data') |> 
+bp_viz_10yr <- bp_viz %>% 
+  getElement('levels') %>% 
+  getElement('data') %>% 
   filter(source=='EHR',
@@
 
-bp_diff_5yr <- bp_viz |> 
-  getElement('diff') |> 
-  getElement('data') |> 
+bp_diff_5yr <- bp_viz %>% 
+  getElement('diff') %>% 
+  getElement('data') %>% 
   filter(time_months == 60)
 
-t_atten_sbp <- bp_viz |> 
-  getElement('diff') |> 
-  getElement('data') |> 
+t_atten_sbp <- bp_viz %>% 
+  getElement('diff') %>% 
+  getElement('data') %>% 
   filter(source == 'EHR',
-         find_signchange_lgl(Lower)) |> 
-  pull(time_months) |> 
+         find_signchange_lgl(Lower)) %>% 
+  pull(time_months) %>% 
   divide_by(12)

### 2024-07-28 index.qmd
adjust legend

@@
     slide-number: true
-    footer: Slides available at https://www.byronjaeger.com/talk/ (google byron jaeger biostats talks)
     width: 1280
@@
 - Additional follow-up from 2020-2023 provides an opportunity to think carefully about interventions that had a beneficial effect.
+
+- Slides are available at https://www.byronjaeger.com/talk/ (google byron jaeger biostats talks)
 
@@
            alpha = 1) + 
-  theme_fig(legend.position = 'inside',
-            legend.position.inside = c(0.75, 0.20))
+  theme_fig(legend.position = 'top')
 
@@
 base_viz_inc$inc_acm[[1]]$plot + 
-  theme_fig(legend.position = 'inside',
-            legend.position.inside = c(0.75, 0.20))
+  theme_fig(legend.position = 'top')
 
@@
            alpha = 1) + 
-  theme_fig(legend.position = 'inside',
-            legend.position.inside = c(0.5, 0.69))
+  theme_fig(legend.position = 'top')
 
@@
 base_viz_inc$inc_cvd[[1]] +
-  theme_fig(legend.position = 'inside',
-            legend.position.inside = c(0.5, 0.69))
+  theme_fig(legend.position = 'top')
 

### 2024-07-28 R/ndi_base_viz_incidence.R

@@ -53,6 +53,9 @@ ndi_base_viz_incidence <- function(ndi_data_list) {
         x_cohort_phase = 7
       )
 
+      fig_kap$plot <- fig_kap$plot +
+        scale_x_continuous(expand = c(0, 0, 0, 0),
+                           breaks = 1:12)
 
       fit_cr <- cuminc(ftime   = ndi_data$acm_years,
                        fstatus = factor(ndi_data$cvd_event_cr,
